---
title: "Chapter 4: A reproducible modeling workflow"
---

This chapter turns model building into a repeatable workflow. Instead of fitting, preprocessing, and scoring in disconnected scripts, we keep the full pipeline in one object so comparisons stay fair.

The teaching task is Adelie vs Gentoo classification. We use bill measures, island, sex, and year; we leave out flipper length and body mass so tuning remains informative rather than trivially high-scoring.

## How to read this page

You already know `lm` and `glm`. Here we extend that mental model:

- A **recipe** is preprocessing.
- A **model specification** declares the learner.
- A **workflow** binds them.
- **`tune_grid()`** explores settings with CV.

1. First: one train/test split and `last_fit()` to show the smallest honest flow.  
2. Then: the canonical Day 2 tuned tree include (shared with slides) for the full pattern.

[← Reading companion](index.qmd) · [Back: Chapter 3](module-03-trees-and-cultures.qmd) · [Next: Chapter 5](module-05-advanced-models.qmd)

---

## First: one split, one model, no tuning (`last_fit`) {#train-test-last-fit}

Day 2 (Tuesday) slides focus on **cross-validation + tuning** (faster to teach one rhythm in the room). If you want the **smallest** end-to-end `tidymodels` example first — **hold out test rows**, fit on training only, then score — work through this block.

**Predictors (both sections):** **bill length & depth**, **island**, **sex**, **year** — we **drop flipper length and body mass** so the tuned tree section below is not “everything wins at 99%.” The [classification notebook](../notebooks/penguins-classification.Rmd) still uses full morphology for classical `glm` / `rpart` plots.

**Packages:** `tidymodels`, `palmerpenguins`, `dplyr`, `tidyr`.

```{r}
#| label: m04-tt-setup
#| message: false
#| warning: false
library(tidymodels)
library(palmerpenguins)
library(dplyr)
library(tidyr)

peng_cls <- palmerpenguins::penguins |>
  dplyr::filter(species %in% c("Adelie", "Gentoo")) |>
  dplyr::mutate(
    y = factor(species, levels = c("Adelie", "Gentoo")),
    year = as.numeric(year)
  ) |>
  dplyr::select(-species, -flipper_length_mm, -body_mass_g) |>
  tidyr::drop_na()
```

```{r}
#| label: m04-tt-split
set.seed(11)
p_split <- initial_split(peng_cls, prop = 0.75, strata = y)
train <- training(p_split)
test <- testing(p_split)

count(train, y, name = "n_train")
count(test, y, name = "n_test")
```

```{r}
#| label: m04-tt-recipe
rec_tt <- recipe(y ~ ., data = train) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())
```

```{r}
#| label: m04-tt-model
mod_glm <- logistic_reg() |>
  set_engine("glm") |>
  set_mode("classification")

wf_tt <- workflow() |>
  add_recipe(rec_tt) |>
  add_model(mod_glm)
```

```{r}
#| label: m04-tt-last-fit
fit_split <- last_fit(wf_tt, split = p_split, metrics = metric_set(accuracy, roc_auc))
collect_metrics(fit_split)
```

```{r}
#| label: m04-tt-conf-mat
collect_predictions(fit_split) |>
  conf_mat(truth = y, estimate = .pred_class)
```

When you continue to the canonical pipeline below, the include redefines `peng_cls` in a self-contained way (so slide chunks remain portable). The modeling grammar stays the same; only the resampling and tuning machinery becomes richer.

---

## Canonical pipeline (Day 2 slides + tuning) {#canonical-pipeline-tuesday}

The next block is the same canonical walkthrough used in the Day 2 lecture deck. Keeping one shared include means the chapter and slides do not drift apart.

{{< include ../_includes/day02-penguins-setup.qmd >}}

{{< include ../_includes/day02-tidymodels-walkthrough.qmd >}}

---

Lecture: [Day 2 (Tuesday) pipeline slides](../slides/day-02-tuesday.html#/part-c-pipeline)  
Notebook: [penguins-canonical-pipeline.Rmd](../notebooks/penguins-canonical-pipeline.Rmd)  
Next: [Chapter 5 — Stronger learners, same discipline](module-05-advanced-models.qmd)
