---
title: "Module 06: Metrics, Imbalance, and SHAP"
---

## Learning objectives

- Pick **metrics** that match prevalence and error costs (not accuracy by default).
- Read **confusion matrices**, **ROC**, and **PR** curves on imbalanced data.
- Use **VIP** and **SHAP** as **communication** tools — not causal proof.

[← Module library](index.qmd) · [Module 07 — Metric tuning lab](module-07-penguins-choose-metrics.qmd) · [Day 4 Part E](../slides/day-04-thursday.html#/part-e-metrics)

---

## On the slides (Day 4)

| Part | Topic | Slide link |
|------|--------|------------|
| E (first) | Multiclass imbalance (rare Chinstrap), majority baseline | [Imbalance block](../slides/day-04-thursday.html#/part-e-metrics) |
| E (metrics) | Macro metrics + per-class recall + confusion matrix | [Part E](../slides/day-04-thursday.html#/part-e-metrics) |
| F | Random forest VIP | [Part F — VIP](../slides/day-04-thursday.html#/part-f-vip) |
| F | SHAP on sex (no species) | [SHAP introduction](../slides/day-04-thursday.html#/shap---what-it-means-game-theory-intuition) |

Tuesday lecture tunes on **accuracy only**; Thursday Part E shows why multiclass imbalance needs **macro/per-class** metrics.

---

## Metrics toolbox

| Metric | Question it answers | When to use |
|--------|---------------------|-------------|
| **accuracy** | Overall correct? | Balanced classes, symmetric costs |
| **sensitivity** (recall) | Of true positives, how many caught? | Screening — missing a case is costly |
| **specificity** | Of true negatives, how many caught? | Avoid false alarms |
| **precision** | Of predicted positives, how many correct? | Confirmation after a screen |
| **F1** | Balance precision & recall | Single summary when both matter |
| **ROC-AUC** | Ranking across thresholds | Separability; can look fine when positives are rare |
| **PR-AUC** | Precision–recall area | **Rare positive** class |

In code, use **`yardstick::spec`** for specificity so R does not confuse it with a **`spec`** model object.

---

## Why accuracy misleads

On **imbalanced** data (many Adelie, few Gentoo in the teaching slice), a model that **always predicts the majority class** can show **high accuracy** while missing **every** minority case — useless for screening.

Numeric example: if prevalence is 90% negative and 10% positive, an always-negative model gets 90% accuracy but 0% sensitivity for the positive class.

The slides compute a **majority-class baseline** before comparing upsampling on a multiclass imbalanced slice ([Part E](../slides/day-04-thursday.html#/part-e-metrics)).

---

## Class imbalance and resampling

- **`step_upsample()`** duplicates minority rows **in training folds** — helps when the learner **ignores** the rare class.  
- In multiclass settings, upsampling is judged by **minority-class recall** and **macro metrics**, not only accuracy.
- Upsampling duplicates minority rows **inside training folds only**; assessment folds stay untouched for honest scoring.
- It can improve underrepresented-class detection at the cost of extra confusion between majority classes.

Lab: [Day 4 Block D — Imbalance](../exercises/mlss2026-lab-exercises.qmd) (microbiome `Late`). Minimal notebook: [multiclass upsampling](../notebooks/penguins-multiclass-upsampling.Rmd).

---

## Confusion matrix

Rows = **true** class; columns = **predicted** class.

- **False negatives** (true Gentoo, predicted Adelie) — costly in screening.  
- **False positives** — costly if treatment is invasive.

Practice reading the **minority row** on the [Day 4 slides](../slides/day-04-thursday.html#/part-e-metrics).

```{r}
#| label: m06-conf-demo
#| echo: true
#| eval: false
#| message: false
library(yardstick)
# After augment(fit, test_data):
# conf_mat(truth = y, estimate = .pred_class)
```

---

## ROC vs PR

- **ROC:** sensitivity vs \(1 - \text{specificity}\) across thresholds — good when both classes matter.  
- **PR:** precision vs recall — often more informative when **positives are rare**.

Threshold example: at threshold 0.2 you might get recall 0.90 but precision 0.18; ROC can still look decent while PR reveals large false-alert burden.

---

## Variable importance (VIP)

- **`vip`** on tree ensembles: how often splits use each predictor (impurity).  
- **Correlated** predictors **share** importance — not proof that changing one mm of bill **causes** species shift.  
- Full chapter: [Day 4 Part F](../slides/day-04-thursday.html#/part-f-vip).

---

## SHAP: Shapley values on penguins (sex) {#shap-shapley-values-on-penguins-sex}

**SHAP** = **SHapley additive explanations**. This section mirrors [Day 4 Part F SHAP slides](../slides/day-04-thursday.html#/shap---what-it-means-game-theory-intuition). We use a **non-trivial** task: predict **`sex`** from morphometrics and island/year while **omitting `species`** ([sex notebook](../notebooks/penguins-sex-classification.Rmd), [data card](../data/cards/palmer-penguins.qmd)).

### 1. What SHAP means (game theory intuition)

**Coalition story (no formulas required in class):**

1. Fix one penguin and a fitted model.
2. Imagine each **feature** is a player who can join a **team** (coalition).
3. For each coalition, the model predicts using only those features; others are integrated over the data (background).
4. The **Shapley value** is the fair average of a player's **marginal contributions** as players join in every possible order.

**Additivity (check on one bird):**

| Piece | Meaning |
|-------|---------|
| \(\phi_0\) | **Baseline** — average model output over background penguins |
| \(\phi_j\) | SHAP value for feature \(j\) on this row |
| Sum | \(\phi_0 + \sum_j \phi_j \approx \hat{p}(\text{male})\) for that row |

**Toy numeric row (illustration only):**

| Feature | SHAP \(\phi\) |
|---------|----------------|
| Baseline \(\phi_0\) | 0.42 |
| body_mass_g | +0.18 |
| bill_depth_mm | +0.09 |
| island_Dream | −0.05 |
| … | … |
| **Sum** | **0.71 = model P(male)** |

**VIP vs SHAP:** VIP = **global** split usage; SHAP = **local** decomposition for **one** penguin.

### 2. How to calculate and plot (R)

**Packages:** [`shapviz`](https://cran.r-project.org/package=shapviz), [`kernelshap`](https://cran.r-project.org/package=kernelshap) (plus course **`tidymodels`** stack). Install:

```r
install.packages(c("shapviz", "kernelshap"))
```

Helpers live in [`R/slide-viz-helpers.R`](../R/slide-viz-helpers.R). The slides source them automatically; in your own project:

```r
source("R/slide-viz-helpers.R")  # adjust path
```

```{r}
#| label: m06-shap-setup
#| include: false
#| message: false
#| warning: false
suppressPackageStartupMessages({
  library(tidymodels)
  library(shapviz)
  library(kernelshap)
})
source("../R/slide-viz-helpers.R")
pg_sex_m06 <- prep_penguins_sex()
sex_shap_m06 <- fit_sex_shap_bundle(
  pg_sex_m06,
  trees = 300,
  n_sample = 80,
  n_bg = 35,
  seed = 11
)
```

**Pipeline (same logic as slides):**

```{r}
#| label: m06-shap-code
#| echo: true
#| eval: false
pg_sex <- prep_penguins_sex()

rec_sex <- recipe(sex ~ bill_length_mm + bill_depth_mm + flipper_length_mm +
                    body_mass_g + island + year, data = pg_sex) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

rf_sex_spec <- rand_forest(trees = 300, mtry = 3, min_n = 2) |>
  set_engine("ranger", probability = TRUE) |>
  set_mode("classification")

fit_sex <- fit(
  workflow() |> add_recipe(rec_sex) |> add_model(rf_sex_spec),
  pg_sex
)

rec_prepped <- prep(extract_recipe(fit_sex, estimated = TRUE), training = pg_sex)
X_model <- bake(rec_prepped, new_data = pg_sex, all_predictors())

rf_engine <- extract_fit_parsnip(fit_sex)$fit
pred_fun <- function(object, X_new) {
  as.numeric(predict(object, data = X_new)$predictions[, "male"])
}
X_shap <- dplyr::slice_sample(X_model, n = 80)
bg_X <- dplyr::slice_sample(X_model, n = 35)
ks <- kernelshap(rf_engine, X = X_shap, pred_fun = pred_fun, bg_X = bg_X)
shp <- shapviz(ks, X_pred = X_shap)
```

**`bake()`** supplies the encoded predictor matrix the forest was trained on. **`kernelshap()`** compares each row to a background sample; **`shapviz()`** draws the plots. (Same steps as Tuesday's **`prep` / `bake`** demo — see [Day 2 pipeline](../slides/day-02-tuesday.html#/prep-and-bake---learn-rules-then-apply).)

For ROC on imbalanced Gentoo, always set **`event_level = "second"`** when using `.pred_Gentoo` — otherwise the curve can look **flipped** (AUC near 0). The slides use `plot_roc_gentoo()` from [`slide-viz-helpers.R`](../R/slide-viz-helpers.R).

**Beeswarm (global):**

```{r}
#| label: m06-shap-beeswarm
#| echo: false
#| fig-width: 9
#| fig-height: 6
#| eval: true
plot_shap_beeswarm(sex_shap_m06$shp)
```

**Waterfall (local) — male vs female in the SHAP sample:**

```{r}
#| label: m06-shap-waterfall-male
#| echo: false
#| fig-width: 9
#| fig-height: 5
#| eval: true
plot_shap_waterfall(
  sex_shap_m06$shp,
  row_id = sex_shap_m06$male_row,
  title = "Waterfall: one male penguin"
)
```

```{r}
#| label: m06-shap-waterfall-female
#| echo: false
#| fig-width: 9
#| fig-height: 5
#| eval: true
plot_shap_waterfall(
  sex_shap_m06$shp,
  row_id = sex_shap_m06$female_row,
  title = "Waterfall: one female penguin"
)
```

**VIP vs SHAP on the same sex model (module only):**

```{r}
#| label: m06-shap-vip-compare
#| echo: false
#| fig-width: 9
#| fig-height: 5
#| eval: true
plot_sex_rf_vip(sex_shap_m06$fit)
```

### 3. How to interpret results

**Say**

- "For **this** bird, the model leaned **male** partly because of **body mass** and **bill depth** in the encoded feature space."
- "The beeswarm shows **which features matter in this RF** across the sample we explained."
- "SHAP is a property of **this fitted model** on **observational** data."

**Do not say**

- "Bill length **causes** male penguins."
- "If we shortened flippers, sex would change" (intervention claim).
- "SHAP shows the model is **fair** by island" (needs subgroup evaluation).

**Assessment prompt (from [curriculum outline](../curriculum-outline.qmd)):** *What would you **not** claim from this SHAP summary?*

**Honest limits:** we fit on all rows for teaching plots; in a study you would compute SHAP on **training** rows only and hold out test data for performance. See [Palmer Penguins data card](../data/cards/palmer-penguins.qmd) for observational caveats.

---

## Course links

- **Slides:** [Day 4 (Thursday)](../slides/day-04-thursday.qmd)
- **Metric choice in tuning:** [Module 07](module-07-penguins-choose-metrics.qmd)
- **Notebook:** [penguins-sex-classification.Rmd](../notebooks/penguins-sex-classification.qmd)
- **Curriculum:** [outline](../curriculum-outline.qmd)
