---
title: "Module 07: Pick the score you care about (accuracy, kappa, ROC AUC)"
---

::: {.callout-important}
## Extension lab — not the Tuesday lecture default

**Day 2 slides** tune the classification tree on **cross-validated accuracy** only ([Part C](../slides/day-02-tuesday.html#/part-c-pipeline)). Use **this module** for the **afternoon lab** and **Day 4** when you compare `select_best(metric = "accuracy")` vs `"roc_auc"` vs `"kappa"` on imbalanced or multi-metric problems ([Part E](../slides/day-04-thursday.html#/part-e-metrics), [Day 4 lab](../exercises/index.qmd)).
:::

## Learning objectives

- Run **one** tuning exercise while asking R to track **several** metrics at once (`accuracy`, Cohen’s **`kappa`**, **`roc_auc`**).
- Use **`select_best()`** with different `metric =` arguments and see that the **best hyperparameters** need not agree across scores.
- Say in one sentence when **accuracy** misleads, why **kappa** adjusts for “getting lucky” on skewed labels, and what **ROC AUC** measures.

## Before you start

Build on [Module 04](module-04-pipeline.qmd#train-test-last-fit) (train/test + `glm`) and [Module 04 — canonical pipeline](module-04-pipeline.qmd#canonical-pipeline-tuesday) (recipe + tree + `tune_grid` from the shared include). Here we keep **cross-validated tuning** on the **full** `peng_cls` table — same Adelie vs Gentoo columns as **Day 2 (Tuesday)** slides (bill + island + sex + year; no flipper or body mass). Teaching default: tune on all rows; in a paper you would nest tuning inside training after a test lockbox.

**Packages:** same as Module 04 (`tidymodels`, `palmerpenguins`, `dplyr`, `tidyr`).

[← Module 04 (train/test beat)](module-04-pipeline.qmd#train-test-last-fit) · [Module library](index.qmd) · [Next: compare models →](module-08-penguins-compare-models.qmd)

```{r}
#| label: m07m-setup
#| message: false
#| warning: false
library(tidymodels)
library(palmerpenguins)
library(dplyr)
library(tidyr)

peng_cls <- palmerpenguins::penguins |>
  dplyr::filter(species %in% c("Adelie", "Gentoo")) |>
  dplyr::mutate(
    y = factor(species, levels = c("Adelie", "Gentoo")),
    year = as.numeric(year)
  ) |>
  dplyr::select(-species, -flipper_length_mm, -body_mass_g) |>
  tidyr::drop_na()

rec <- recipe(y ~ ., data = peng_cls) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

mod_spec <- decision_tree(
  tree_depth = tune(),
  min_n = tune()
) |>
  set_engine("rpart") |>
  set_mode("classification")

wf <- workflow() |>
  add_recipe(rec) |>
  add_model(mod_spec)

folds <- vfold_cv(peng_cls, v = 5, strata = y)

metrics_all <- metric_set(accuracy, kap, roc_auc)
```

## One grid, three metrics

`kap` is **Cohen’s kappa**: it subtracts what you would expect from guessing the majority class, so it is often more informative than raw **accuracy** when class counts differ.

```{r}
#| label: m07m-tune
set.seed(42)
tune_res <- tune_grid(
  wf,
  resamples = folds,
  grid = grid_regular(
    tree_depth(range = c(2, 6)),
    min_n(range = c(2, 20)),
    levels = 3
  ),
  metrics = metrics_all
)
```

## What cross-validation reported (mean per setting)

```{r}
#| label: m07m-collect
collect_metrics(tune_res) |>
  dplyr::filter(.metric %in% c("accuracy", "kap", "roc_auc")) |>
  dplyr::arrange(.metric, desc(mean))
```

## Same tuning run — three different “winners”

```{r}
#| label: m07m-select-best
best_acc <- select_best(tune_res, metric = "accuracy")
best_kap <- select_best(tune_res, metric = "kap")
best_auc <- select_best(tune_res, metric = "roc_auc")

best_acc
best_kap
best_auc
```

If the three tibbles differ, that is the lesson: **the metric is part of the scientific question**, not an afterthought.

## Optional — tuning surface (one panel per metric)

`autoplot()` draws **one** metric at a time here; re-run the chunk with `metric = "accuracy"` or `"kap"` to compare shapes.

```{r}
#| label: m07m-autoplot
#| fig-width: 9
#| fig-height: 4.5
autoplot(tune_res, metric = "roc_auc")
```

## Takeaways

- You can **commit to several metrics** before tuning, then **declare** which one picks hyperparameters for this exercise (or use a rule: e.g. maximize **kappa** subject to **sensitivity** ≥ some floor).
- **ROC AUC** ranks risk; **accuracy** counts hard calls; **kappa** penalizes “always guess the common species” shortcuts.
- [Module 08](module-08-penguins-compare-models.qmd) holds the **metric** fixed and swaps **model families**.

## Course links

- **Concepts:** [Module 06 — Metrics & SHAP](module-06-evaluation-and-interpretability.qmd)
- **Day 4 (Thursday) slides:** [Metrics toolbox](../slides/day-04-thursday.html)
