---
title: "Module 04: Modeling pipeline in R"
---

- **Learning objective:** follow a full **classification** workflow in R on **Palmer Penguins** (Adelie vs Gentoo): clean the table, preprocess, pick a model, evaluate honestly, then (below) cross-validate and tune like Day 2 (Tuesday) slides.
- **Key message:** keep **preprocessing, fitting, and scoring** in one reproducible object so every model you compare is treated **fairly** — same **scientific task** (Adelie vs Gentoo) as the [penguin classification notebook](../notebooks/penguins-classification.Rmd), but that notebook uses **classical `glm` / `rpart`** with **all** morphometrics; **this page** uses **bill + island + sex + year** so **tree tuning** shows a real bias–variance tradeoff (flipper and body mass make CV scores too flat). Follow-up labs: [Module 07](module-07-penguins-choose-metrics.qmd) (several metrics while tuning), [Module 08](module-08-penguins-compare-models.qmd) (compare RF / XGBoost / MLP).

## How to read this page

You already know **`lm`** and **`glm`**. Here, R uses a slightly richer vocabulary: a **recipe** is “the list of preprocessing steps,” a **workflow** glues those steps to a **model specification**, and **`tune_grid`** tries several settings using **cross-validation**.

1. **First section** — one **train/test split**, **no tuning**: smallest honest **`last_fit`** story (logistic regression) on the **same `peng_cls` columns** as below.  
2. **Second section** — the **same** Day 2 (Tuesday) walk-through as the slides: tuned **classification tree**, wider `tune_grid`, figures (shared include).

[← Module library (all modules)](index.qmd)

---

## On the slides (Day 2)

| Section | Topic | Slide link |
|---------|--------|------------|
| Part C | Incremental `recipe` → `workflow` → **`tune_grid`** (tree, **accuracy**) | [Part C pipeline](../slides/day-02-tuesday.html#/part-c-pipeline) |
| Source | Same code as this page (canonical include) | [`day02-tidymodels-walkthrough.qmd`](../_includes/day02-tidymodels-walkthrough.qmd) |

Thursday swaps **`spec`** only — [Part B models](../slides/day-04-thursday.html#/part-b-models).

---

## First: one split, one model, no tuning (`last_fit`) {#train-test-last-fit}

Day 2 (Tuesday) slides focus on **cross-validation + tuning** (faster to teach one rhythm in the room). If you want the **smallest** end-to-end `tidymodels` example first — **hold out test rows**, fit on training only, then score — work through this block.

**Predictors (both sections):** **bill length & depth**, **island**, **sex**, **year** — we **drop flipper length and body mass** so the tuned tree section below is not “everything wins at 99%.” The [classification notebook](../notebooks/penguins-classification.Rmd) still uses full morphology for classical `glm` / `rpart` plots.

**Packages:** `tidymodels`, `palmerpenguins`, `dplyr`, `tidyr`.

```{r}
#| label: m04-tt-setup
#| message: false
#| warning: false
library(tidymodels)
library(palmerpenguins)
library(dplyr)
library(tidyr)

peng_cls <- palmerpenguins::penguins |>
  dplyr::filter(species %in% c("Adelie", "Gentoo")) |>
  dplyr::mutate(
    y = factor(species, levels = c("Adelie", "Gentoo")),
    year = as.numeric(year)
  ) |>
  dplyr::select(-species, -flipper_length_mm, -body_mass_g) |>
  tidyr::drop_na()
```

```{r}
#| label: m04-tt-split
set.seed(11)
p_split <- initial_split(peng_cls, prop = 0.75, strata = y)
train <- training(p_split)
test <- testing(p_split)

count(train, y, name = "n_train")
count(test, y, name = "n_test")
```

```{r}
#| label: m04-tt-recipe
rec_tt <- recipe(y ~ ., data = train) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())
```

```{r}
#| label: m04-tt-model
mod_glm <- logistic_reg() |>
  set_engine("glm") |>
  set_mode("classification")

wf_tt <- workflow() |>
  add_recipe(rec_tt) |>
  add_model(mod_glm)
```

```{r}
#| label: m04-tt-last-fit
fit_split <- last_fit(wf_tt, split = p_split, metrics = metric_set(accuracy, roc_auc))
collect_metrics(fit_split)
```

```{r}
#| label: m04-tt-conf-mat
collect_predictions(fit_split) |>
  conf_mat(truth = y, estimate = .pred_class)
```

When you continue to the **canonical pipeline** below, you **redefine `peng_cls` in the include** with the same columns (the chunk is self-contained for slides) and **cross-validation + `tune_grid`** replace this single split — same vocabulary, heavier machinery. Expect **`tree_depth`** and **`min_n`** to move metrics in the tuning heatmap.

---

## Canonical pipeline (Day 2 slides + tuning) {#canonical-pipeline-tuesday}

Read the **concept slide** before each **code chunk** in the include — chunk names match the Day 2 (Tuesday) deck. Tuning uses **`tree_depth` 1–6** and **`min_n` 2–50** (four levels each).

{{< include ../_includes/day02-penguins-setup.qmd >}}

{{< include ../_includes/day02-tidymodels-walkthrough.qmd >}}
