---
title: "Chapter 8: Comparing learners fairly (random forest, XGBoost, small neural net)"
---

This chapter focuses on fair comparison design. We keep folds, recipe, and metrics fixed, then change only the model family.

This is the natural follow-up to [Chapter 4](module-04-pipeline.qmd#canonical-pipeline-tuesday) (tuned tree) and [Chapter 7](module-07-penguins-choose-metrics.qmd) (which metric to optimize). The **Thursday lecture** runs the same comparison on slides with **`metric_set(accuracy)`** ([Part B](../slides/day-04-thursday.html#/part-b-models)); this page also reports **ROC AUC** for side-by-side reading.

Here we **fix** reasonable hyperparameters so the page stays fast — in coursework you could wrap each in `tune_grid` like Module 04.

**Packages:** `tidymodels`, `palmerpenguins`, `dplyr`, `tidyr`, `purrr`, plus **`ranger`** and **`xgboost`** for the engines used below (install from CRAN if `fit_resamples()` asks for them).

[← Chapter 7](module-07-penguins-choose-metrics.qmd) · [Reading companion](index.qmd)

```{r}
#| label: m08c-setup
#| message: false
#| warning: false
library(tidymodels)
library(palmerpenguins)
library(dplyr)
library(tidyr)
library(purrr)

peng_cls <- palmerpenguins::penguins |>
  dplyr::filter(species %in% c("Adelie", "Gentoo")) |>
  dplyr::mutate(
    y = factor(species, levels = c("Adelie", "Gentoo")),
    year = as.numeric(year)
  ) |>
  dplyr::select(-species, -flipper_length_mm, -body_mass_g) |>
  tidyr::drop_na()

rec <- recipe(y ~ ., data = peng_cls) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

folds <- vfold_cv(peng_cls, v = 5, strata = y)
met <- metric_set(roc_auc, accuracy)
```

## Three model specifications (no tuning — fixed settings)

These settings are intentionally reasonable-but-not-optimized so students can focus on **fair comparison design** (same folds, same recipe) before running larger tuning grids.

- RF: many trees, modest `mtry` for robust baseline behavior.
- XGBoost: shallow trees + small learning rate for smoother boosting.
- MLP: small hidden layer and weight decay to limit overfitting on small \(n\).

```{r}
#| label: m08c-specs
rf_spec <- rand_forest(
  mtry = 4,
  trees = 400,
  min_n = 2
) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("classification")

xgb_spec <- boost_tree(
  trees = 200,
  tree_depth = 3,
  learn_rate = 0.03
) |>
  set_engine("xgboost") |>
  set_mode("classification")

mlp_spec <- mlp(
  hidden_units = 8,
  penalty = 0.1,
  epochs = 150
) |>
  set_engine("nnet", trace = FALSE) |>
  set_mode("classification")
```

## Three workflows — same recipe, different engine

```{r}
#| label: m08c-workflows
wf_rf <- workflow() |> add_recipe(rec) |> add_model(rf_spec)
wf_xgb <- workflow() |> add_recipe(rec) |> add_model(xgb_spec)
wf_mlp <- workflow() |> add_recipe(rec) |> add_model(mlp_spec)
```

## Cross-validated resampling (`fit_resamples`)

Each call refits the recipe on **analysis** folds and scores on **assessment** folds — fair side-by-side comparison.

What is held constant: data rows, preprocessing steps, folds, and metrics. What changes: only the model family and hyperparameters.

```{r}
#| label: m08c-fit-resamples
set.seed(7)
rs_rf <- fit_resamples(wf_rf, folds, metrics = met)
rs_xgb <- fit_resamples(wf_xgb, folds, metrics = met)
rs_mlp <- fit_resamples(wf_mlp, folds, metrics = met)
```

## Compare metrics (mean across folds)

Interpretation tip: compare both `mean` and `std_err`. A tiny difference in mean (e.g. 0.01 AUC) with overlapping uncertainty is often not a meaningful winner on small data.

```{r}
#| label: m08c-collect-bind
cmp <- purrr::imap_dfr(
  list(random_forest = rs_rf, xgboost = rs_xgb, mlp_nnet = rs_mlp),
  \(rs, name) collect_metrics(rs) |> dplyr::mutate(model = name)
)

cmp |>
  dplyr::select(model, .metric, mean, std_err) |>
  dplyr::arrange(.metric, dplyr::desc(mean))
```

## Optional — which fold was hardest?

```{r}
#| label: m08c-per-fold
collect_metrics(rs_xgb, summarize = FALSE) |>
  dplyr::select(id, .metric, .estimate) |>
  tidyr::pivot_wider(names_from = .metric, values_from = .estimate)
```

If one fold is consistently harder, inspect whether it has unusual class balance, covariate mix, or missingness patterns before concluding a model is worse.

## Takeaways

- **Same folds + same recipe** means differences come mainly from **model family** and your chosen **settings**, not from a sneaky preprocessing shortcut.
- On **small** datasets, simpler models sometimes **win by noise**; report **uncertainty** (`std_err` from `collect_metrics`) and avoid over-interpreting tiny AUC gaps.
- A small MLP can underperform on tabular data because neural nets are sensitive to feature representation and tuning; with limited \(n\), tree ensembles often provide stronger bias-variance tradeoffs out of the box.
- To go deeper on **metrics** and **explainability**, continue with [Chapter 6](module-06-evaluation-and-interpretability.qmd) and [Chapter 5](module-05-advanced-models.qmd).

## Chapter links

- **Advanced models (concepts):** [Chapter 5](module-05-advanced-models.qmd)
- **Day 2 (Tuesday) / Day 4 (Thursday) slides:** [day-02](../slides/day-02-tuesday.html), [day-04](../slides/day-04-thursday.html)
