---
title: "Solution 4B — Microbiome class imbalance and upsampling"
format:
  html:
    toc: true
    code-tools: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(tidymodels)
  library(themis)
  library(dplyr)
  library(ggplot2)
})
source("_load_microbiome.R")
```

```{r mic-imb}
set.seed(4)
mic <- load_microbiome()
otu_cols <- mic_otu_cols(mic)

mic_imb <- mic |>
  group_by(Label) |>
  filter(Label == "Early" | (Label == "Late" & row_number() <= max(1L, round(0.10 * sum(Label == "Late"))))) |>
  ungroup()

ggplot(mic_imb, aes(Label)) +
  geom_bar() +
  labs(title = "Manufactured imbalance (all Early, ~10% of Late)")
```

```{r specs}
xgb_spec <- boost_tree(trees = 80, tree_depth = 3, learn_rate = 0.08) |>
  set_engine("xgboost") |>
  set_mode("classification")

rec_base <- recipe(Label ~ ., data = mic_imb) |>
  update_role(sample_id, Individual, Sex, Day, new_role = "id") |>
  step_mutate(across(all_of(otu_cols), ~ log1p(.x))) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())

rec_up <- recipe(Label ~ ., data = mic_imb) |>
  update_role(sample_id, Individual, Sex, Day, new_role = "id") |>
  step_upsample(Label) |>
  step_mutate(across(all_of(otu_cols), ~ log1p(.x))) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())

folds_imb <- mic_group_folds(mic_imb, v = 5)
metrics_imb <- metric_set(roc_auc, accuracy, sens, yardstick::spec, pr_auc)
```

```{r resample}
set.seed(11)
rs_no <- fit_resamples(
  workflow() |> add_recipe(rec_base) |> add_model(xgb_spec),
  folds_imb, metrics_imb
)
set.seed(12)
rs_up <- fit_resamples(
  workflow() |> add_recipe(rec_up) |> add_model(xgb_spec),
  folds_imb, metrics_imb
)

bind_rows(
  collect_metrics(rs_no) |> mutate(model = "No upsample"),
  collect_metrics(rs_up) |> mutate(model = "Upsample")
) |>
  filter(.metric %in% c("accuracy", "sens", "pr_auc")) |>
  select(model, .metric, mean, std_err)
```

## Interpretation

Compare **accuracy**, **sensitivity**, and **`pr_auc`** between the two models.

- If metrics were already strong **without** upsampling, say so — upsampling duplicates minority rows in training; it helps when the learner **under-predicts** the rare class (`Late`), not when separation is already easy.
- For a **screening** study, emphasize **sensitivity** and **`pr_auc`** (cost of missing `Late`); for a methods comparison on balanced prevalence, **ROC-AUC** may suffice.
- Tie conclusions to the confusion matrix below (false negatives on `Late`).

```{r conf-mat}
set.seed(13)
split_imb <- initial_split(mic_imb, prop = 0.8, strata = Label)
fit_imb <- fit(
  workflow() |> add_recipe(rec_base) |> add_model(xgb_spec),
  training(split_imb)
)
augment(fit_imb, testing(split_imb)) |>
  conf_mat(truth = Label, estimate = .pred_class)
```

```{r session, echo=FALSE}
sessionInfo()
```
