---
title: "Solution — Day 2 microbiome (tidymodels pipeline)"
format:
  html:
    toc: true
    toc_depth: 3
    code-tools: true
---

Tasks **2.1–2.5** on the [lab exercises page](../index.qmd). Outcome: **`Label`** (Early vs Late). We use **`accuracy`** throughout and compare **`rpart`**, unpenalized logistic regression (`glm`), and **lasso** (`glmnet`) on the same train/test split.

Microbiome helpers (`load_microbiome()`, `mic_otu_cols()`, …) live in [`_load_microbiome.R` on GitHub](https://github.com/s-peischl/ML_SS_2026/blob/main/exercises/solutions/_load_microbiome.R). When you run this file locally, `source()` it from `exercises/solutions/`.

New to `tidymodels`? Use the companion explainer: [Day 2 function guide (plain language)](../tidymodels-day2-guide.qmd).

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 120
)
suppressPackageStartupMessages({
  library(tidymodels)
  library(dplyr)
  library(ggplot2)
  library(tidyr)
  library(patchwork)
})
theme_set(theme_minimal(base_size = 13))
```

```{r setup-visible}
source("_load_microbiome.R")

# One metric only — pass this to last_fit(), tune_grid(), etc.
accuracy_only <- metric_set(accuracy)
```

## 2.1 Recipe

```{r data}
set.seed(7)
mic <- load_microbiome()
otu_cols <- mic_otu_cols(mic)
mic_model <- mic |> select(Label, all_of(otu_cols))

table(mic_model$Label)
```

```{r recipe}
rec <- recipe(Label ~ ., data = mic_model) |>
  step_mutate(across(all_of(otu_cols), ~ log1p(.x))) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())
rec
```

## 2.2 `rpart` tree spec

```{r tree-spec}
tree_spec <- decision_tree(tree_depth = 4, min_n = 10) |>
  set_engine("rpart") |>
  set_mode("classification")
tree_spec
```

## 2.3 Train / test split

We hold out **25%** of rows for a stratified test set. The recipe is **prepped on training rows only** when we `fit()` — that is the leakage-safe pattern from the slides.

```{r split}
split <- initial_split(mic_model, prop = 0.75, strata = Label)
train <- training(split)
test <- testing(split)

tibble(
  set = c("Train", "Test"),
  n = c(nrow(train), nrow(test)),
  pct_late = c(mean(train$Label == "Late"), mean(test$Label == "Late"))
) |>
  knitr::kable(digits = 3, caption = "Split sizes and Late prevalence")
```

```{r split-plot, fig.width=7, fig.height=3.5}
bind_rows(
  train |> mutate(set = "Train"),
  test |> mutate(set = "Test")
) |>
  ggplot(aes(set, fill = Label)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Class balance in train vs test", x = NULL, y = "Proportion")
```

```{r tree-workflow}
wf_tree <- workflow() |> add_recipe(rec) |> add_model(tree_spec)
```

```{r tree-fit}
set.seed(7)
fit_tree <- last_fit(wf_tree, split, metrics = accuracy_only)
collect_metrics(fit_tree)
```

```{r tree-fit-train}
set.seed(7)
fitted_tree <- fit(wf_tree, train)
pred_tree_train <- augment(fitted_tree, train)
pred_tree_test <- augment(fitted_tree, test)

accuracy(pred_tree_train, truth = Label, estimate = .pred_class)
accuracy(pred_tree_test, truth = Label, estimate = .pred_class)
```

## 2.4 Logistic regression (unpenalized `glm`)

```{r glm-spec}
log_spec <- logistic_reg() |>
  set_engine("glm") |>
  set_mode("classification")
wf_log <- workflow() |> add_recipe(rec) |> add_model(log_spec)
```

```{r glm-fit}
set.seed(7)
fit_log <- last_fit(wf_log, split, metrics = accuracy_only)
collect_metrics(fit_log)
```

Fit on the **training set** so we can score train and test explicitly with `yardstick::accuracy()`:

```{r glm-fit-train}
set.seed(7)
fitted_log <- fit(wf_log, train)
pred_train <- augment(fitted_log, train)
pred_test <- augment(fitted_log, test)

# yardstick::accuracy() — truth and estimate column names must match
accuracy(pred_train, truth = Label, estimate = .pred_class)
accuracy(pred_test, truth = Label, estimate = .pred_class)
```

## 2.5 Lasso logistic regression (`glmnet`)

Same recipe, but **`mixture = 1`** is lasso and **`penalty`** (λ) is tuned by **5-fold CV on the training set only** — the Day 1 idea inside a `tidymodels` workflow.

```{r lasso-spec}
lasso_spec <- logistic_reg(penalty = tune(), mixture = 1) |>
  set_engine("glmnet") |>
  set_mode("classification")
wf_lasso <- workflow() |> add_recipe(rec) |> add_model(lasso_spec)
```

```{r lasso-tune}
set.seed(7)
folds <- vfold_cv(train, v = 5, strata = Label)
grid <- grid_regular(penalty(range = c(-4, 0)), levels = 10)

tuned_lasso <- tune_grid(
  wf_lasso,
  resamples = folds,
  grid = grid,
  metrics = accuracy_only
)

select_best(tuned_lasso, metric = "accuracy")
```

```{r lasso-final}
wf_lasso_final <- finalize_workflow(wf_lasso, select_best(tuned_lasso, metric = "accuracy"))

set.seed(7)
fit_lasso <- last_fit(wf_lasso_final, split, metrics = accuracy_only)
collect_metrics(fit_lasso)
```

## 2.6 Compare models on the test set (accuracy)

All three models use the **same** split and recipe. **`last_fit()`** fits on `training(split)` and reports **`accuracy`** on `testing(split)`:

```{r compare}
cmp <- bind_rows(
  collect_metrics(fit_tree) |> mutate(model = "rpart (tree)"),
  collect_metrics(fit_log) |> mutate(model = "Logistic (glm)"),
  collect_metrics(fit_lasso) |> mutate(model = "Lasso (glmnet)")
) |>
  select(model, .metric, .estimate)

majority_acc <- max(prop.table(table(test$Label)))

cmp |>
  bind_rows(tibble(model = "Majority class", .metric = "accuracy", .estimate = majority_acc)) |>
  knitr::kable(digits = 3, caption = "Held-out test accuracy (75/25 split)")
```

```{r compare-plot, fig.width=7, fig.height=4}
cmp |>
  ggplot(aes(reorder(model, .estimate), .estimate, fill = model)) +
  geom_col(show.legend = FALSE) +
  geom_hline(yintercept = majority_acc, linetype = 2, linewidth = 0.7) +
  coord_flip() +
  labs(
    title = "Test accuracy — rpart vs glm vs lasso",
    subtitle = "Dashed line = majority-class baseline on the test set",
    x = NULL,
    y = "Accuracy"
  )
```

## 2.7 Train vs test accuracy — all three models

Unpenalized logistic regression has **more OTU predictors than training samples** (`r length(otu_cols)` OTUs, `r nrow(train)` train rows). `glm` may not converge and can achieve **perfect training accuracy** while doing **worse than the majority class** on the test set. **`rpart`** and **lasso** handle high-dimensional predictors more safely (splits vs penalised coefficients).

```{r train-test-compare}
fitted_lasso <- fit(wf_lasso_final, train)
pred_lasso_train <- augment(fitted_lasso, train)
pred_lasso_test <- augment(fitted_lasso, test)

acc_value <- function(df) {
  accuracy(df, truth = Label, estimate = .pred_class)$.estimate
}

acc_tbl <- tibble(
  model = rep(c("rpart (tree)", "Logistic (glm)", "Lasso (glmnet)"), each = 2),
  set = rep(c("Train", "Test"), times = 3),
  accuracy = c(
    acc_value(pred_tree_train), acc_value(pred_tree_test),
    acc_value(pred_train), acc_value(pred_test),
    acc_value(pred_lasso_train), acc_value(pred_lasso_test)
  )
)

knitr::kable(acc_tbl, digits = 3, caption = "Accuracy via yardstick::accuracy()")
```

```{r train-test-plot, fig.width=7, fig.height=4}
acc_tbl |>
  mutate(
    model = factor(
      model,
      levels = c("rpart (tree)", "Logistic (glm)", "Lasso (glmnet)")
    ),
    set = factor(set, levels = c("Train", "Test"))
  ) |>
  ggplot(aes(set, accuracy, fill = model)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_hline(yintercept = majority_acc, linetype = 2) +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0, 0.03))) +
  labs(
    title = "Training vs test accuracy",
    subtitle = "Dashed line = majority-class accuracy on the test set",
    x = NULL,
    y = "Accuracy",
    fill = NULL
  )
```

```{r confusion-both, fig.width=10, fig.height=4}
conf_to_df <- function(x, model) {
  as.data.frame(x$table) |>
    mutate(model = model)
}

bind_rows(
  conf_to_df(conf_mat(pred_tree_test, truth = Label, estimate = .pred_class), "rpart (tree)"),
  conf_to_df(conf_mat(pred_test, truth = Label, estimate = .pred_class), "Logistic (glm)"),
  conf_to_df(conf_mat(pred_lasso_test, truth = Label, estimate = .pred_class), "Lasso (glmnet)")
) |>
  ggplot(aes(Prediction, Truth, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "white", fontface = "bold") +
  facet_wrap(~ model) +
  scale_fill_gradient(low = "grey85", high = "steelblue") +
  labs(title = "Test-set confusion matrices", x = "Predicted", y = "Observed")
```

### Takeaways

1. **Specify the metric:** `accuracy_only <- metric_set(accuracy)` for resampling; `accuracy(data, truth = Label, estimate = .pred_class)` on prediction tables.
2. **Unpenalized glm + p ≫ n** → unreliable test accuracy; **`rpart`** and **lasso** are safer choices on the same OTU matrix.
3. **`rpart`** uses recursive splits (morning Part B); **lasso** uses penalised coefficients (Monday's `glmnet`, now in a workflow).
4. **Repeated measures:** `initial_split()` can put the same mouse in train and test; Day 4 uses `group_vfold_cv(group = Individual)`.

**Leakage reminder:** prep the recipe inside `fit()` / `workflow` on training data only; never normalize on the full table before splitting.

```{r session, echo=FALSE}
sessionInfo()
```
