---
title: "Solution — Day 2 microbiome (tidymodels pipeline)"
format:
  html:
    toc: true
    toc_depth: 3
    code-tools: true
---

Tasks **2.1–2.5** on the [lab exercises page](../index.qmd). Outcome: **`Label`** (Early vs Late). Same **`recipe` → `workflow` → fit** pattern as the Palmer Penguins deck; here we also contrast **training-set** vs **held-out test** metrics for logistic regression.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 120
)
suppressPackageStartupMessages({
  library(tidymodels)
  library(dplyr)
  library(ggplot2)
  library(tidyr)
  library(patchwork)
})
theme_set(theme_minimal(base_size = 13))
source("_load_microbiome.R")

class_metrics <- metric_set(accuracy, precision, recall, roc_auc)
late_event <- "second" # Label levels: Early, Late → Late is the positive class
```

## 2.1 Recipe

```{r data}
set.seed(7)
mic <- load_microbiome()
otu_cols <- mic_otu_cols(mic)
mic_model <- mic |> select(Label, all_of(otu_cols))

table(mic_model$Label)
```

```{r recipe}
rec <- recipe(Label ~ ., data = mic_model) |>
  step_mutate(across(all_of(otu_cols), ~ log1p(.x))) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())
rec
```

## 2.2 Tree spec

```{r tree-spec}
tree_spec <- decision_tree(tree_depth = 4, min_n = 10) |>
  set_engine("rpart") |>
  set_mode("classification")
tree_spec
```

## 2.3 Train / test split

We hold out **25%** of rows for a stratified test set. The recipe is **prepped on training rows only** when we `fit()` — that is the leakage-safe pattern from the slides.

```{r split}
split <- initial_split(mic_model, prop = 0.75, strata = Label)
train <- training(split)
test <- testing(split)

tibble(
  set = c("Train", "Test"),
  n = c(nrow(train), nrow(test)),
  pct_late = c(mean(train$Label == "Late"), mean(test$Label == "Late"))
) |>
  knitr::kable(digits = 3, caption = "Split sizes and Late prevalence")
```

```{r split-plot, fig.width=7, fig.height=3.5}
bind_rows(
  train |> mutate(set = "Train"),
  test |> mutate(set = "Test")
) |>
  ggplot(aes(set, fill = Label)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Class balance in train vs test", x = NULL, y = "Proportion")
```

```{r tree-workflow}
wf_tree <- workflow() |> add_recipe(rec) |> add_model(tree_spec)
```

```{r tree-fit}
set.seed(7)
fit_tree <- last_fit(wf_tree, split, metrics = class_metrics)
collect_metrics(fit_tree)
```

## 2.4 Logistic workflow

```{r glm-spec}
log_spec <- logistic_reg() |>
  set_engine("glm") |>
  set_mode("classification")
wf_log <- workflow() |> add_recipe(rec) |> add_model(log_spec)
```

```{r glm-fit}
set.seed(7)
fit_log <- last_fit(wf_log, split, metrics = class_metrics)
collect_metrics(fit_log)
```

Fit once on the **training set** so we can score the same model on train and test rows explicitly.

```{r glm-fit-train}
set.seed(7)
fitted_log <- fit(wf_log, train)
pred_train <- augment(fitted_log, train)
pred_test <- augment(fitted_log, test)
```

## 2.5 Compare models on the test set

```{r compare}
cmp <- bind_rows(
  collect_metrics(fit_tree) |> mutate(model = "Decision tree (rpart)"),
  collect_metrics(fit_log) |> mutate(model = "Logistic regression")
) |>
  filter(.metric %in% c("accuracy", "precision", "roc_auc")) |>
  select(model, .metric, .estimate)

knitr::kable(cmp, digits = 3, caption = "Held-out test metrics (last_fit)")
```

```{r compare-plot, fig.width=8, fig.height=4}
cmp |>
  mutate(.metric = recode(.metric, roc_auc = "ROC AUC")) |>
  ggplot(aes(reorder(model, .estimate), .estimate, fill = model)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ .metric, scales = "free_x") +
  coord_flip() +
  labs(title = "Test-set performance (75/25 split)", x = NULL, y = NULL)
```

## 2.6 Logistic regression — in-sample vs test precision

**Precision** (positive predictive value) answers: among samples we call **Late**, what fraction truly are Late? Training precision is computed on the **same rows used to fit** the model; test precision uses **held-out** rows only. A large gap is a sign of **optimism** — the model may be memorising noise in the training OTUs.

```{r precision-compare}
score_set <- function(df, set_label) {
  tibble(
    set = set_label,
    accuracy = accuracy_vec(df$Label, df$.pred_class),
    precision = precision_vec(df$Label, df$.pred_class, event_level = late_event),
    recall = recall_vec(df$Label, df$.pred_class, event_level = late_event)
  )
}

log_scores <- bind_rows(
  score_set(pred_train, "Train (in-sample)"),
  score_set(pred_test, "Test (held out)")
)

knitr::kable(log_scores, digits = 3, caption = "Logistic regression — train vs test")
```

```{r precision-plot, fig.width=7, fig.height=4}
log_scores |>
  pivot_longer(c(accuracy, precision, recall), names_to = "metric", values_to = "value") |>
  mutate(
    metric = recode(metric,
      accuracy = "Accuracy",
      precision = "Precision (Late)",
      recall = "Recall (Late)"
    ),
    set = factor(set, levels = c("Train (in-sample)", "Test (held out)"))
  ) |>
  ggplot(aes(set, value, fill = set)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ metric) +
  scale_fill_manual(values = c("Train (in-sample)" = "steelblue", "Test (held out)" = "darkorange")) +
  scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Logistic regression: training vs test metrics", x = NULL, y = NULL)
```

```{r precision-gap, echo=FALSE, results='asis'}
train_prec <- log_scores$precision[log_scores$set == "Train (in-sample)"]
test_prec <- log_scores$precision[log_scores$set == "Test (held out)"]
cat(
  sprintf(
    "Precision drops from **%.3f** on the training set to **%.3f** on the test set (Δ = %.3f). ",
    train_prec, test_prec, train_prec - test_prec
  ),
  "That is expected when **p ≫ n**: a linear model on hundreds of OTUs can fit training noise even after preprocessing. ",
  "Accuracy and recall show the same train-vs-test pattern — check the table above.\n",
  sep = ""
)
```

```{r confusion-logistic, fig.width=9, fig.height=4}
conf_mat_train <- conf_mat(pred_train, truth = Label, estimate = .pred_class)
conf_mat_test <- conf_mat(pred_test, truth = Label, estimate = .pred_class)

bind_rows(
  as.data.frame(conf_mat_train$table) |> mutate(set = "Train"),
  as.data.frame(conf_mat_test$table) |> mutate(set = "Test")
) |>
  ggplot(aes(Prediction, Truth, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "white", fontface = "bold") +
  facet_wrap(~ set) +
  scale_fill_gradient(low = "grey85", high = "steelblue") +
  labs(title = "Logistic regression confusion matrices", x = "Predicted", y = "Observed")
```

### Takeaways

1. **Fair model comparison** uses the **same** train/test split and the **same** recipe (tasks 2.3–2.5).
2. **In-sample precision** is not a honest estimate of future performance — always report **test** (or CV) metrics when you tune or select models.
3. **Repeated measures:** rows from the same `Individual` can land in both train and test with `initial_split()`, which can **inflate** test precision. Research workflows use `group_vfold_cv(group = Individual)` (Day 4 solution).

**Leakage reminder:** prep the recipe inside `fit()` / `workflow` on training data only; never normalize on the full table before splitting.

```{r session, echo=FALSE}
sessionInfo()
```
