---
title: "Day 4 (Thursday): Models, Preprocessing, Metrics, and Importance"
metadata-files: [_reveal-config.yml]
---

## Learning objectives

- Swap **`parsnip` specs** (bagging, boosting, neural nets) on the **same recipe + workflow** from Tuesday
- Compare **many models** fairly with **`fit_resamples()`** using **accuracy** until imbalance
- Add **PCA**, **imputation**, and **resampling** as **recipe steps** inside grouped CV
- Choose **metrics beyond accuracy** (sensitivity, PR-AUC, confusion matrices) for imbalanced data
- Read **variable importance** without causal over-interpretation

## Navigation

- [Teaching Hub](../index.qmd)
- [Curriculum Outline](../curriculum-outline.qmd)
- [Module 05 — Advanced models](../modules/module-05-advanced-models.qmd)
- [Module 08 — Compare learners](../modules/module-08-penguins-compare-models.qmd)
- [Back: Day 2 (Tuesday) deck](day-02-tuesday.qmd)

# Part A — Recap (Tuesday’s pipeline)

## What you already have

- **`rec_base`** — `step_zv` → `step_dummy` → `step_normalize` on **`peng_bal`**
- **`workflow()`** + **`decision_tree` / `rpart`** tuned on Tuesday
- **`vfold_cv(..., strata = y)`** and **`metric_set(accuracy)`** for Parts B–D today

```{r}
#| label: day04-setup
#| include: false
suppressPackageStartupMessages({
  library(tidymodels)
  library(palmerpenguins)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(rlang)
  library(vip)
  library(purrr)
  library(themis)
})
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  out.width = "100%",
  dpi = 120
)

peng_cls <- palmerpenguins::penguins |>
  filter(species %in% c("Adelie", "Gentoo")) |>
  mutate(
    y = factor(species, levels = c("Adelie", "Gentoo")),
    year = as.numeric(year)
  ) |>
  dplyr::select(-species, -flipper_length_mm, -body_mass_g) |>
  tidyr::drop_na()

folds <- vfold_cv(peng_cls, v = 5, strata = y)
metrics_acc <- metric_set(accuracy)

peng_bal <- peng_cls

rec_base <- recipe(y ~ ., data = peng_bal) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

tree_spec <- decision_tree(tree_depth = 4, min_n = 10) |>
  set_engine("rpart") |>
  set_mode("classification")

wf_tree <- workflow() |>
  add_recipe(rec_base) |>
  add_model(tree_spec)

set.seed(6)
rs_tree <- fit_resamples(wf_tree, folds, metrics = metrics_acc)

# Teaching scenarios (artificial problems — document real prevalence on the data card)
set.seed(4)
peng_imb <- peng_bal |>
  group_by(y) |>
  filter(y == "Adelie" | (y == "Gentoo" & dplyr::row_number() <= 10L)) |>
  ungroup()

peng_na <- peng_bal
na_bill <- sample(which(!is.na(peng_na$bill_length_mm)), size = 18L)
peng_na$bill_length_mm[na_bill] <- NA
na_sex <- sample(which(!is.na(peng_na$sex)), size = 5L)
peng_na$sex[na_sex] <- NA

peng_pca <- palmerpenguins::penguins |>
  filter(species %in% c("Adelie", "Gentoo")) |>
  mutate(
    y = factor(species, levels = c("Adelie", "Gentoo")),
    year = as.numeric(year)
  ) |>
  dplyr::select(-species) |>
  tidyr::drop_na()

folds_imb <- vfold_cv(peng_imb, v = 5, strata = y)
folds_pca <- vfold_cv(peng_pca, v = 5, strata = y)
metrics_cls_imb <- metric_set(accuracy, sens, yardstick::spec, roc_auc, pr_auc)

plot_class_balance <- function(df, title = "Class counts") {
  ggplot(df, aes(y, fill = y)) +
    geom_bar(show.legend = FALSE) +
    geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.3, size = 4) +
    scale_fill_manual(values = c(Adelie = "turquoise", Gentoo = "magenta")) +
    theme_minimal() +
    labs(title = title, x = "Species (y)", y = "n")
}

plot_na_highlight <- function(df, xvar, yvar, title = NULL) {
  df <- df |>
    mutate(
      missing_x = is.na(.data[[xvar]]),
      x_plot = ifelse(missing_x, stats::median(.data[[xvar]], na.rm = TRUE), .data[[xvar]])
    )
  ggplot(df, aes(x = x_plot, y = .data[[yvar]])) +
    geom_point(
      data = dplyr::filter(df, !missing_x),
      aes(color = y),
      size = 2.4
    ) +
    geom_point(
      data = dplyr::filter(df, missing_x),
      aes(x = x_plot, y = .data[[yvar]]),
      shape = 4,
      color = "firebrick",
      size = 3,
      stroke = 1.2,
      inherit.aes = FALSE
    ) +
    scale_color_manual(values = c(Adelie = "black", Gentoo = "gray40"), name = "Species") +
    theme_minimal() +
    labs(
      title = title,
      x = paste0(xvar, " (× = missing)"),
      y = yvar,
      caption = "Red × marks rows with NA on the x-axis"
    )
}

metrics_summary_table <- function(fit_res, label) {
  collect_metrics(fit_res) |>
    dplyr::filter(.metric %in% c("accuracy", "sens", "spec", "roc_auc", "pr_auc")) |>
    dplyr::select(.metric, mean, std_err) |>
    dplyr::mutate(model = label) |>
    dplyr::relocate(model)
}

compare_metrics_tbl <- function(...) {
  tab <- purrr::imap_dfr(rlang::list2(...), ~ metrics_summary_table(.x, .y)) |>
    dplyr::select(model, .metric, mean, std_err) |>
    dplyr::mutate(cell = sprintf("%.3f (%.3f)", mean, std_err)) |>
    dplyr::select(model, .metric, cell) |>
    tidyr::pivot_wider(names_from = model, values_from = cell)
  knitr::kable(tab, caption = "Cross-validated metrics: mean (std err)")
}

plot_prob_boundary <- function(
    model,
    data,
    f1 = "bill_length_mm",
    f2 = "bill_depth_mm",
    title = NULL
) {
  r1 <- range(data[[f1]], na.rm = TRUE)
  r2 <- range(data[[f2]], na.rm = TRUE)
  grid <- expand.grid(
    seq(r1[1] - 0.5, r1[2] + 0.5, length.out = 120),
    seq(r2[1] - 0.5, r2[2] + 0.5, length.out = 120)
  )
  names(grid) <- c(f1, f2)
  for (nm in setdiff(names(data), c(f1, f2, "y"))) {
    v <- data[[nm]]
    grid[[nm]] <- if (is.numeric(v)) stats::median(v, na.rm = TRUE) else {
      tab <- table(v)
      names(tab)[which.max(tab)]
    }
  }
  grid$prob <- predict(model, new_data = grid, type = "prob")$.pred_Gentoo
  ggplot(data, aes(!!sym(f1), !!sym(f2))) +
    geom_raster(
      data = grid,
      aes(!!sym(f1), !!sym(f2), fill = prob),
      interpolate = TRUE,
      inherit.aes = FALSE
    ) +
    geom_point(aes(color = y), size = 2.2) +
    scale_fill_gradient(low = "turquoise", high = "magenta", name = "P(Gentoo)") +
    scale_color_manual(values = c(Adelie = "black", Gentoo = "white"), name = "Species") +
    theme_minimal() +
    labs(title = title, x = f1, y = f2)
}

save_day04_composite <- function(model, data, title, chunk_id, f1 = "bill_length_mm", f2 = "bill_depth_mm") {
  path <- paste0(knitr::fig_chunk(chunk_id), ".png")
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  if (requireNamespace("ragg", quietly = TRUE)) {
    ragg::agg_png(path, width = 10, height = 5, units = "in", res = 120)
  } else {
    png(path, width = 10, height = 5, units = "in", res = 120)
  }
  on.exit(dev.off(), add = TRUE)
  layout(matrix(1:2, nrow = 1, ncol = 2))
  par(mar = c(1, 1, 3, 1))
  if (inherits(model, "workflow")) {
    fit_obj <- extract_fit_parsnip(model)
    if (inherits(fit_obj$fit, "rpart")) {
      rpart.plot::rpart.plot(fit_obj$fit, type = 4, extra = 104, main = title, cex = 0.8)
    } else {
      plot(1, type = "n", xlab = "", ylab = "", main = title)
      text(1, 1, "Tree plot: rpart engine only", cex = 0.9)
    }
  } else {
    plot(1, type = "n", xlab = "", ylab = "", main = title)
    text(1, 1, "See boundary panel", cex = 0.9)
  }
  print(plot_prob_boundary(model, data, f1, f2))
  path
}
```

## Teaching scenarios (Part C extensions)

We deliberately stress the penguins in three ways — document real prevalence and missingness on a **data card** in production; do not manufacture problems silently.

```{r}
#| label: day04-scenarios-show
#| echo: true
#| message: false
dplyr::count(peng_imb, y, name = "n")
sum(is.na(peng_na$bill_length_mm))
ncol(peng_pca)
```

| Object | Stress | Recipe extension (Part C) |
|--------|--------|---------------------------|
| **`peng_imb`** | Rare Gentoo (~10 vs ~150 Adelie) | `step_upsample(y)` — when it helps |
| **`peng_na`** | Missing bill length and some `sex` | `step_impute_*()` |
| **`peng_pca`** | Extra size measures (correlated) | `step_pca()` |

# Part B — Many models (accuracy only) {#part-b-models}

**Rule:** keep **`add_recipe(rec_base)`** and **`folds`** — change **`add_model(spec)`** only.

## Model catalog and engines

**Same `rec_base` from Tuesday** — change **`spec`** (and `set_engine()`); keep **`workflow()`** + **`folds`**.

| Model spec | Typical tuned knobs | Common engines |
|------------|---------------------|----------------|
| `decision_tree()` | `tree_depth`, `min_n`, `cost_complexity` | `rpart`, `C5.0` |
| `rand_forest()` | `mtry`, `min_n`, `trees` | `ranger`, `randomForest` |
| `boost_tree()` | `trees`, `tree_depth`, `learn_rate` | `xgboost`, `lightgbm`, `C5.0` |
| `logistic_reg()` | `penalty` | `glm`, `glmnet` |
| `linear_reg()` | `penalty` | `lm`, `glmnet` |
| `mlp()` | `hidden_units`, `penalty`, `epochs` | `nnet`, `keras` |
| `svm_rbf()` | `cost`, `rbf_sigma` | `kernlab` |
| `nearest_neighbor()` | `neighbors` | `kknn` |

- **`tune()`** marks hyperparameters for **`tune_grid()`**.
- Lookup: [tidymodels — find models](https://www.tidymodels.org/find/parsnip/) · [Engines at a glance](https://www.tidymodels.org/find/parsnip/#engines-at-a-glance).
- More depth: [Module 05](../modules/module-05-advanced-models.qmd) · [Module 08](../modules/module-08-penguins-compare-models.qmd).

## Parsnip: type → engine → mode

Every spec needs: **model type** → **`set_engine()`** → **`set_mode("classification")`** (or regression).

```{r}
#| label: day04-catalog-dials
#| echo: true
#| message: false
rf_tune_spec <- rand_forest(trees = tune(), mtry = tune()) |>
  set_engine("ranger") |>
  set_mode("classification")
extract_parameter_set_dials(rf_tune_spec)
```

## Same recipe, swap the `spec`

Part A gave **`wf_tree`** (`rec_base` + `tree_spec`). Below: **random forest** (bagging) and **XGBoost** (boosting) on the same **`folds`**.

## Bagging: random forests

- **Many trees** on **bootstrap** samples of rows; each split sees only a **random subset** of predictors → lower variance than one deep tree.
- **`rand_forest(mtry = …, trees = …, min_n = …) |> set_engine("ranger")`**

## Boosting: gradient boosted trees

- Trees are added **sequentially**; each new tree targets **errors left by the ensemble** so far.
- **`boost_tree(trees = …, tree_depth = …, learn_rate = …) |> set_engine("xgboost")`**
- **`learn_rate`**: shrink each tree’s contribution (smaller → more trees needed, often smoother).

## Forest and boosting specs (code)

```{r}
#| label: day04-ensemble-specs
#| echo: true
#| message: false
rf_spec <- rand_forest(mtry = 3, trees = 400, min_n = 2) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("classification")
xgb_spec <- boost_tree(trees = 150, tree_depth = 3, learn_rate = 0.05) |>
  set_engine("xgboost") |>
  set_mode("classification")

list(tree = tree_spec, rf = rf_spec, xgb = xgb_spec)
```

```{r}
#| label: day04-ensemble-prep
#| echo: true
#| message: false
wf_rf <- workflow() |> add_recipe(rec_base) |> add_model(rf_spec)
wf_xgb <- workflow() |> add_recipe(rec_base) |> add_model(xgb_spec)
```

### Cross-validated ROC AUC (code + table)

```{r}
#| label: day04-ensemble-resamples
#| echo: true
#| message: false
set.seed(7)
rs_rf <- fit_resamples(wf_rf, folds, metrics = metrics_acc)
rs_xgb <- fit_resamples(wf_xgb, folds, metrics = metrics_acc)

cmp_ens <- bind_rows(
  collect_metrics(rs_tree) |> mutate(model = "Single tree"),
  collect_metrics(rs_rf) |> mutate(model = "Random forest"),
  collect_metrics(rs_xgb) |> mutate(model = "XGBoost")
) |>
  filter(.metric == "accuracy")

cmp_ens |>
  select(model, mean, std_err) |>
  arrange(desc(mean))
```

```{r}
#| label: day04-ensemble-auc-bar
#| echo: true
#| fig-width: 8
#| fig-height: 4
cmp_ens |>
  ggplot(aes(reorder(model, mean), mean, fill = model)) +
  geom_col(show.legend = FALSE) +
  geom_errorbar(aes(ymin = mean - std_err, ymax = mean + std_err), width = 0.12) +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Same folds, same recipe — different spec (accuracy)",
    x = NULL,
    y = "Mean accuracy"
  )
```

Fixed settings here for speed — in projects, wrap each spec in **`tune_grid()`** like [Module 04](../modules/module-04-pipeline.qmd#canonical-pipeline-tuesday).

## Boosting: watch the boundary sharpen

We fix **`tree_depth = 2`** and **`learn_rate = 0.3`**, then increase **`trees`** from 1 to 6. Each round adds small trees that correct leftover mistakes — the **Gentoo probability surface** becomes more **nonlinear**.

```{r}
#| label: day04-boost-spec-show
#| echo: true
#| message: false
boost_tree(trees = 6, tree_depth = 2, learn_rate = 0.3) |>
  set_engine("xgboost") |>
  set_mode("classification")
```

```{r}
#| label: day04-boost-boundary-prep
#| echo: false
#| results: hide
boost_trees <- 1:6
for (i in boost_trees) {
  spec_i <- boost_tree(
    trees = i,
    tree_depth = 2,
    learn_rate = 0.3
  ) |>
    set_engine("xgboost") |>
    set_mode("classification")
  wf_i <- workflow() |>
    add_recipe(rec_base) |>
    add_model(spec_i)
  assign(paste0("boost_fit_", i), fit(wf_i, peng_cls))
}
```

## Boosting round 1

```{r}
#| label: day04-boost-step-1
#| echo: false
#| fig-width: 9
#| fig-height: 5
print(
  plot_prob_boundary(
    get("boost_fit_1"), peng_cls,
    title = "XGBoost after 1 tree (bill length × bill depth)"
  )
)
```

One shallow tree: rough split — mostly **underfits** the Gentoo cloud.

## Boosting round 2

```{r}
#| label: day04-boost-step-2
#| echo: false
#| fig-width: 9
#| fig-height: 5
print(
  plot_prob_boundary(
    get("boost_fit_2"), peng_cls,
    title = "XGBoost after 2 trees"
  )
)
```

## Boosting round 3

```{r}
#| label: day04-boost-step-3
#| echo: false
#| fig-width: 9
#| fig-height: 5
print(
  plot_prob_boundary(
    get("boost_fit_3"), peng_cls,
    title = "XGBoost after 3 trees"
  )
)
```

## Boosting round 4

```{r}
#| label: day04-boost-step-4
#| echo: false
#| fig-width: 9
#| fig-height: 5
print(
  plot_prob_boundary(
    get("boost_fit_4"), peng_cls,
    title = "XGBoost after 4 trees"
  )
)
```

## Boosting round 5

```{r}
#| label: day04-boost-step-5
#| echo: false
#| fig-width: 9
#| fig-height: 5
print(
  plot_prob_boundary(
    get("boost_fit_5"), peng_cls,
    title = "XGBoost after 5 trees"
  )
)
```

## Boosting round 6

```{r}
#| label: day04-boost-step-6
#| echo: false
#| fig-width: 9
#| fig-height: 5
print(
  plot_prob_boundary(
    get("boost_fit_6"), peng_cls,
    title = "XGBoost after 6 trees"
  )
)
```

## Bagging vs boosting (same bill plane)

```{r}
#| label: day04-rf-boundary-prep
#| echo: false
#| results: hide
rf_boundary_fit <- fit(wf_rf, peng_bal)
```

```{r}
#| label: day04-rf-boundary
#| echo: false
#| fig-width: 9
#| fig-height: 5
print(
  plot_prob_boundary(
    rf_boundary_fit, peng_bal,
    title = "Random forest — smoother averaged surface"
  )
)
```

Contrast **bagging**: a random forest **averages** many trees — boundaries are often **smoother** than boosting’s **staged** corrections (previous slide).

## Bagging vs boosting (side by side)

```{r}
#| label: day04-rf-xgb-compare
#| echo: false
#| fig-width: 11
#| fig-height: 4.5
p_rf <- plot_prob_boundary(
  rf_boundary_fit, peng_bal,
  title = "Random forest (bagging)"
)
p_xgb <- plot_prob_boundary(
  get("boost_fit_6"), peng_cls,
  title = "XGBoost after 6 trees"
)
if (requireNamespace("patchwork", quietly = TRUE)) {
  print(p_rf + p_xgb + patchwork::plot_layout(ncol = 2))
} else {
  print(p_rf)
  print(p_xgb)
}
```

## Neural network (small MLP) — API literacy

- **`mlp(hidden_units = …, penalty = …, epochs = …) |> set_engine("nnet")`** — one hidden layer, weight decay, counts passes through the data.
- **Requires scaling** (already in `rec_base` via `step_normalize`).
- On **small tabular** data, trees often match or beat MLPs — still worth knowing the **same workflow** interface.

```{r}
#| label: day04-mlp-spec
#| echo: true
mlp_spec <- mlp(hidden_units = 8, penalty = 0.1, epochs = 150) |>
  set_engine("nnet", trace = FALSE) |>
  set_mode("classification")

wf_mlp <- workflow() |>
  add_recipe(rec_base) |>
  add_model(mlp_spec)
```

```{r}
#| label: day04-mlp-resamples
#| echo: true
#| message: false
set.seed(9)
rs_mlp <- fit_resamples(wf_mlp, folds, metrics = metrics_acc)
mlp_cmp <- bind_rows(
  collect_metrics(rs_rf) |> mutate(model = "Random forest"),
  collect_metrics(rs_xgb) |> mutate(model = "XGBoost"),
  collect_metrics(rs_mlp) |> mutate(model = "MLP (nnet)")
) |>
  filter(.metric == "accuracy")

mlp_cmp |>
  select(model, .metric, mean, std_err)
```

```{r}
#| label: day04-mlp-bar
#| echo: true
#| fig-width: 8
#| fig-height: 4
mlp_cmp |>
  ggplot(aes(model, mean, fill = model)) +
  geom_col(show.legend = FALSE) +
  geom_errorbar(aes(ymin = mean - std_err, ymax = mean + std_err), width = 0.12) +
  theme_minimal() +
  labs(title = "Same recipe — accuracy across specs", x = NULL, y = "Mean accuracy")
```

Full three-way comparison: [Module 08](../modules/module-08-penguins-compare-models.qmd).

# Part C — Dimension reduction (accuracy only) {#part-c-pca}

## Extension: PCA inside the recipe (`peng_pca`)

Add **`step_pca()`** to the Part A recipe chain — fit on training folds only, same as `step_normalize()`.

## PCA motivation: correlated measurements

For PCA we add back **flipper length** and **body mass** — bills and size track together on penguins.

```{r}
#| label: day04-pca-corr
#| echo: true
#| fig-width: 7
#| fig-height: 5
num_cols <- c("bill_length_mm", "bill_depth_mm", "flipper_length_mm", "body_mass_g", "year")
cor_pca <- cor(peng_pca[, num_cols])
cor_df <- as.data.frame(as.table(cor_pca)) |>
  dplyr::rename(var1 = Var1, var2 = Var2, r = Freq)
ggplot(cor_df, aes(var1, var2, fill = r)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick", midpoint = 0, limit = c(-1, 1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_fixed() +
  labs(title = "Correlated morphometrics (why PCA?)", fill = "r")
```

## Recipe with `step_pca()`

```{r}
#| label: day04-pca-recipe-code
#| echo: true
rec_pca <- recipe(y ~ ., data = peng_pca) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_pca(all_numeric_predictors(), num_comp = 2)

rec_pca
```

```{r}
#| label: day04-pca-variance
#| echo: false
#| fig-width: 7
#| fig-height: 4
num_scaled <- peng_pca |>
  dplyr::select(dplyr::all_of(num_cols)) |>
  scale()
pc_fit <- prcomp(num_scaled, center = FALSE, scale. = FALSE)
var_pct <- pc_fit$sdev^2 / sum(pc_fit$sdev^2) * 100
var_df <- tibble(PC = paste0("PC", seq_along(var_pct)), variance_pct = var_pct)
ggplot(var_df, aes(PC, variance_pct)) +
  geom_col(fill = "steelblue") +
  theme_minimal() +
  labs(
    title = "Variance explained (same numerics as step_pca)",
    x = NULL,
    y = "% variance"
  )
```

## Model with PCA vs without (same folds)

```{r}
#| label: day04-pca-prep-models
#| echo: false
#| results: hide
xgb_pca_spec <- boost_tree(trees = 100, tree_depth = 3, learn_rate = 0.05) |>
  set_engine("xgboost") |>
  set_mode("classification")

rec_no_pca <- recipe(y ~ ., data = peng_pca) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

wf_pca <- workflow() |> add_recipe(rec_pca) |> add_model(xgb_pca_spec)
wf_no_pca <- workflow() |> add_recipe(rec_no_pca) |> add_model(xgb_pca_spec)
```

```{r}
#| label: day04-pca-fit-compare
#| echo: true
#| message: false
set.seed(8)
rs_pca <- fit_resamples(wf_pca, folds_pca, metrics = metrics_acc)
set.seed(8)
rs_no_pca <- fit_resamples(wf_no_pca, folds_pca, metrics = metrics_acc)
bind_rows(
  collect_metrics(rs_pca) |> mutate(model = "XGB + PCA (2 comp.)"),
  collect_metrics(rs_no_pca) |> mutate(model = "XGB, no PCA")
) |>
  filter(.metric == "accuracy") |>
  select(model, mean, std_err)
```

```{r}
#| label: day04-pca-auc-bar
#| echo: false
#| fig-width: 7
#| fig-height: 3.5
bind_rows(
  collect_metrics(rs_pca) |> mutate(model = "With PCA"),
  collect_metrics(rs_no_pca) |> mutate(model = "No PCA")
) |>
  filter(.metric == "accuracy") |>
  ggplot(aes(model, mean, fill = model)) +
  geom_col(show.legend = FALSE) +
  geom_errorbar(aes(ymin = mean - std_err, ymax = mean + std_err), width = 0.12) +
  theme_minimal() +
  labs(title = "Accuracy: PCA vs all scaled numerics", x = NULL, y = "Mean accuracy")
```

**Tradeoff:** components are **not** “bill mm” anymore — VIP on PCA columns is harder to explain biologically.

**Leakage check:** always fit PCA **inside** `workflow()` + resampling, never on the full table before splitting.

# Part D — Missing data (accuracy only) {#part-d-impute}

## Extension: imputation inside the recipe (`peng_na`)

Add **`step_impute_*()`** before `step_zv()` / dummies / normalize — never `drop_na()` on the full table before splitting.

## Artificial missing data: bill length

We set **~15%** of `bill_length_mm` (and a few `sex`) to **NA** to practice imputation — not a real sensor failure pattern.

```{r}
#| label: day04-na-scatter
#| echo: true
#| fig-width: 8
#| fig-height: 4.5
plot_na_highlight(
  peng_na, "bill_length_mm", "bill_depth_mm",
  title = "Missing bill length (red ×)"
)
```

## Wrong fix: drop rows with `drop_na()`

```{r}
#| label: day04-na-drop-demo
#| echo: true
tibble(
  rows_before = nrow(peng_na),
  rows_after_drop_na = nrow(tidyr::drop_na(peng_na))
)
```

Dropping loses penguins and can **bias** who remains — never drop on the **full** table before splitting.

## Right fix: impute inside the recipe

```{r}
#| label: day04-impute-recipe
#| echo: true
rec_impute <- recipe(y ~ ., data = peng_na) |>
  step_impute_median(bill_length_mm) |>
  step_impute_mode(sex) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())
```

## Before vs after imputation (plots)

```{r}
#| label: day04-impute-before-after
#| echo: true
#| fig-width: 10
#| fig-height: 4.5
prep_impute <- prep(rec_impute, peng_na)
peng_baked <- bake(prep_impute, peng_na)

p_before <- plot_na_highlight(
  peng_na, "bill_length_mm", "bill_depth_mm",
  title = "Before imputation"
)
p_after <- ggplot(peng_baked, aes(bill_length_mm, bill_depth_mm, color = y)) +
  geom_point(size = 2.4) +
  scale_color_manual(values = c(Adelie = "black", Gentoo = "gray40")) +
  theme_minimal() +
  labs(title = "After median/mode imputation", x = "bill_length_mm", y = "bill_depth_mm")

if (requireNamespace("patchwork", quietly = TRUE)) {
  print(p_before + p_after + patchwork::plot_layout(ncol = 2))
} else {
  print(p_before)
  print(p_after)
}
```

```{r}
#| label: day04-impute-na-count
#| echo: true
sum(is.na(bake(prep_impute, peng_na)$bill_length_mm))
```

## Imputation inside cross-validation

```{r}
#| label: day04-impute-cv
#| echo: true
#| message: false
xgb_na_spec <- boost_tree(trees = 80, tree_depth = 3, learn_rate = 0.08) |>
  set_engine("xgboost") |>
  set_mode("classification")
wf_na <- workflow() |>
  add_recipe(rec_impute) |>
  add_model(xgb_na_spec)
folds_na <- vfold_cv(peng_na, v = 5, strata = y)
set.seed(14)
rs_na <- fit_resamples(wf_na, folds_na, metrics = metrics_acc)
collect_metrics(rs_na) |>
  filter(.metric == "accuracy") |>
  select(mean, std_err)
```

The workflow accepts **NA** in raw rows; each fold **learns** imputation on analysis data only.

## Metrics and deployment (reminders)

- Pick **metrics before** peeking at holdout data; report **subgroup** tables when ethically appropriate.
- **Data card:** document NA handling and class prevalence **before** modeling.
- **Model card:** intended use, limits, monitoring — finish the draft you started Monday.

Continue: [Module 06 — Metrics & SHAP](../modules/module-06-evaluation-and-interpretability.qmd) · [Day 2 (Tuesday) pipeline](../slides/day-02-tuesday.qmd).

# Part E — Imbalance, metrics, and confusion matrices {#part-e-metrics}

## Extension: class imbalance (`peng_imb`)

**Recipe extension:** add **`step_upsample(y)`** (or **`step_downsample()`** on the majority) near the top of the chain — still inside CV folds.

## Artificial imbalance: rare Gentoo

We **downsampled** Gentoo (~10 vs ~150 Adelie). Real cohorts can look like this when the event is rare.

```{r}
#| label: day04-imb-balance-plot
#| echo: false
#| fig-width: 7
#| fig-height: 4
print(plot_class_balance(peng_imb, "Manufactured class imbalance"))
```

A model that always predicts **Adelie** can still show **high accuracy** — but **miss every Gentoo** (low **sensitivity**).

## Bill space is still separable

```{r}
#| label: day04-imb-scatter
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
ggplot(peng_imb, aes(bill_length_mm, bill_depth_mm, color = y)) +
  geom_point(size = 2.6) +
  scale_color_manual(values = c(Adelie = "turquoise", Gentoo = "magenta")) +
  theme_minimal() +
  labs(
    title = "Rare Gentoo — bills still separate species",
    x = "Bill length (mm)",
    y = "Bill depth (mm)"
  )
```

**Upsampling duplicates minority rows in training** — useful when the learner **ignores** the rare class, not when the boundary is already easy.

## Recipes on `peng_imb` (code)

```{r}
#| label: day04-imb-recipes
#| echo: true
#| message: false
xgb_imb_spec <- boost_tree(trees = 80, tree_depth = 3, learn_rate = 0.08) |>
  set_engine("xgboost") |>
  set_mode("classification")

rec_imb <- recipe(y ~ ., data = peng_imb) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

rec_imb_up <- recipe(y ~ ., data = peng_imb) |>
  step_upsample(y) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

rec_imb
rec_imb_up
```

```{r}
#| label: day04-imb-prep
#| echo: true
#| message: false
wf_imb <- workflow() |> add_recipe(rec_imb) |> add_model(xgb_imb_spec)
wf_imb_up <- workflow() |> add_recipe(rec_imb_up) |> add_model(xgb_imb_spec)
```

## Cross-validated metrics: no upsample

```{r}
#| label: day04-imb-no-upsample
#| echo: true
#| message: false
set.seed(11)
rs_imb <- fit_resamples(wf_imb, folds_imb, metrics = metrics_cls_imb)
metrics_summary_table(rs_imb, "XGBoost, no upsample")
```

On this DGP we often see **very high accuracy and ROC-AUC** already.

## Cross-validated metrics: with `step_upsample(y)`

```{r}
#| label: day04-imb-with-upsample
#| echo: true
#| message: false
set.seed(12)
rs_imb_up <- fit_resamples(wf_imb_up, folds_imb, metrics = metrics_cls_imb)
metrics_summary_table(rs_imb_up, "XGBoost + upsample")
```

**Upsampling is not a free lunch** — compare tables in the next section (confusion matrix, ROC/PR, facet plot).

## Why accuracy misleads (imbalanced data)

```{r}
#| label: day04-metrics-majority
#| echo: true
#| message: false
tibble(
  strategy = "Always predict Adelie",
  accuracy_on_peng_imb = mean(peng_imb$y == "Adelie")
)
```

- **~94% accuracy** from never calling **Gentoo** — useless for screening rare events.
- Parts B–D used **`accuracy`** on **balanced** `peng_bal`; from here we add **metric_set** members for **rare positives**.

## Metrics toolbox (`yardstick`)

| Metric | Question it answers | When it shines |
|--------|---------------------|----------------|
| **accuracy** | Overall correct? | Balanced classes, equal error costs |
| **sens** (sensitivity, recall) | Of true Gentoo, how many caught? | Screening — missing positives is costly |
| **yardstick::spec** (specificity) | Of true Adelie, how many caught? | Avoid false alarms |
| **precision** | Of predicted Gentoo, how many correct? | Confirmation after a screen |
| **f_meas** (F1) | Balance precision & recall | Single score when both matter |
| **roc_auc** | Ranking quality across thresholds | Class separability, balanced-ish prevalence |
| **pr_auc** | Precision–recall area | **Rare** positive class |

```{r}
#| label: day04-metrics-set-show
#| echo: true
#| message: false
metrics_cls_imb
```

- Use **`yardstick::spec`** in code so R does not confuse **specificity** with a **`spec`** model object.

## Confusion matrix (read the rare-class row)

From the imbalance block — **held-out 20%**, no upsampling:

```{r}
#| label: day04-metrics-conf-mat
#| echo: true
#| message: false
set.seed(13)
split_imb <- initial_split(peng_imb, prop = 0.8, strata = y)
fit_imb <- fit(wf_imb, training(split_imb))
pred_imb <- augment(fit_imb, testing(split_imb))
pred_imb |>
  conf_mat(truth = y, estimate = .pred_class)
```

- **Rows** = true species; **columns** = predictions.
- **Gentoo row, Adelie column** = false negatives (missed Gentoo) — costly in screening.

## ROC vs PR curve (rare positive)

```{r}
#| label: day04-metrics-roc-pr
#| echo: true
#| fig-width: 9
#| fig-height: 4
#| message: false
set.seed(15)
split_pr <- initial_split(peng_imb, prop = 0.8, strata = y)
fit_pr <- fit(wf_imb, training(split_pr))
pred_pr <- augment(fit_pr, testing(split_pr))

p_roc <- pred_pr |>
  roc_curve(truth = y, .pred_Gentoo) |>
  autoplot() +
  labs(title = "ROC curve (held-out 20%)") +
  theme_minimal()

p_pr <- pred_pr |>
  pr_curve(truth = y, .pred_Gentoo) |>
  autoplot() +
  labs(title = "PR curve (held-out 20%)") +
  theme_minimal()

if (requireNamespace("patchwork", quietly = TRUE)) {
  print(p_roc + p_pr)
} else {
  print(p_roc)
  print(p_pr)
}
```

## Side-by-side CV metrics (imbalanced data)

```{r}
#| label: day04-metrics-compare-kable
#| echo: true
#| message: false
compare_metrics_tbl(
  "No upsample" = rs_imb,
  "Upsample" = rs_imb_up
)
```

```{r}
#| label: day04-metrics-compare-bar
#| echo: false
#| fig-width: 9
#| fig-height: 4.5
bind_rows(
  collect_metrics(rs_imb) |> mutate(model = "No upsample"),
  collect_metrics(rs_imb_up) |> mutate(model = "Upsample")
) |>
  filter(.metric %in% c("accuracy", "sens", "roc_auc", "pr_auc")) |>
  ggplot(aes(model, mean, fill = model)) +
  geom_col(show.legend = FALSE) +
  geom_errorbar(aes(ymin = mean - std_err, ymax = mean + std_err), width = 0.15) +
  facet_wrap(~.metric, scales = "free_y") +
  theme_minimal() +
  labs(
    title = "Imbalanced data: metrics need not move together",
    x = NULL,
    y = "Mean metric (5-fold CV)"
  )
```

- Pick the metric that matches **deployment cost** before tuning (`select_best(metric = "...")` in the lab).

# Part F — Variable importance {#part-f-vip}

## Data the forest sees (bill plane)

```{r}
#| label: day04-vip-scatter
#| echo: true
#| fig-width: 7
#| fig-height: 4.5
ggplot(peng_bal, aes(bill_length_mm, bill_depth_mm, color = y)) +
  geom_point(size = 2.6) +
  scale_color_manual(values = c(Adelie = "turquoise", Gentoo = "magenta")) +
  theme_minimal() +
  labs(
    title = "Adelie vs Gentoo in bill space",
    x = "Bill length (mm)",
    y = "Bill depth (mm)"
  )
```

## Variable importance (random forest)

- **`vip`** on a fitted **`ranger`** model: how often splits use each **baked** predictor (impurity).
- **Correlated** bill measures can **share** importance — not proof that changing one bill mm **causes** species shift.

```{r}
#| label: day04-vip-prep
#| echo: false
#| results: hide
rf_fit <- fit(wf_rf, peng_cls)
```

```{r}
#| label: day04-vip-plot
#| echo: true
#| fig-width: 9
#| fig-height: 5
rf_fit |>
  extract_fit_parsnip() |>
  vip(geom = "point", num_features = 12) +
  theme_minimal() +
  labs(
    title = "Variable importance (random forest)",
    subtitle = "Adelie vs Gentoo — descriptive ranking, not causal"
  )
```

**Model-agnostic alternative:** **permutation importance** (shuffle one column, watch metric drop) — see [Module 06](../modules/module-06-evaluation-and-interpretability.qmd).

**Local explanations:** SHAP (TreeSHAP for forests) — same module; treat as **communication**, not intervention.

## SHAP (pointer)

- **SHAP:** local explanations for **one prediction** — communication, not causation
- **TreeSHAP** for forests/boosting; details in [Module 06](../modules/module-06-evaluation-and-interpretability.qmd)

# Part G — Wrap-up

## Nested resampling (when tuning gets serious)

- **Outer** CV for performance; **inner** CV for hyperparameters inside each analysis fold
- Avoid tuning and scoring on the same held-out rows

## End-of-day checklist

- Name three **`parsnip` model types** and a valid **engine** for each
- Why must **`step_pca()`** and **`step_impute_*()`** live **inside** the recipe?
- When might **`step_upsample()`** not improve CV scores?
- Which metric would you report for a **rare-event screening** study?
- One reason **VIP** on correlated bills is **not** a causal claim
