## PCA inside a recipe

- **`step_pca(all_numeric_predictors(), num_comp = k)`** learns orthogonal directions from **training rows only** (inside each CV fold when resampled).
- Use when you have **many correlated** measurements and want a compact input layer for a flexible model.
- **Tradeoff:** components are **not** “bill length” anymore — harder to tell a biological story from loadings alone.

```{r}
#| label: day04-pca-recipe-code
#| echo: true
rec_pca <- recipe(y ~ ., data = peng_cls) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_pca(all_numeric_predictors(), num_comp = 3)

rec_pca
```

```{r}
#| label: day04-pca-fit
#| echo: true
#| message: false
xgb_pca_spec <- boost_tree(trees = 100, tree_depth = 3, learn_rate = 0.05) |>
  set_engine("xgboost") |>
  set_mode("classification")

wf_pca <- workflow() |>
  add_recipe(rec_pca) |>
  add_model(xgb_pca_spec)

set.seed(8)
rs_pca <- fit_resamples(wf_pca, folds, metrics = metrics_cls)
collect_metrics(rs_pca) |>
  filter(.metric == "roc_auc") |>
  select(mean, std_err)
```

Compare to **no PCA** XGBoost from the ensemble slide — similar AUC is common here because we only have a few bill + year numerics after dummy coding.

**Leakage check:** never `step_pca()` on the **full** table before splitting; always inside **`workflow()`** + **`fit_resamples()`** or train-only `fit()`.
