## Extension: imputation inside the recipe (`peng_na`)

Add **`step_impute_*()`** before `step_zv()` / dummies / normalize — never `drop_na()` on the full table before splitting.

## Artificial missing data: bill length

We set **~15%** of `bill_length_mm` (and a few `sex`) to **NA** (**missing** values — “not available”) to practice imputation — not a real sensor failure pattern.

```{r}
#| label: day04-na-scatter
#| echo: false
#| fig-width: 9
#| fig-height: 6
#| out-width: 100%
plot_na_highlight(
  peng_na, "bill_length_mm", "bill_depth_mm",
  title = "Missing bill length (red ×)"
)
```

## Wrong fix: drop rows with `drop_na()`

```{r}
#| label: day04-na-drop-demo
#| echo: true
tibble(
  rows_before = nrow(peng_na),
  rows_after_drop_na = nrow(tidyr::drop_na(peng_na))
)
```

Dropping loses penguins and can **bias** who remains — never drop on the **full** table before splitting.

## Right fix: impute inside the recipe

- **`step_impute_median()`** for numeric **`bill_length_mm`**; **`step_impute_mode()`** for categorical **`sex`**.
- Then the usual **`step_zv` → `step_dummy` → `step_normalize`** chain from Part A.

```{r}
#| label: day04-impute-recipe
#| echo: true
rec_impute <- recipe(y ~ ., data = peng_na) |>
  step_impute_median(bill_length_mm) |>
  step_impute_mode(sex) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())
```

## Before imputation

Rows with **missing bill length** sit in the **left margin** (pink band) — we do **not** know their x-position yet.

```{r}
#| label: day04-impute-prep-plots
#| echo: false
#| results: hide
prep_impute <- prep(rec_impute, peng_na)
rec_impute_fill <- recipe(y ~ ., data = peng_na) |>
  step_impute_median(bill_length_mm) |>
  step_impute_mode(sex)
peng_filled <- bake(prep(rec_impute_fill, peng_na), peng_na)
peng_na_plot <- peng_na |>
  mutate(bill_length_true = bill_length_true)
```

```{r}
#| label: day04-impute-before
#| echo: false
#| fig-width: 9
#| fig-height: 6
#| out-width: 100%
print(
  plot_na_highlight(
    peng_na, "bill_length_mm", "bill_depth_mm",
    title = "Before imputation — missing bill length"
  )
)
```

## After imputation

**Open circles** = true bill length (simulation only). **Gold** = median imputed value for the **same** penguin. **Orange arrows** connect true → imputed. Gray points are other birds (observed bill length).

```{r}
#| label: day04-impute-after
#| echo: false
#| fig-width: 9
#| fig-height: 6
#| out-width: 100%
print(
  plot_imputed_highlight(
    peng_na_plot,
    peng_filled,
    "bill_length_mm",
    "bill_depth_mm",
    true_xvar = "bill_length_true",
    title = "After median imputation — bill length filled in"
  )
)
```

```{r}
#| label: day04-impute-na-count
#| echo: true
sum(is.na(bake(prep_impute, peng_na)$bill_length_mm))
```

## Imputation inside cross-validation

```{r}
#| label: day04-impute-cv
#| echo: true
#| message: false
xgb_na_spec <- boost_tree(trees = 80, tree_depth = 3, learn_rate = 0.08) |>
  set_engine("xgboost") |>
  set_mode("classification")
wf_na <- workflow() |>
  add_recipe(rec_impute) |>
  add_model(xgb_na_spec)
folds_na <- vfold_cv(peng_na, v = 5, strata = y)
set.seed(14)
rs_na <- fit_resamples(wf_na, folds_na, metrics = metrics_acc)
collect_metrics(rs_na) |>
  filter(.metric == "accuracy") |>
  select(mean, std_err)
```

The workflow accepts **NA** in raw rows; each fold **learns** imputation on analysis data only.

## Metrics and deployment (reminders)

- Pick **metrics before** peeking at holdout data; report **subgroup** tables when ethically appropriate.
- **Data card:** document NA handling and class prevalence **before** modeling.
- **Model card:** intended use, limits, monitoring — finish the draft you started Monday.

Continue: [Day 2 (Tuesday) pipeline](../slides/day-02-tuesday.html#/part-c-pipeline).
