## Data the forest sees (bill plane)

```{r}
#| label: day04-vip-scatter
#| echo: true
#| fig-width: 7
#| fig-height: 4.5
ggplot(peng_bal, aes(bill_length_mm, bill_depth_mm, color = y)) +
  geom_point(size = 2.6) +
  scale_color_manual(values = c(Adelie = "turquoise", Gentoo = "magenta")) +
  theme_minimal() +
  labs(
    title = "Adelie vs Gentoo in bill space",
    x = "Bill length (mm)",
    y = "Bill depth (mm)"
  )
```

## Variable importance (random forest)

- **`vip`** on a fitted **`ranger`** model: how often splits use each **baked** predictor (impurity).
- **Correlated** bill measures can **share** importance — not proof that changing one bill mm **causes** species shift.

```{r}
#| label: day04-vip-prep
#| echo: false
#| results: hide
rf_fit <- fit(wf_rf, peng_cls)
```

```{r}
#| label: day04-vip-plot
#| echo: true
#| fig-width: 9
#| fig-height: 5
rf_fit |>
  extract_fit_parsnip() |>
  vip(geom = "point", num_features = 12) +
  theme_minimal() +
  labs(
    title = "Variable importance (random forest)",
    subtitle = "Adelie vs Gentoo — descriptive ranking, not causal"
  )
```

**Model-agnostic alternative:** **permutation importance** (shuffle one column, watch metric drop) — see the **VIP plot** on the Day 4 deck.

**Local explanations:** [SHAP on sex (slides)](../slides/day-04-thursday.html#/shap---what-it-means-game-theory-intuition) · [SHAP notebook](../notebooks/penguins-shap.html) — treat as **communication**, not intervention.
