## Palmer Penguins: one table, one model family (trees)

- **Outcome `y`:** Adelie vs Gentoo (complete cases).
- **Predictors:** bill length & depth, **island**, **sex**, **year** — no flipper or body mass (so **`tree_depth`** and **`min_n`** show a real tuning tradeoff).
- **Today’s engine:** **`decision_tree()` + `rpart` only** — forests and boosting come on **Thursday**.

```{r}
#| label: day02-walk-balance
#| echo: false
#| fig-width: 7
#| fig-height: 4
plot_class_balance(peng_bal, "Balanced classes")
```

```{r}
#| label: day02-walk-scatter
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
ggplot(peng_bal, aes(bill_length_mm, bill_depth_mm, color = y)) +
  geom_point(size = 2.6) +
  scale_color_manual(values = c(Adelie = "turquoise", Gentoo = "magenta")) +
  theme_minimal() +
  labs(
    title = "Bill space — Adelie vs Gentoo",
    x = "Bill length (mm)",
    y = "Bill depth (mm)"
  )
```

## Pipeline map

**Data → recipe → spec → workflow → resampling/tuning → fit → figures**

| Package | Job |
|---------|-----|
| **`rsample`** | Training vs assessment rows (`initial_split`, `vfold_cv`) |
| **`recipes`** | Preprocessing checklist (dummy, scale, …) |
| **`parsnip`** | Model **blueprint** (`decision_tree`, `set_engine("rpart")`) |
| **`workflows`** | Glue recipe + model |
| **`tune`** | Search `tree_depth` and `min_n` |
| **`yardstick`** | Scores — **accuracy only** today; more metrics Thursday |

# Step 1 — Preprocessing (`recipe`)

A **recipe** defines transforms **before** the model sees data. Steps run top to bottom.

## Start the recipe

```{r}
#| label: day02-walk-rec-start
#| echo: true
#| message: false
rec <- recipe(y ~ ., data = peng_bal)
rec
```

- **`recipe(y ~ ., data = peng_bal)`** — `y` is the outcome; **`.`** = all other columns are predictors.
- Blueprint only — nothing transformed yet.

## `step_zv()` — drop zero-variance columns

```{r}
#| label: day02-walk-rec-zv
#| echo: true
#| message: false
rec <- rec |>
  step_zv(all_predictors())
rec
```

- Removes columns with one unique value (no predictive power; can break algorithms in small folds).

## `step_dummy()` — encode categories

```{r}
#| label: day02-walk-rec-dummy
#| echo: true
#| message: false
rec <- rec |>
  step_dummy(all_nominal_predictors())
rec
```

- **`island`** and **`sex`** → numeric 0/1 columns for **`rpart`**.

## `step_normalize()` — scale numeric predictors

```{r}
#| label: day02-walk-rec-norm
#| echo: true
#| message: false
rec <- rec |>
  step_normalize(all_numeric_predictors())
rec
```

- Z-score each numeric column using **training-fold rows only** inside CV.
- Good habit before you swap in other model types later in the week.

```{r}
#| label: day02-walk-rec-base-assign
#| echo: true
#| message: false
rec_base <- rec
rec_base
```

::: {.callout-note}
## Afternoon lab (microbiome)

Same pattern on OTU counts — add **`step_mutate(..., log1p(.x))`** on skewed abundances before `step_zv`.
:::

**`prep()`** learns rules; **`bake()`** applies them. A **`workflow()`** does both inside each CV fold.

# Step 2 — Model blueprint (`spec`)

**`parsnip`** stores the family and hyperparameters. **Nothing is fitted yet.**

## Classification tree with `rpart`

```{r}
#| label: day02-walk-tree-spec-fixed
#| echo: true
#| message: false
tree_spec_demo <- decision_tree(tree_depth = 4, min_n = 10) |>
  set_engine("rpart") |>
  set_mode("classification")
tree_spec_demo
```

- **`decision_tree()`** — fixed **parsnip** function name.
- **`tree_depth`**, **`min_n`** — depth limit and minimum leaf size.
- **`set_engine("rpart")`** — same engine as the gene trees this morning.
- **`set_mode("classification")`** — factor outcome **`y`**.

## Tuning knobs (`tune()`)

```{r}
#| label: day02-walk-tree-spec-tune
#| echo: true
#| message: false
tree_spec <- decision_tree(
  tree_depth = tune(),
  min_n = tune()
) |>
  set_engine("rpart") |>
  set_mode("classification")
tree_spec
```

- **`tune()`** marks arguments **`tune_grid()`** will search — we pick winners by **cross-validated accuracy**.

## Parsnip: type → engine → mode

| Model type | Valid engines (examples) | Modes |
|------------|--------------------------|-------|
| `decision_tree()` | `"rpart"`, `"C5.0"` | classification, regression |
| `rand_forest()` | `"ranger"`, `"randomForest"` | classification, regression |
| `boost_tree()` | `"xgboost"`, `"lightgbm"` | classification, regression |

Thursday swaps **`spec`** — same **`recipe`** and **`workflow()`** pattern.

```{r}
#| label: day02-walk-bad-engine
#| echo: true
#| eval: false
#| message: false
decision_tree() |>
  set_engine("lm") |>
  set_mode("classification")
# Error: 'lm' is not a valid engine for decision_tree()
```

# Step 3 — Workflow

```{r}
#| label: day02-walk-wf
#| echo: true
#| message: false
wf <- workflow() |>
  add_recipe(rec_base) |>
  add_model(tree_spec)
wf
```

- Preprocessing + model in **one** object — no scaling the full table before splitting.

## One-shot fit (intuition)

```{r}
#| label: day02-walk-fit-demo
#| echo: true
#| message: false
wf_demo <- workflow() |>
  add_recipe(rec_base) |>
  add_model(tree_spec_demo)
fit_demo <- fit(wf_demo, data = peng_bal)
```

```{r}
#| label: day02-walk-boundary
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
print(plot_prob_boundary(fit_demo, peng_bal, title = "Fitted tree — P(Gentoo)"))
```

# Step 4 — Metrics (accuracy today)

```{r}
#| label: day02-walk-metrics
#| echo: true
#| message: false
metrics_day2
```

- **`accuracy`** — fraction of correct class labels.
- Thursday adds sensitivity, ROC-AUC, PR-AUC, and more when classes are imbalanced.

# Step 5 — Cross-validation and tuning

```{r}
#| label: day02-walk-tune
#| echo: true
#| message: false
set.seed(2)
tune_res <- tune_grid(
  wf,
  resamples = folds,
  grid = grid_regular(
    tree_depth(range = c(1, 6)),
    min_n(range = c(2, 50)),
    levels = 4
  ),
  metrics = metrics_day2
)
```

- **`vfold_cv(..., strata = y)`** — balanced species per fold.
- Grid spans underfit (depth 1, large `min_n`) → more flexible trees.

## Read the tuning plot

```{r}
#| label: day02-walk-autoplot
#| echo: true
#| fig-width: 10
#| fig-height: 5
tune_res |>
  autoplot() +
  theme_minimal() +
  labs(
    title = "Tuning: accuracy vs tree_depth and min_n",
    subtitle = "Bill + island + sex + year — no flipper/mass"
  )
```

- Look for **lower accuracy** at **`tree_depth = 1`** (stump) and with **large `min_n`** (heavy pruning).

## Finalize and fit

```{r}
#| label: day02-walk-finalize
#| echo: true
#| message: false
best <- select_best(tune_res, metric = "accuracy")
final_wf <- finalize_workflow(wf, best)
final_fit <- fit(final_wf, data = peng_cls)
best
```

```{r}
#| label: day02-walk-fit-resamples
#| echo: true
#| message: false
set.seed(6)
wf_tree <- final_wf
rs_tree <- fit_resamples(wf_tree, folds, metrics = metrics_day2)
metrics_summary_table(rs_tree, "Tuned tree (5-fold CV)")
```

Teaching speed: we **`fit` on all rows** after tuning; in projects, hold out a true test set with **`last_fit`**.

## Fitted tree structure

```{r}
#| label: day02-walk-rpart-plot
#| echo: true
#| fig-width: 10
#| fig-height: 6
final_fit |>
  extract_fit_engine() |>
  rpart.plot(
    roundint = FALSE,
    type = 4,
    extra = 104,
    box.palette = "RdBu",
    main = "Final decision tree (rpart)"
  )
```

## Variable importance (this tree)

```{r}
#| label: day02-walk-vip
#| echo: true
#| fig-width: 9
#| fig-height: 5
final_fit |>
  extract_fit_parsnip() |>
  vip(geom = "point", aesthetics = list(color = "midnightblue", size = 3)) +
  theme_minimal() +
  labs(
    title = "Split-based importance (single tree)",
    subtitle = "Correlated bills ≠ causation — full VIP chapter on Thursday"
  )
```
