Short **tidyverse** refresher before the `tidymodels` pipeline (same pipe grammar as recipes and workflows).

```{r}
#| label: day02-tidy-intro-setup
#| include: false
suppressPackageStartupMessages(library(tidyverse))
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Tidy data and the pipe `|>`

- Each **column** is one variable; each **row** is one penguin (or one sample)
- **`x |> f()`** means: take the result of `x` and pass it as the **first argument** to `f()`
- Read chains **top to bottom** — same habit you will use for `recipe() |> step_*()`

## Example: `filter` and `select`

```{r}
#| label: day02-tidy-filter-select
#| echo: true
pg <- palmerpenguins::penguins |>
  filter(!is.na(sex), !is.na(bill_length_mm), !is.na(bill_depth_mm)) |>
  select(species, island, sex, bill_length_mm, bill_depth_mm, body_mass_g)

head(pg, 4)
```

## Example: `mutate` (new columns)

```{r}
#| label: day02-tidy-mutate
#| echo: true
pg |>
  mutate(bill_ratio = bill_length_mm / bill_depth_mm) |>
  select(species, bill_length_mm, bill_depth_mm, bill_ratio) |>
  head(4)
```

## Same grammar in `tidymodels`

- **Table pipelines:** `penguins |> filter(...) |> mutate(...)`
- **Preprocessing:** `recipe(y ~ ., data = train) |> step_dummy() |> step_normalize()`
- **Modeling objects:** `workflow() |> add_recipe(rec) |> add_model(spec)` — here you chain **objects**, not tables, but the **read left-to-right** habit is the same
