## Why accuracy misleads (imbalanced data)

```{r}
#| label: day04-metrics-majority
#| echo: true
#| message: false
tibble(
  strategy = "Always predict Adelie",
  accuracy_on_peng_imb = mean(peng_imb$y == "Adelie")
)
```

- **~94% accuracy** from never calling **Gentoo** — useless for screening rare events.
- Parts B–D used **`accuracy`** on **balanced** `peng_bal`; from here we add **metric_set** members for **rare positives**.

## Metrics toolbox (`yardstick`)

| Metric | Question it answers | When it shines |
|--------|---------------------|----------------|
| **accuracy** | Overall correct? | Balanced classes, equal error costs |
| **sens** (sensitivity, recall) | Of true Gentoo, how many caught? | Screening — missing positives is costly |
| **yardstick::spec** (specificity) | Of true Adelie, how many caught? | Avoid false alarms |
| **precision** | Of predicted Gentoo, how many correct? | Confirmation after a screen |
| **f_meas** (F1) | Balance precision & recall | Single score when both matter |
| **roc_auc** | Ranking quality across thresholds | Class separability, balanced-ish prevalence |
| **pr_auc** | Precision–recall area | **Rare** positive class |

```{r}
#| label: day04-metrics-set-show
#| echo: true
#| message: false
metrics_cls_imb
```

- Use **`yardstick::spec`** in code so R does not confuse **specificity** with a **`spec`** model object.

## Confusion matrix (read the rare-class row)

From the imbalance block — **held-out 20%**, no upsampling:

```{r}
#| label: day04-metrics-conf-mat
#| echo: true
#| message: false
set.seed(13)
split_imb <- initial_split(peng_imb, prop = 0.8, strata = y)
fit_imb <- fit(wf_imb, training(split_imb))
pred_imb <- augment(fit_imb, testing(split_imb))
pred_imb |>
  conf_mat(truth = y, estimate = .pred_class)
```

- **Rows** = true species; **columns** = predictions.
- **Gentoo row, Adelie column** = false negatives (missed Gentoo) — costly in screening.

## ROC vs PR curve (rare positive)

```{r}
#| label: day04-metrics-roc-pr
#| echo: true
#| fig-width: 9
#| fig-height: 4
#| message: false
set.seed(15)
split_pr <- initial_split(peng_imb, prop = 0.8, strata = y)
fit_pr <- fit(wf_imb, training(split_pr))
pred_pr <- augment(fit_pr, testing(split_pr))

p_roc <- pred_pr |>
  roc_curve(truth = y, .pred_Gentoo) |>
  autoplot() +
  labs(title = "ROC curve (held-out 20%)") +
  theme_minimal()

p_pr <- pred_pr |>
  pr_curve(truth = y, .pred_Gentoo) |>
  autoplot() +
  labs(title = "PR curve (held-out 20%)") +
  theme_minimal()

if (requireNamespace("patchwork", quietly = TRUE)) {
  print(p_roc + p_pr)
} else {
  print(p_roc)
  print(p_pr)
}
```

## Side-by-side CV metrics (imbalanced data)

```{r}
#| label: day04-metrics-compare-kable
#| echo: true
#| message: false
compare_metrics_tbl(
  "No upsample" = rs_imb,
  "Upsample" = rs_imb_up
)
```

```{r}
#| label: day04-metrics-compare-bar
#| echo: false
#| fig-width: 9
#| fig-height: 4.5
bind_rows(
  collect_metrics(rs_imb) |> mutate(model = "No upsample"),
  collect_metrics(rs_imb_up) |> mutate(model = "Upsample")
) |>
  filter(.metric %in% c("accuracy", "sens", "roc_auc", "pr_auc")) |>
  ggplot(aes(model, mean, fill = model)) +
  geom_col(show.legend = FALSE) +
  geom_errorbar(aes(ymin = mean - std_err, ymax = mean + std_err), width = 0.15) +
  facet_wrap(~.metric, scales = "free_y") +
  theme_minimal() +
  labs(
    title = "Imbalanced data: metrics need not move together",
    x = NULL,
    y = "Mean metric (5-fold CV)"
  )
```

- Pick the metric that matches **deployment cost** before tuning (`select_best(metric = "...")` in the lab).
