Visuals below follow the same simulations as the **student notebooks** in this repo:

- [Linear regression + LASSO](../notebooks/linear-regression-lasso.Rmd)
- [Logistic regression + tree](../notebooks/logistic-regression-gene-disease.Rmd)

```{r}
#| label: day01-examples-libs
#| include: false
suppressPackageStartupMessages({
  library(glmnet)
  library(caret)
  library(dplyr)
  library(ggplot2)
  library(GGally)
  library(rlang)
  library(MASS)
})
theme_set(theme_classic())
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  out.width = "100%",
  dpi = 120
)
```

## Example: synthetic genes → continuous trait

Same data-generating story as the linear / LASSO notebook: GeneA and GeneB drive the trait; correlated noise genes; GeneJ correlated with GeneA but not causal.

```{r}
#| label: day01-linear-sim-data
#| echo: false
#| results: hide
set.seed(123)
n <- 150
p <- 10
gene_names <- c("GeneA", "GeneB", "GeneC", "GeneD", "GeneE", "GeneF", "GeneG", "GeneH", "GeneI", "GeneJ")
gene_data <- matrix(rnorm(n * p), nrow = n)
colnames(gene_data) <- gene_names
trait <- -0.5 * gene_data[, "GeneA"] + 0.8 * gene_data[, "GeneB"] + rnorm(n, sd = 0.3)
gene_data[, "GeneC"] <- gene_data[, "GeneA"] + rnorm(n, sd = 0.2)
gene_data[, "GeneD"] <- -0.3 * gene_data[, "GeneB"] + rnorm(n, sd = 0.2)
for (i in 5:p) gene_data[, gene_names[i]] <- rnorm(n)
gene_data[, "GeneF"] <- gene_data[, "GeneE"] + rnorm(n, sd = 0.2)
gene_data[, "GeneH"] <- -0.5 * gene_data[, "GeneG"] + rnorm(n, sd = 0.2)
gene_data[, "GeneJ"] <- 0.7 * gene_data[, "GeneA"] + rnorm(n, sd = 0.2)
data_trait <- as.data.frame(cbind(gene_data, trait))
```

```{r}
#| label: day01-correlation-heatmap
#| echo: false
#| fig-height: 4.5
#| fig-width: 8
correlation_matrix <- cor(data_trait)
cor_data <- as.data.frame(as.table(correlation_matrix)) |>
  dplyr::rename(variable1 = Var1, variable2 = Var2, correlation = Freq)
ggplot(cor_data, aes(variable1, variable2, fill = correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "dodgerblue", high = "firebrick4", mid = "white",
    midpoint = 0, limit = c(-1, 1), name = "r"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_fixed() +
  labs(title = "Correlation among genes and trait")
```

```{r}
#| label: day01-ggpairs-snippet
#| echo: false
#| fig-height: 5.5
#| fig-width: 8
ggpairs(
  data_trait,
  upper = list(continuous = wrap("cor", size = 2.2)),
  aes(alpha = 0.15)
) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Ridge vs Lasso: cross-validated error (`cv.glmnet`)

Train/test split and scaling match the original notebook (`createDataPartition`, `preProcess`).

```{r}
#| label: day01-train-test-glmnet
#| echo: false
#| results: hide
X <- as.matrix(dplyr::select(data_trait, -trait))
y <- data_trait$trait
trainIndex <- createDataPartition(y, p = 0.8, list = FALSE)
X_train <- X[trainIndex, , drop = FALSE]
y_train <- y[trainIndex]
X_test <- X[-trainIndex, , drop = FALSE]
y_test <- y[-trainIndex]
preproc <- preProcess(X_train, method = c("center", "scale"))
X_train_s <- predict(preproc, X_train)
X_test_s <- predict(preproc, X_test)
lam_seq <- 10^seq(-3, -0.1, length.out = 25)
cv_ridge <- cv.glmnet(X_train_s, y_train, alpha = 0, lambda = lam_seq)
cv_lasso <- cv.glmnet(X_train_s, y_train, alpha = 1, lambda = lam_seq)
```

```{r}
#| label: day01-cv-plots
#| echo: false
#| fig-height: 4
#| fig-width: 9
oldpar <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1), mgp = c(2, 0.7, 0))
plot(cv_ridge, main = "Ridge (alpha = 0)")
plot(cv_lasso, main = "Lasso (alpha = 1)")
par(oldpar)
```

```{r}
#| label: day01-coef-paths
#| echo: false
#| fig-height: 4.2
#| fig-width: 9
fit_ridge <- glmnet(X_train_s, y_train, alpha = 0, lambda = lam_seq)
fit_lasso <- glmnet(X_train_s, y_train, alpha = 1, lambda = lam_seq)
oldpar <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1), mgp = c(2, 0.7, 0))
plot(fit_ridge, xvar = "lambda", label = TRUE, main = "Ridge coefficients vs log(lambda)")
plot(fit_lasso, xvar = "lambda", label = TRUE, main = "Lasso coefficients vs log(lambda)")
par(oldpar)
```

## OLS linear regression (matches `LinearRegressionLasso.Rmd`)

Ordinary least squares on the **same scaled training matrix** as LASSO: `lm(trait ~ .)` with all genes. This is the “full complexity” linear benchmark before discrete model search (AIC) or continuous shrinkage (ridge/lasso).

```{r}
#| label: day01-lm-full-train
#| echo: false
#| results: hide
X_train_df <- as.data.frame(X_train_s)
X_train_df$trait <- y_train
X_test_df <- as.data.frame(X_test_s)
full_lm <- lm(trait ~ ., data = X_train_df)
```

```{r}
#| label: day01-lm-coef-plot
#| echo: false
#| fig-height: 4.2
#| fig-width: 8
cf <- coef(full_lm)
cf <- cf[names(cf) != "(Intercept)"]
ci <- tryCatch(
  confint(full_lm, parm = names(cf)),
  error = function(e) matrix(NA_real_, length(cf), 2, dimnames = list(names(cf), c("2.5 %", "97.5 %")))
)
coef_tbl <- data.frame(
  gene = names(cf),
  estimate = as.numeric(cf),
  lo = as.numeric(ci[, 1]),
  hi = as.numeric(ci[, 2])
)
ggplot(coef_tbl, aes(x = estimate, y = reorder(gene, estimate))) +
  geom_vline(xintercept = 0, linetype = 2, color = "gray50") +
  geom_pointrange(aes(xmin = lo, xmax = hi), fatten = 3) +
  labs(
    title = "OLS coefficients ± 95% CI (train, all predictors)",
    x = "Effect on trait (scaled gene units)", y = NULL
  ) +
  theme_minimal()
```

```{r}
#| label: day01-lm-test-metrics
#| echo: false
#| results: asis
pred_lm_test <- predict(full_lm, newdata = X_test_df)
mse_ols <- mean((y_test - pred_lm_test)^2)
rmse_ols <- sqrt(mse_ols)
r2_ols <- cor(y_test, pred_lm_test)^2
best_lam <- cv_lasso$lambda.1se
lasso_best_fit <- glmnet(X_train_s, y_train, alpha = 1, lambda = best_lam)
pred_lasso_test <- as.numeric(predict(lasso_best_fit, newx = X_test_s))
mse_las <- mean((y_test - pred_lasso_test)^2)
rmse_las <- sqrt(mse_las)
r2_las <- cor(y_test, pred_lasso_test)^2
knitr::kable(
  data.frame(
    model = c("OLS full (train fit)", "Lasso (lambda.1se)"),
    RMSE_test = c(rmse_ols, rmse_las),
    R2_test = c(r2_ols, r2_las)
  ),
  digits = 3,
  caption = "Held-out test performance (same split, scaled features)"
)
```

## AIC: backward and forward selection (same notebook)

`stepAIC` greedily adds or drops terms to improve the Akaike information criterion \( \mathrm{AIC} = 2k - 2\log L \), trading fit against number of parameters \(k\).

```{r}
#| label: day01-step-aic
#| echo: false
#| results: hide
backward_model <- stepAIC(full_lm, direction = "backward", trace = FALSE)
start_model <- lm(trait ~ 1, data = X_train_df)
forward_model <- stepAIC(
  start_model,
  direction = "forward",
  scope = list(lower = start_model, upper = full_lm),
  trace = FALSE
)
```

```{r}
#| label: day01-aic-model-sizes
#| echo: false
#| fig-height: 3.2
#| fig-width: 6.5
aic_tbl <- data.frame(
  approach = c("OLS full", "AIC backward", "AIC forward"),
  n_terms = c(
    length(coef(full_lm)),
    length(coef(backward_model)),
    length(coef(forward_model))
  ),
  AIC = c(AIC(full_lm), AIC(backward_model), AIC(forward_model))
)
ggplot(aic_tbl, aes(x = reorder(approach, AIC), y = AIC, fill = approach)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "AIC across nested-search models (train)", x = NULL, y = "AIC (lower is better)") +
  theme_minimal()
```

```{r}
#| label: day01-aic-terms-text
#| echo: false
#| results: asis
fmt_f <- function(f) paste(deparse(f), collapse = " ")
cat("**Backward AIC formula:** `", fmt_f(formula(backward_model)), "`\n\n", sep = "")
cat("**Forward AIC formula:** `", fmt_f(formula(forward_model)), "`\n\n", sep = "")
```

```{r}
#| label: day01-aic-vs-lasso-coefs
#| echo: false
#| fig-height: 4
#| fig-width: 9
coef_back <- coef(backward_model)
nm <- names(coef_back)
nm <- nm[nm != "(Intercept)"]
lc_mat <- as.matrix(coef(lasso_best_fit, s = best_lam))
lasso_vals <- ifelse(nm %in% rownames(lc_mat), lc_mat[nm, 1], NA_real_)
coef_compare_long <- rbind(
  data.frame(gene = nm, model = "backward", estimate = as.numeric(coef_back[nm])),
  data.frame(gene = nm, model = "lasso", estimate = lasso_vals)
)
ggplot(coef_compare_long, aes(x = estimate, y = reorder(gene, estimate), color = model)) +
  geom_vline(xintercept = 0, linetype = 2, color = "gray60") +
  geom_point(position = position_dodge(width = 0.4), size = 2) +
  labs(
    title = "Backward-AIC vs Lasso(1se) coefficients (train)",
    x = "Coefficient", y = NULL, color = NULL
  ) +
  theme_minimal()
```

## Example: gene expression → disease (logistic notebook)

Same simulation as `LogReg.Rmd`: causal roles for Gene1, Gene3, Gene4 and an interaction term in the generative linear predictor.

```{r}
#| label: day01-logistic-sim
#| echo: false
#| results: hide
set.seed(123)
n <- 200
p <- 10
gene_names <- paste0("Gene", 1:p)
genes_data <- data.frame(matrix(rnorm(n * p), nrow = n))
colnames(genes_data) <- gene_names
disease_presence <- ifelse(
  0.5 * genes_data$Gene1 + 0.8 * genes_data$Gene3 - 0.3 * genes_data$Gene4 -
    genes_data$Gene1 * genes_data$Gene3 * 2 + rnorm(n, sd = 0.5) > 0,
  1, 0
)
disease_presence <- factor(disease_presence, levels = c(0, 1), labels = c("Absent", "Present"))
bio_data <- data.frame(genes_data, disease_presence)
logistic_model <- glm(disease_presence ~ ., data = bio_data, family = binomial())
```

```{r}
#| label: day01-ggpairs-logistic
#| echo: false
#| fig-height: 5.2
#| fig-width: 8
ggpairs(bio_data, columns = 1:5, aes(color = disease_presence)) +
  theme_minimal()
```

```{r}
#| label: day01-logistic-boundary
#| echo: false
#| fig-height: 4.5
#| fig-width: 7
plot_decision_boundary_logistic <- function(model, data, feature1, feature2) {
  r1 <- range(data[[feature1]])
  r2 <- range(data[[feature2]])
  grid <- expand.grid(
    seq(r1[1] - 1, r1[2] + 1, length.out = 160),
    seq(r2[1] - 1, r2[2] + 1, length.out = 160)
  )
  names(grid) <- c(feature1, feature2)
  for (feat in setdiff(names(data), c(feature1, feature2, "disease_presence"))) {
    grid[[feat]] <- mean(data[[feat]])
  }
  grid$response_prob <- predict(model, newdata = grid, type = "response")
  grid$response_cls <- factor(
    ifelse(grid$response_prob > 0.5, "Present", "Absent"),
    levels = c("Absent", "Present")
  )
  ggplot(data, aes(!!sym(feature1), !!sym(feature2))) +
    geom_tile(
      data = grid,
      aes(!!sym(feature1), !!sym(feature2), fill = response_cls),
      alpha = 0.28,
      inherit.aes = FALSE
    ) +
    geom_point(aes(fill = disease_presence), shape = 21, color = "black", size = 2.5) +
    scale_fill_manual(values = c(Absent = "turquoise", Present = "magenta")) +
    theme_minimal() +
    labs(
      title = "Logistic boundary (Gene1 vs Gene3; other genes at mean)",
      x = feature1, y = feature2, fill = "Class"
    )
}
print(plot_decision_boundary_logistic(logistic_model, bio_data, "Gene1", "Gene3"))
```
