**Gene × disease** — describe data → null vs full `glm` → AIC forward/backward (all $n$ samples).

- [Logistic regression + tree (genes)](../notebooks/logistic-regression-gene-disease.html)

```{r}
#| label: day01-logistic-libs
#| include: false
suppressPackageStartupMessages({
  library(broom)
  library(glmnet)
  library(recipes)
  library(dplyr)
  library(ggplot2)
  library(GGally)
  library(MASS)
  library(rlang)
  library(rcartocolor)
})
```

## Gene × disease: the dataset

Causal roles for **Gene1**, **Gene3**, **Gene4** and a Gene1×Gene3 interaction in the generative linear predictor.

```{r}
#| label: day01-logistic-sim
#| echo: false
#| results: hide
set.seed(123)
n <- 200
p <- 10
gene_names <- paste0("Gene", 1:p)
genes_data <- data.frame(matrix(rnorm(n * p), nrow = n))
colnames(genes_data) <- gene_names
disease_presence <- ifelse(
  0.5 * genes_data$Gene1 + 0.8 * genes_data$Gene3 - 0.3 * genes_data$Gene4 -
    genes_data$Gene1 * genes_data$Gene3 * 2 + rnorm(n, sd = 0.5) > 0,
  1, 0
)
disease_presence <- factor(disease_presence, levels = c(0, 1), labels = c("Absent", "Present"))
bio_data <- data.frame(genes_data, disease_presence)
```

```{r}
#| label: day01-ggpairs-logistic
#| echo: false
#| fig-height: 5.2
#| fig-width: 8
ggpairs(bio_data, columns = 1:5, aes(color = disease_presence)) +
  theme_minimal()
```

```{r}
#| label: day01-logistic-prep-full
#| echo: false
#| results: hide
rec_scale_log <- recipe(disease_presence ~ ., data = bio_data) |>
  step_normalize(all_numeric_predictors())
bio_scaled <- bake(prep(rec_scale_log, bio_data), bio_data)
y_log <- bio_data$disease_presence
y_bin <- ifelse(y_log == "Present", 1, 0)
X_log_s <- as.matrix(dplyr::select(bio_scaled, -disease_presence))
n_obs_log <- nrow(bio_scaled)
```

## Null model (intercept only)

Predicts the overall disease rate — ignores genes → **underfitting**.

```{r}
#| label: day01-logistic-null
#| echo: false
#| results: hide
null_glm <- glm(disease_presence ~ 1, data = bio_scaled, family = binomial())
```

## Full logistic model (all genes)

```{r}
#| label: day01-logistic-full
#| echo: false
#| results: hide
full_glm <- glm(disease_presence ~ ., data = bio_scaled, family = binomial())
```

```{r}
#| label: day01-logistic-boundary
#| echo: false
#| fig-height: 4.5
#| fig-width: 7
plot_decision_boundary_logistic <- function(model, data, feature1, feature2) {
  r1 <- range(data[[feature1]])
  r2 <- range(data[[feature2]])
  grid <- expand.grid(
    seq(r1[1] - 1, r1[2] + 1, length.out = 160),
    seq(r2[1] - 1, r2[2] + 1, length.out = 160)
  )
  names(grid) <- c(feature1, feature2)
  for (feat in setdiff(names(data), c(feature1, feature2, "disease_presence"))) {
    grid[[feat]] <- mean(data[[feat]])
  }
  grid$response_prob <- predict(model, newdata = grid, type = "response")
  grid$response_cls <- factor(
    ifelse(grid$response_prob > 0.5, "Present", "Absent"),
    levels = c("Absent", "Present")
  )
  ggplot(data, aes(!!sym(feature1), !!sym(feature2))) +
    geom_tile(
      data = grid,
      aes(!!sym(feature1), !!sym(feature2), fill = response_cls),
      alpha = 0.28,
      inherit.aes = FALSE
    ) +
    geom_point(aes(fill = disease_presence), shape = 21, color = "black", size = 2.5) +
    scale_fill_manual(values = c(Absent = "turquoise", Present = "magenta")) +
    theme_minimal() +
    labs(
      title = "Logistic boundary (Gene1 vs Gene3; other genes at mean)",
      x = feature1, y = feature2, fill = "Class"
    )
}
print(plot_decision_boundary_logistic(full_glm, bio_scaled, "Gene1", "Gene3"))
```

## Null vs full: in-sample misclassification

```{r}
#| label: day01-logistic-null-vs-full
#| echo: false
#| fig-height: 3.5
#| fig-width: 6.5
class_err <- function(model, data) {
  prob <- predict(model, newdata = data, type = "response")
  pred <- factor(ifelse(prob > 0.5, "Present", "Absent"), levels = c("Absent", "Present"))
  mean(pred != data$disease_presence)
}
cmp_log <- data.frame(
  model = c("Null (intercept only)", "Full (all genes)"),
  AIC = c(AIC(null_glm), AIC(full_glm)),
  error_rate = c(class_err(null_glm, bio_scaled), class_err(full_glm, bio_scaled))
)
ggplot(cmp_log, aes(x = reorder(model, -error_rate), y = error_rate, fill = model)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "In-sample misclassification: null vs full",
    subtitle = "Null ignores genes; full uses all predictors",
    x = NULL, y = "Misclassification rate (same n samples)"
  ) +
  theme_minimal()
```

## AIC: backward and forward selection

```{r}
#| label: day01-logistic-step-aic
#| echo: false
#| results: hide
backward_glm <- stepAIC(full_glm, direction = "backward", trace = FALSE)
forward_glm <- stepAIC(
  null_glm,
  direction = "forward",
  scope = list(lower = null_glm, upper = full_glm),
  trace = FALSE
)
```

```{r}
#| label: day01-logistic-aic-sizes
#| echo: false
#| fig-height: 3.2
#| fig-width: 6.5
aic_log_tbl <- data.frame(
  approach = c("Null", "GLM full", "AIC backward", "AIC forward"),
  n_terms = c(
    length(coef(null_glm)),
    length(coef(full_glm)),
    length(coef(backward_glm)),
    length(coef(forward_glm))
  ),
  AIC = c(AIC(null_glm), AIC(full_glm), AIC(backward_glm), AIC(forward_glm))
)
ggplot(aic_log_tbl, aes(x = reorder(approach, AIC), y = AIC, fill = approach)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "AIC across logistic models (same data)", x = NULL, y = "AIC (lower is better)") +
  theme_minimal()
```

```{r}
#| label: day01-logistic-aic-terms
#| echo: false
#| results: asis
fmt_f <- function(f) paste(deparse(f), collapse = " ")
cat("**Backward AIC formula:** `", fmt_f(formula(backward_glm)), "`\n\n", sep = "")
cat("**Forward AIC formula:** `", fmt_f(formula(forward_glm)), "`\n\n", sep = "")
```
