**Gene × trait** — describe data → null vs full model → AIC forward/backward (all on the same $n$ samples).

- [Linear regression + LASSO (genes)](../notebooks/linear-regression-lasso.Rmd)

```{r}
#| label: day01-trait-libs
#| include: false
suppressPackageStartupMessages({
  library(broom)
  library(glmnet)
  library(recipes)
  library(dplyr)
  library(ggplot2)
  library(GGally)
  library(MASS)
  library(rcartocolor)
  library(tidyr)
})
my_pal <- carto_pal(12, "Bold")
theme_set(theme_classic())
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  out.width = "100%",
  dpi = 120
)
```

## Gene × trait: the dataset

**GeneA** and **GeneB** drive the trait; correlated noise genes; **GeneJ** correlated with GeneA but not causal.

```{r}
#| label: day01-linear-sim-data
#| echo: false
#| results: hide
set.seed(123)
n <- 150
p <- 10
gene_names <- c("GeneA", "GeneB", "GeneC", "GeneD", "GeneE", "GeneF", "GeneG", "GeneH", "GeneI", "GeneJ")
gene_data <- matrix(rnorm(n * p), nrow = n)
colnames(gene_data) <- gene_names
trait <- -0.5 * gene_data[, "GeneA"] + 0.8 * gene_data[, "GeneB"] + rnorm(n, sd = 0.3)
gene_data[, "GeneC"] <- gene_data[, "GeneA"] + rnorm(n, sd = 0.2)
gene_data[, "GeneD"] <- -0.3 * gene_data[, "GeneB"] + rnorm(n, sd = 0.2)
for (i in 5:p) gene_data[, gene_names[i]] <- rnorm(n)
gene_data[, "GeneF"] <- gene_data[, "GeneE"] + rnorm(n, sd = 0.2)
gene_data[, "GeneH"] <- -0.5 * gene_data[, "GeneG"] + rnorm(n, sd = 0.2)
gene_data[, "GeneJ"] <- 0.7 * gene_data[, "GeneA"] + rnorm(n, sd = 0.2)
data_trait <- as.data.frame(cbind(gene_data, trait))
```

```{r}
#| label: day01-correlation-heatmap
#| echo: false
#| fig-height: 4.5
#| fig-width: 8
correlation_matrix <- cor(data_trait)
cor_data <- as.data.frame(as.table(correlation_matrix)) |>
  dplyr::rename(variable1 = Var1, variable2 = Var2, correlation = Freq)
ggplot(cor_data, aes(variable1, variable2, fill = correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "dodgerblue", high = "firebrick4", mid = "white",
    midpoint = 0, limit = c(-1, 1), name = "r"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_fixed() +
  labs(title = "Correlation among genes and trait")
```

```{r}
#| label: day01-ggpairs-snippet
#| echo: false
#| fig-height: 5.5
#| fig-width: 8
ggpairs(
  data_trait,
  upper = list(continuous = wrap("cor", size = 2.2)),
  aes(alpha = 0.15)
) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| label: day01-trait-prep-full
#| echo: false
#| results: hide
rec_scale <- recipe(trait ~ ., data = data_trait) |>
  step_normalize(all_numeric_predictors())
data_scaled <- bake(prep(rec_scale, data_trait), data_trait)
X_s <- as.matrix(dplyr::select(data_scaled, -trait))
y <- data_scaled$trait
n_obs <- nrow(data_scaled)
```

## Null model (intercept only)

Too simple: ignores genes → **underfitting**.

```{r}
#| label: day01-lm-null
#| echo: false
#| results: hide
null_lm <- lm(trait ~ 1, data = data_scaled)
```

## Full model (all genes)

Every gene in the model — flexible, but many correlated predictors → unstable coefficients (**overfitting risk**).

```{r}
#| label: day01-lm-full
#| echo: false
#| results: hide
full_lm <- lm(trait ~ ., data = data_scaled)
```

## Regression table (full model)

```{r}
#| label: day01-lm-regression-table
#| echo: false
#| results: asis
reg_tbl <- broom::tidy(full_lm)
knitr::kable(head(reg_tbl, 6), digits = 3, caption = "OLS summary (first terms; all genes in notebook)")
```

```{r}
#| label: day01-lm-coef-plot
#| echo: false
#| fig-height: 4.2
#| fig-width: 8
cf <- coef(full_lm)
cf <- cf[names(cf) != "(Intercept)"]
ci <- tryCatch(
  confint(full_lm, parm = names(cf)),
  error = function(e) matrix(NA_real_, length(cf), 2, dimnames = list(names(cf), c("2.5 %", "97.5 %")))
)
coef_tbl <- data.frame(
  gene = names(cf),
  estimate = as.numeric(cf),
  lo = as.numeric(ci[, 1]),
  hi = as.numeric(ci[, 2])
)
ggplot(coef_tbl, aes(x = estimate, y = reorder(gene, estimate))) +
  geom_vline(xintercept = 0, linetype = 2, color = "gray50") +
  geom_pointrange(aes(xmin = lo, xmax = hi), fatten = 3) +
  labs(
    title = "OLS coefficients ± 95% CI (all predictors)",
    x = "Effect on trait (scaled gene units)", y = NULL
  ) +
  theme_minimal()
```

## Null vs full: in-sample fit

```{r}
#| label: day01-null-vs-full-complexity
#| echo: false
#| fig-height: 3.5
#| fig-width: 6.5
cmp_tbl <- data.frame(
  model = c("Null (intercept only)", "Full (all genes)"),
  AIC = c(AIC(null_lm), AIC(full_lm)),
  RMSE_in = c(
    sqrt(mean(residuals(null_lm)^2)),
    sqrt(mean(residuals(full_lm)^2))
  )
)
ggplot(cmp_tbl, aes(x = reorder(model, -RMSE_in), y = RMSE_in, fill = model)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "In-sample RMSE: too few vs too many predictors",
    subtitle = "Null underfits; full fits training data tighter (more parameters)",
    x = NULL, y = "RMSE (same n samples)"
  ) +
  theme_minimal()
```

## AIC: backward and forward selection

```{r}
#| label: day01-step-aic
#| echo: false
#| results: hide
backward_model <- stepAIC(full_lm, direction = "backward", trace = FALSE)
forward_model <- stepAIC(
  null_lm,
  direction = "forward",
  scope = list(lower = null_lm, upper = full_lm),
  trace = FALSE
)
```

```{r}
#| label: day01-aic-model-sizes
#| echo: false
#| fig-height: 3.2
#| fig-width: 6.5
aic_tbl <- data.frame(
  approach = c("Null", "OLS full", "AIC backward", "AIC forward"),
  n_terms = c(
    length(coef(null_lm)),
    length(coef(full_lm)),
    length(coef(backward_model)),
    length(coef(forward_model))
  ),
  AIC = c(AIC(null_lm), AIC(full_lm), AIC(backward_model), AIC(forward_model))
)
ggplot(aic_tbl, aes(x = reorder(approach, AIC), y = AIC, fill = approach)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "AIC across models (same data)", x = NULL, y = "AIC (lower is better)") +
  theme_minimal()
```

```{r}
#| label: day01-aic-terms-text
#| echo: false
#| results: asis
fmt_f <- function(f) paste(deparse(f), collapse = " ")
cat("**Backward AIC formula:** `", fmt_f(formula(backward_model)), "`\n\n", sep = "")
cat("**Forward AIC formula:** `", fmt_f(formula(forward_model)), "`\n\n", sep = "")
```

## AIC vs full OLS: coefficients (±2 SE)

Dashed lines: known effects for **GeneB** (0.8) and **GeneA** (−0.5).

```{r}
#| label: day01-aic-coef-compare
#| echo: false
#| fig-height: 4.5
#| fig-width: 9
eff_plot <- bind_rows(
  broom::tidy(backward_model) %>% mutate(model = "Backward"),
  broom::tidy(forward_model) %>% mutate(model = "Forward"),
  broom::tidy(full_lm) %>% mutate(model = "Full")
) %>%
  filter(term != "(Intercept)")

ggplot(eff_plot, aes(x = estimate, y = term, color = model)) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(xmin = estimate - 2 * std.error, xmax = estimate + 2 * std.error),
    width = 0.15,
    linewidth = 0.5
  ) +
  geom_vline(xintercept = c(0.8, -0.5), linetype = "dashed", color = "gray45") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "gray70") +
  coord_flip() +
  scale_color_manual(values = my_pal[c(1, 4, 8)]) +
  labs(
    title = "Coefficient estimates: full vs AIC backward vs AIC forward",
    x = "Estimate (scaled gene units)", y = NULL, color = NULL
  ) +
  theme_minimal()
```
