## Extension: imputation inside the recipe (`peng_na`)

Add **`step_impute_*()`** before `step_zv()` / dummies / normalize — never `drop_na()` on the full table before splitting.

## Artificial missing data: bill length

We set **~15%** of `bill_length_mm` (and a few `sex`) to **NA** to practice imputation — not a real sensor failure pattern.

```{r}
#| label: day04-na-scatter
#| echo: true
#| fig-width: 8
#| fig-height: 4.5
plot_na_highlight(
  peng_na, "bill_length_mm", "bill_depth_mm",
  title = "Missing bill length (red ×)"
)
```

## Wrong fix: drop rows with `drop_na()`

```{r}
#| label: day04-na-drop-demo
#| echo: true
tibble(
  rows_before = nrow(peng_na),
  rows_after_drop_na = nrow(tidyr::drop_na(peng_na))
)
```

Dropping loses penguins and can **bias** who remains — never drop on the **full** table before splitting.

## Right fix: impute inside the recipe

```{r}
#| label: day04-impute-recipe
#| echo: true
rec_impute <- recipe(y ~ ., data = peng_na) |>
  step_impute_median(bill_length_mm) |>
  step_impute_mode(sex) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())
```

## Before vs after imputation (plots)

```{r}
#| label: day04-impute-before-after
#| echo: true
#| fig-width: 10
#| fig-height: 4.5
prep_impute <- prep(rec_impute, peng_na)
peng_baked <- bake(prep_impute, peng_na)

p_before <- plot_na_highlight(
  peng_na, "bill_length_mm", "bill_depth_mm",
  title = "Before imputation"
)
p_after <- ggplot(peng_baked, aes(bill_length_mm, bill_depth_mm, color = y)) +
  geom_point(size = 2.4) +
  scale_color_manual(values = c(Adelie = "black", Gentoo = "gray40")) +
  theme_minimal() +
  labs(title = "After median/mode imputation", x = "bill_length_mm", y = "bill_depth_mm")

if (requireNamespace("patchwork", quietly = TRUE)) {
  print(p_before + p_after + patchwork::plot_layout(ncol = 2))
} else {
  print(p_before)
  print(p_after)
}
```

```{r}
#| label: day04-impute-na-count
#| echo: true
sum(is.na(bake(prep_impute, peng_na)$bill_length_mm))
```

## Imputation inside cross-validation

```{r}
#| label: day04-impute-cv
#| echo: true
#| message: false
xgb_na_spec <- boost_tree(trees = 80, tree_depth = 3, learn_rate = 0.08) |>
  set_engine("xgboost") |>
  set_mode("classification")
wf_na <- workflow() |>
  add_recipe(rec_impute) |>
  add_model(xgb_na_spec)
folds_na <- vfold_cv(peng_na, v = 5, strata = y)
set.seed(14)
rs_na <- fit_resamples(wf_na, folds_na, metrics = metrics_acc)
collect_metrics(rs_na) |>
  filter(.metric == "accuracy") |>
  select(mean, std_err)
```

The workflow accepts **NA** in raw rows; each fold **learns** imputation on analysis data only.

## Metrics and deployment (reminders)

- Pick **metrics before** peeking at holdout data; report **subgroup** tables when ethically appropriate.
- **Data card:** document NA handling and class prevalence **before** modeling.
- **Model card:** intended use, limits, monitoring — finish the draft you started Monday.

Continue: [Module 06 — Metrics & SHAP](../modules/module-06-evaluation-and-interpretability.qmd) · [Day 2 (Tuesday) pipeline](../slides/day-02-tuesday.qmd).
