## SHAP  -  what it means (game theory intuition)

- **One penguin, one prediction:** the model outputs **P(male)** (or log-odds). SHAP asks: *how much did each measurement move this bird away from a baseline?*
- **Players and coalitions:** treat each feature as a **player**. A **coalition** is any subset of features with the others **averaged over** the data. The model's output on that coalition is the coalition's **payout**.
- **Shapley value:** for each feature, average its **marginal contribution** over all orders in which features can join the coalition. That fair share is the **SHAP value** for feature *j* (written `phi_j`).
- **Additivity:** start from a baseline score, then add one credit per feature. For penguin *i*:

```
baseline + phi_i1 + phi_i2 + ...  ≈  P(male) for this bird
```

(all credits are for the **same** fitted model on the link/probability scale)

- **VIP vs SHAP:** **VIP** ranks predictors **globally** (how often splits use them). **SHAP** explains **one row** (why *this* penguin scored more male or female).

## SHAP task — sex without `species`

- **Outcome:** `sex` (female / male). **`species` is omitted** on purpose so the task is not trivial (see [sex notebook](../notebooks/penguins-sex-classification.html)).
- **Model:** same **`recipe` + random forest (`ranger`)** pattern as Tuesday; fit on all complete rows here so SHAP curves are stable in class (in production, fit SHAP on training data only).
- **Packages:** **`shapviz`** for plots; **`kernelshap`** to compute values on the fitted **`ranger`** forest.

```{r}
#| label: day04-shap-sex-setup
#| include: false
suppressPackageStartupMessages({
  library(shapviz)
  library(kernelshap)
})
pg_sex <- prep_penguins_sex()
sex_shap <- fit_sex_shap_bundle(pg_sex, trees = 300, n_sample = 80, n_bg = 35, seed = 11)
shp_sex <- sex_shap$shp
X_model <- bake_workflow_predictors(sex_shap$fit, pg_sex)
```

```{r}
#| label: day04-shap-sex-code
#| echo: true
#| eval: false
pg_sex <- prep_penguins_sex()

rec_sex <- recipe(sex ~ bill_length_mm + bill_depth_mm + flipper_length_mm +
                    body_mass_g + island + year, data = pg_sex) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

rf_sex_spec <- rand_forest(trees = 300, mtry = 3, min_n = 2) |>
  set_engine("ranger", probability = TRUE) |>
  set_mode("classification")

wf_sex <- workflow() |>
  add_recipe(rec_sex) |>
  add_model(rf_sex_spec)

fit_sex <- fit(wf_sex, pg_sex)
```

## From fitted workflow to SHAP values

1. **Fit** `workflow(recipe + forest)` on raw penguins.
2. **`bake()`** into **`X_model`** — predictors-only table in the forest's language (scaled bills, dummy `island`, …).
3. **Pick** birds to explain (`X_shap`) and a **reference flock** (`bg_X`) for coalition baselines.
4. **Perturb** baked columns many times: for each feature, ask *if this bird's value changed, how would **P(male)** move?*
5. **Average** those pushes into one credit per feature → **SHAP values** → beeswarm / waterfall plots.

**Why `bake()`?** The forest only saw **stage-2** columns (scaled numerics, dummies). SHAP must perturb features in that language — not raw mm/island labels. **`bake()` = recover the spreadsheet the trees learned from.**

## Where we intervene (schematic) {.flowchart-slide}

```{mermaid}
%%{init: {"theme":"base", "themeVariables": {"fontSize": "15px", "lineColor": "#2d2d2d", "clusterBkg": "#fafafa", "clusterBorder": "#555555"}, "flowchart": {"useMaxWidth": false, "htmlLabels": true, "wrappingWidth": 300, "padding": 14, "nodeSpacing": 36, "rankSpacing": 40}}}%%
flowchart TD
    Raw["Raw row from pg_sex"]

    subgraph TrainOnce ["Train once inside fit()"]
        Rec["Recipe<br/>zv, dummy, normalize"]
        BakeIn["Baked columns internal"]
        RF["ranger learns trees"]
        Raw --> Rec --> BakeIn --> RF
    end

    subgraph ShapLoop ["Explain after fit frozen model"]
        BakeOut["bake to X_model"]
        Intervene["INTERVENE HERE<br/>kernelshap edits baked columns"]
        Pred["pred_fun asks ranger<br/>P(male)"]
        ShapOut["SHAP credits"]
        BakeOut --> Intervene --> Pred --> ShapOut
    end

    Raw --> BakeOut
    RF -.->|"extract_fit_parsnip same forest"| Pred

    classDef train fill:#e8f1fb,stroke:#1f4f99,color:#0f2b56,stroke-width:2px;
    classDef shap fill:#ffffff,stroke:#3d3d3d,color:#1a1a1a,stroke-width:1.5px;
    classDef intervene fill:#fff3cd,stroke:#b8860b,color:#3d2f00,stroke-width:3px;
    classDef output fill:#e9f9ef,stroke:#1f7a3f,color:#124825,stroke-width:2px;

    class Raw,Rec,BakeIn,RF train;
    class BakeOut,Pred shap;
    class Intervene intervene;
    class ShapOut output;

    style TrainOnce fill:#f3f8ff,stroke:#1f4f99,stroke-width:1px,color:#0f2b56
    style ShapLoop fill:#fafafa,stroke:#555555,stroke-width:1px,color:#222222
```

- **Intervene on:** numeric columns in **`X_model`** (scaled bills, dummy islands, …).
- **Do not intervene on:** raw mm/island labels, recipe parameters, or tree weights (all frozen after `fit()`).
- **Why:** the forest only understands the baked spreadsheet — SHAP asks *what-if* questions in that language.

## `X_model` — bake the forest's spreadsheet

**`X_model`** = `bake(rec_prepped, new_data = pg_sex, all_predictors())`

- **Rows:** one per penguin (same birds as `pg_sex`, after recipe row handling).
- **Columns:** **predictors only** — numeric columns `ranger` actually split on. Outcome `sex` is omitted.
- **`kernelshap()`** edits rows of this object; **`X_shap`** = rows we explain; **`bg_X`** = background reference flock.

| Raw in `pg_sex` | Becomes in `X_model` |
|-----------------|----------------------|
| `bill_length_mm` (mm) | z-scored `bill_length_mm` |
| `island` (Biscoe / Dream / …) | dummy columns like `island_Biscoe` (0/1) |
| `sex` (female / male) | **omitted** — not a predictor here |

**Wrong:** pass raw `bill_length_mm = 45` while the forest expects a z-scored value. **Right:** pass the baked row from `X_model`.

```{r}
#| label: day04-shap-sex-bake
#| echo: true
#| eval: false
rec_prepped <- prep(extract_recipe(fit_sex, estimated = TRUE), training = pg_sex)
X_model <- bake(rec_prepped, new_data = pg_sex, all_predictors())

head(X_model, 3)
```

```{r}
#| label: day04-x-model-show
#| echo: false
knitr::kable(
  tibble::tibble(
    baked_column = names(X_model),
    row1_value = as.numeric(X_model[1, ])
  ),
  digits = 3,
  caption = paste0(
    "X_model has ", nrow(X_model), " rows and ", ncol(X_model),
    " baked predictor columns (row 1 shown)"
  )
)
```

## KernelSHAP — loop and code

For **one penguin** in `X_shap`: start from a **baseline** built from `bg_X`; perturb each baked feature; call **`pred_fun`** → **P(male)**; **average** moves into one SHAP value per feature (they sum ≈ this bird's final score).

- **`rf_engine`** — unwrap **`ranger`** from tidymodels via **`extract_fit_parsnip()`**
- **`pred_fun`** — baked matrix → **P(male)** (SHAP does not know about recipes)
- **`X_shap`** — ~80 random rows to explain; **`bg_X`** — ~35 background rows for coalitions
- **`kernelshap(...)`** then **`shapviz(...)`** for beeswarm / waterfall plots

```{r}
#| label: day04-shap-sex-shapviz
#| echo: true
#| eval: false
rf_engine <- extract_fit_parsnip(fit_sex)$fit

pred_fun <- function(object, X_new) {
  as.numeric(predict(object, data = X_new)$predictions[, "male"])
}

set.seed(11)
X_shap <- dplyr::slice_sample(X_model, n = 80)
bg_X <- dplyr::slice_sample(X_model, n = 35)

ks <- kernelshap(rf_engine, X = X_shap, pred_fun = pred_fun, bg_X = bg_X)
shp_sex <- shapviz(ks, X_pred = X_shap)
```

## Afternoon lab — Task 4.4

{{< include ../_includes/lab/d04-44.qmd >}}

## SHAP beeswarm  -  global pattern

```{r}
#| label: day04-shap-beeswarm
#| echo: false
#| fig-width: 9
#| fig-height: 6
#| out-width: 100%
plot_shap_beeswarm(shp_sex)
```

- **y-axis:** features; **x-axis:** SHAP contribution toward **P(male)**.
- **Color:** feature value (high vs low). Wide spread = this predictor matters **for many birds** in this model.

## SHAP waterfall — two local explanations

**Male** and **female** examples from the same fitted model — bars sum from baseline to **P(male)** for each bird.

```{r}
#| label: day04-shap-waterfall-male
#| echo: false
#| fig-width: 9
#| fig-height: 5.5
#| out-width: 100%
plot_shap_waterfall(
  shp_sex,
  row_id = sex_shap$male_row,
  title = "Waterfall: one male penguin",
  subtitle = "Bars sum from baseline to model P(male) for this bird"
)
```

```{r}
#| label: day04-shap-waterfall-female
#| echo: false
#| fig-width: 9
#| fig-height: 5.5
#| out-width: 100%
plot_shap_waterfall(
  shp_sex,
  row_id = sex_shap$female_row,
  title = "Waterfall: one female penguin",
  subtitle = "Same model  -  different feature pushes"
)
```

## Reading SHAP without over-claiming

**Reasonable claims**

- For **this** penguin, **higher body mass** (in the model, holding the coalition story fixed) pushed **P(male)** up or down.
- **Globally**, bill depth and mass show the widest SHAP spread in **this fitted RF**.
- SHAP describes **the trained model**, not a randomized experiment in the wild.

**Avoid**

- "Bill length **causes** males"  -  observational data; morphometrics are **correlated** ([data card](../data/cards/palmer-penguins.html)).
- "We should **intervene** on flipper length"  -  policy from association, not a causal effect shown by the model.

**Humility prompt (assessment):** *What would you **not** claim from this SHAP beeswarm?*

Full write-up: [SHAP notebook](../notebooks/penguins-shap.html)
