## Neural network (small MLP) — API literacy

- **`mlp(hidden_units = …, penalty = …, epochs = …) |> set_engine("nnet")`** — one hidden layer, weight decay, counts passes through the data.
- **Requires scaling** (already in `rec_base` via `step_normalize`).
- On **small tabular** data, trees often match or beat MLPs — still worth knowing the **same workflow** interface.

```{r}
#| label: day04-mlp-spec
#| echo: true
mlp_spec <- mlp(hidden_units = 8, penalty = 0.1, epochs = 150) |>
  set_engine("nnet", trace = FALSE) |>
  set_mode("classification")

wf_mlp <- workflow() |>
  add_recipe(rec_base) |>
  add_model(mlp_spec)
```

```{r}
#| label: day04-mlp-resamples
#| echo: true
#| message: false
set.seed(9)
rs_mlp <- fit_resamples(wf_mlp, folds, metrics = metrics_acc)
mlp_cmp <- bind_rows(
  collect_metrics(rs_rf) |> mutate(model = "Random forest"),
  collect_metrics(rs_xgb) |> mutate(model = "XGBoost"),
  collect_metrics(rs_mlp) |> mutate(model = "MLP (nnet)")
) |>
  filter(.metric == "accuracy")

mlp_cmp |>
  select(model, .metric, mean, std_err)
```

```{r}
#| label: day04-mlp-bar
#| echo: true
#| fig-width: 8
#| fig-height: 4
mlp_cmp |>
  ggplot(aes(model, mean, fill = model)) +
  geom_col(show.legend = FALSE) +
  geom_errorbar(aes(ymin = mean - std_err, ymax = mean + std_err), width = 0.12) +
  theme_minimal() +
  labs(title = "Same recipe — accuracy across specs", x = NULL, y = "Mean accuracy")
```

Three-way comparison: see the **cross-validated accuracy** bar chart on the Day 4 deck.
