## Extension: class imbalance (`peng_imb`)

**Recipe extension:** add **`step_upsample(y)`** (or **`step_downsample()`** on the majority) near the top of the chain — still inside CV folds.

## Artificial imbalance: rare Gentoo

We **downsampled** Gentoo (~10 vs ~150 Adelie). Real cohorts can look like this when the event is rare.

```{r}
#| label: day04-imb-balance-plot
#| echo: false
#| fig-width: 7
#| fig-height: 4
print(plot_class_balance(peng_imb, "Manufactured class imbalance"))
```

A model that always predicts **Adelie** can still show **high accuracy** — but **miss every Gentoo** (low **sensitivity**).

## Bill space is still separable

```{r}
#| label: day04-imb-scatter
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
ggplot(peng_imb, aes(bill_length_mm, bill_depth_mm, color = y)) +
  geom_point(size = 2.6) +
  scale_color_manual(values = c(Adelie = "turquoise", Gentoo = "magenta")) +
  theme_minimal() +
  labs(
    title = "Rare Gentoo — bills still separate species",
    x = "Bill length (mm)",
    y = "Bill depth (mm)"
  )
```

**Upsampling duplicates minority rows in training** — useful when the learner **ignores** the rare class, not when the boundary is already easy.

## Recipes on `peng_imb` (code)

```{r}
#| label: day04-imb-recipes
#| echo: true
#| message: false
xgb_imb_spec <- boost_tree(trees = 80, tree_depth = 3, learn_rate = 0.08) |>
  set_engine("xgboost") |>
  set_mode("classification")

rec_imb <- recipe(y ~ ., data = peng_imb) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

rec_imb_up <- recipe(y ~ ., data = peng_imb) |>
  step_upsample(y) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

rec_imb
rec_imb_up
```

```{r}
#| label: day04-imb-prep
#| echo: true
#| message: false
wf_imb <- workflow() |> add_recipe(rec_imb) |> add_model(xgb_imb_spec)
wf_imb_up <- workflow() |> add_recipe(rec_imb_up) |> add_model(xgb_imb_spec)
```

## Cross-validated metrics: no upsample

```{r}
#| label: day04-imb-no-upsample
#| echo: true
#| message: false
set.seed(11)
rs_imb <- fit_resamples(wf_imb, folds_imb, metrics = metrics_cls_imb)
metrics_summary_table(rs_imb, "XGBoost, no upsample")
```

On this DGP we often see **very high accuracy and ROC-AUC** already.

## Cross-validated metrics: with `step_upsample(y)`

```{r}
#| label: day04-imb-with-upsample
#| echo: true
#| message: false
set.seed(12)
rs_imb_up <- fit_resamples(wf_imb_up, folds_imb, metrics = metrics_cls_imb)
metrics_summary_table(rs_imb_up, "XGBoost + upsample")
```

**Upsampling is not a free lunch** — compare tables in the next section (confusion matrix, ROC/PR, facet plot).
