Visuals below follow the same simulations as the course notebooks:

- [`Examples/Day 1 - Linear Regression/LinearRegressionLasso.Rmd`](../../Examples/Day%201%20-%20Linear%20Regression/LinearRegressionLasso.Rmd) (trait + LASSO / ridge with `glmnet`)
- [`Examples/Day 1 - Logistic Regression/LogReg.Rmd`](../../Examples/Day%201%20-%20Logistic%20Regression/LogReg.Rmd) (disease + `ggpairs` + logistic boundary)

```{r}
#| label: day01-examples-libs
#| include: false
suppressPackageStartupMessages({
  library(glmnet)
  library(caret)
  library(dplyr)
  library(ggplot2)
  library(GGally)
  library(rlang)
})
theme_set(theme_classic())
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  out.width = "100%",
  dpi = 120
)
```

## Example: synthetic genes → continuous trait

Same data-generating story as the linear / LASSO notebook: GeneA and GeneB drive the trait; correlated noise genes; GeneJ correlated with GeneA but not causal.

```{r}
#| label: day01-linear-sim-data
#| echo: false
#| results: hide
set.seed(123)
n <- 150
p <- 10
gene_names <- c("GeneA", "GeneB", "GeneC", "GeneD", "GeneE", "GeneF", "GeneG", "GeneH", "GeneI", "GeneJ")
gene_data <- matrix(rnorm(n * p), nrow = n)
colnames(gene_data) <- gene_names
trait <- -0.5 * gene_data[, "GeneA"] + 0.8 * gene_data[, "GeneB"] + rnorm(n, sd = 0.3)
gene_data[, "GeneC"] <- gene_data[, "GeneA"] + rnorm(n, sd = 0.2)
gene_data[, "GeneD"] <- -0.3 * gene_data[, "GeneB"] + rnorm(n, sd = 0.2)
for (i in 5:p) gene_data[, gene_names[i]] <- rnorm(n)
gene_data[, "GeneF"] <- gene_data[, "GeneE"] + rnorm(n, sd = 0.2)
gene_data[, "GeneH"] <- -0.5 * gene_data[, "GeneG"] + rnorm(n, sd = 0.2)
gene_data[, "GeneJ"] <- 0.7 * gene_data[, "GeneA"] + rnorm(n, sd = 0.2)
data_trait <- as.data.frame(cbind(gene_data, trait))
```

```{r}
#| label: day01-correlation-heatmap
#| echo: false
#| fig-height: 4.5
#| fig-width: 8
correlation_matrix <- cor(data_trait)
cor_data <- as.data.frame(as.table(correlation_matrix)) |>
  dplyr::rename(variable1 = Var1, variable2 = Var2, correlation = Freq)
ggplot(cor_data, aes(variable1, variable2, fill = correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "dodgerblue", high = "firebrick4", mid = "white",
    midpoint = 0, limit = c(-1, 1), name = "r"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_fixed() +
  labs(title = "Correlation among genes and trait")
```

```{r}
#| label: day01-ggpairs-snippet
#| echo: false
#| fig-height: 5.5
#| fig-width: 8
ggpairs(
  data_trait,
  upper = list(continuous = wrap("cor", size = 2.2)),
  aes(alpha = 0.15)
) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Ridge vs Lasso: cross-validated error (`cv.glmnet`)

Train/test split and scaling match the original notebook (`createDataPartition`, `preProcess`).

```{r}
#| label: day01-train-test-glmnet
#| echo: false
#| results: hide
X <- as.matrix(dplyr::select(data_trait, -trait))
y <- data_trait$trait
trainIndex <- createDataPartition(y, p = 0.8, list = FALSE)
X_train <- X[trainIndex, , drop = FALSE]
y_train <- y[trainIndex]
X_test <- X[-trainIndex, , drop = FALSE]
y_test <- y[-trainIndex]
preproc <- preProcess(X_train, method = c("center", "scale"))
X_train_s <- predict(preproc, X_train)
X_test_s <- predict(preproc, X_test)
lam_seq <- 10^seq(-3, -0.1, length.out = 25)
cv_ridge <- cv.glmnet(X_train_s, y_train, alpha = 0, lambda = lam_seq)
cv_lasso <- cv.glmnet(X_train_s, y_train, alpha = 1, lambda = lam_seq)
```

```{r}
#| label: day01-cv-plots
#| echo: false
#| fig-height: 4
#| fig-width: 9
oldpar <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1), mgp = c(2, 0.7, 0))
plot(cv_ridge, main = "Ridge (alpha = 0)")
plot(cv_lasso, main = "Lasso (alpha = 1)")
par(oldpar)
```

```{r}
#| label: day01-coef-paths
#| echo: false
#| fig-height: 4.2
#| fig-width: 9
fit_ridge <- glmnet(X_train_s, y_train, alpha = 0, lambda = lam_seq)
fit_lasso <- glmnet(X_train_s, y_train, alpha = 1, lambda = lam_seq)
oldpar <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1), mgp = c(2, 0.7, 0))
plot(fit_ridge, xvar = "lambda", label = TRUE, main = "Ridge coefficients vs log(lambda)")
plot(fit_lasso, xvar = "lambda", label = TRUE, main = "Lasso coefficients vs log(lambda)")
par(oldpar)
```

## Example: gene expression → disease (logistic notebook)

Same simulation as `LogReg.Rmd`: causal roles for Gene1, Gene3, Gene4 and an interaction term in the generative linear predictor.

```{r}
#| label: day01-logistic-sim
#| echo: false
#| results: hide
set.seed(123)
n <- 200
p <- 10
gene_names <- paste0("Gene", 1:p)
genes_data <- data.frame(matrix(rnorm(n * p), nrow = n))
colnames(genes_data) <- gene_names
disease_presence <- ifelse(
  0.5 * genes_data$Gene1 + 0.8 * genes_data$Gene3 - 0.3 * genes_data$Gene4 -
    genes_data$Gene1 * genes_data$Gene3 * 2 + rnorm(n, sd = 0.5) > 0,
  1, 0
)
disease_presence <- factor(disease_presence, levels = c(0, 1), labels = c("Absent", "Present"))
bio_data <- data.frame(genes_data, disease_presence)
logistic_model <- glm(disease_presence ~ ., data = bio_data, family = binomial())
```

```{r}
#| label: day01-ggpairs-logistic
#| echo: false
#| fig-height: 5.2
#| fig-width: 8
ggpairs(bio_data, columns = 1:5, aes(color = disease_presence)) +
  theme_minimal()
```

```{r}
#| label: day01-logistic-boundary
#| echo: false
#| fig-height: 4.5
#| fig-width: 7
plot_decision_boundary_logistic <- function(model, data, feature1, feature2) {
  r1 <- range(data[[feature1]])
  r2 <- range(data[[feature2]])
  grid <- expand.grid(
    seq(r1[1] - 1, r1[2] + 1, length.out = 160),
    seq(r2[1] - 1, r2[2] + 1, length.out = 160)
  )
  names(grid) <- c(feature1, feature2)
  for (feat in setdiff(names(data), c(feature1, feature2, "disease_presence"))) {
    grid[[feat]] <- mean(data[[feat]])
  }
  grid$response_prob <- predict(model, newdata = grid, type = "response")
  grid$response_cls <- factor(
    ifelse(grid$response_prob > 0.5, "Present", "Absent"),
    levels = c("Absent", "Present")
  )
  ggplot(data, aes(!!sym(feature1), !!sym(feature2))) +
    geom_tile(
      data = grid,
      aes(!!sym(feature1), !!sym(feature2), fill = response_cls),
      alpha = 0.28,
      inherit.aes = FALSE
    ) +
    geom_point(aes(fill = disease_presence), shape = 21, color = "black", size = 2.5) +
    scale_fill_manual(values = c(Absent = "turquoise", Present = "magenta")) +
    theme_minimal() +
    labs(
      title = "Logistic boundary (Gene1 vs Gene3; other genes at mean)",
      x = feature1, y = feature2, fill = "Class"
    )
}
print(plot_decision_boundary_logistic(logistic_model, bio_data, "Gene1", "Gene3"))
```
