## Missing data: impute inside the recipe

- **`step_impute_mean()`** / **`step_impute_median()`** for numerics; **`step_impute_mode()`** for factors — statistics computed on **analysis** folds only when resampled.
- **Never** fill NA using test rows before scoring.

```{r}
#| label: day04-impute-demo
#| echo: true
#| message: false
peng_na <- peng_cls
set.seed(2)
na_idx <- sample(which(!is.na(peng_na$bill_length_mm)), size = 12)
peng_na$bill_length_mm[na_idx] <- NA

rec_impute <- recipe(y ~ ., data = peng_na) |>
  step_impute_median(bill_length_mm) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

prep_impute <- prep(rec_impute, peng_na)
sum(is.na(bake(prep_impute, peng_na)$bill_length_mm))
```

## Class imbalance: upsample in the recipe

- **`themis::step_upsample(y)`** draws extra **minority** rows **inside training** only (with replacement) before fitting.
- Alternatives: **`step_smote()`**, **class weights** in the engine, **threshold tuning** on predicted probabilities.

```{r}
#| label: day04-upsample-recipe
#| echo: true
rec_up <- recipe(y ~ ., data = peng_cls) |>
  step_upsample(y) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())
```

When **false negatives** cost more (miss a rare disease), tune **recall** / **sensitivity** and report **PR-AUC** — not accuracy alone ([Module 07](../modules/module-07-penguins-choose-metrics.qmd)).

## Metrics and deployment (reminders)

- Pick **metrics before** peeking at holdout data; report **subgroup** tables when ethically appropriate.
- **Data card:** document NA handling and class prevalence **before** modeling.
- **Model card:** intended use, limits, monitoring — finish the draft you started Monday.

Continue: [Module 06 — Metrics & SHAP](../modules/module-06-evaluation-and-interpretability.qmd) · [Day 2 pipeline](../slides/day-02-tuesday.qmd).
