## Boosting: watch the boundary sharpen

We fix **`tree_depth = 2`** and **`learn_rate = 0.3`**, then increase **`trees`** from 1 to 6. Each round adds small trees that correct leftover mistakes — the **Gentoo probability surface** becomes more **nonlinear**.

```{r}
#| label: day04-boost-boundary-prep
#| echo: false
#| results: hide
boost_trees <- 1:6
for (i in boost_trees) {
  spec_i <- boost_tree(
    trees = i,
    tree_depth = 2,
    learn_rate = 0.3
  ) |>
    set_engine("xgboost") |>
    set_mode("classification")
  wf_i <- workflow() |>
    add_recipe(rec_base) |>
    add_model(spec_i)
  assign(paste0("boost_fit_", i), fit(wf_i, peng_cls))
}
```

## Boosting round 1

```{r}
#| label: day04-boost-step-1
#| echo: false
#| fig-width: 9
#| fig-height: 5
print(
  plot_prob_boundary(
    get("boost_fit_1"), peng_cls,
    title = "XGBoost after 1 tree (bill length × bill depth)"
  )
)
```

One shallow tree: rough split — mostly **underfits** the Gentoo cloud.

## Boosting round 2

```{r}
#| label: day04-boost-step-2
#| echo: false
#| fig-width: 9
#| fig-height: 5
print(
  plot_prob_boundary(
    get("boost_fit_2"), peng_cls,
    title = "XGBoost after 2 trees"
  )
)
```

## Boosting round 3

```{r}
#| label: day04-boost-step-3
#| echo: false
#| fig-width: 9
#| fig-height: 5
print(
  plot_prob_boundary(
    get("boost_fit_3"), peng_cls,
    title = "XGBoost after 3 trees"
  )
)
```

## Boosting round 4

```{r}
#| label: day04-boost-step-4
#| echo: false
#| fig-width: 9
#| fig-height: 5
print(
  plot_prob_boundary(
    get("boost_fit_4"), peng_cls,
    title = "XGBoost after 4 trees"
  )
)
```

## Boosting round 5

```{r}
#| label: day04-boost-step-5
#| echo: false
#| fig-width: 9
#| fig-height: 5
print(
  plot_prob_boundary(
    get("boost_fit_5"), peng_cls,
    title = "XGBoost after 5 trees"
  )
)
```

## Boosting round 6

```{r}
#| label: day04-boost-step-6
#| echo: false
#| fig-width: 9
#| fig-height: 5
print(
  plot_prob_boundary(
    get("boost_fit_6"), peng_cls,
    title = "XGBoost after 6 trees"
  )
)
```

Contrast **bagging**: a random forest **averages** many independent trees — boundaries are often **smoother** than boosting’s staged corrections.
