## SHAP  -  what it means (game theory intuition)

- **One penguin, one prediction:** the model outputs **P(male)** (or log-odds). SHAP asks: *how much did each measurement move this bird away from a baseline?*
- **Players and coalitions:** treat each feature as a **player**. A **coalition** is any subset of features with the others **averaged over** the data. The model's output on that coalition is the coalition's **payout**.
- **Shapley value:** for each feature, average its **marginal contribution** over all orders in which features can join the coalition. That fair share is the **SHAP value** \(\phi_j\) for feature \(j\).
- **Additivity:** baseline \(\phi_0\) plus all feature contributions sums to the model output on the **link scale** (for this forest, **P(male)**):

$$\phi_0 + \sum_j \phi_{i,j} \approx \hat{p}_i(\text{male})$$

- **VIP vs SHAP:** **VIP** ranks predictors **globally** (how often splits use them). **SHAP** explains **one row** (why *this* penguin scored more male or female).

## SHAP task  -  sex without `species`

- **Outcome:** `sex` (female / male). **`species` is omitted** on purpose so the task is not trivial (see [sex notebook](../notebooks/penguins-sex-classification.Rmd)).
- **Model:** same **`recipe` + random forest (`ranger`)** pattern as Tuesday; we **fit on all complete rows** here so SHAP curves are stable in class (in production, fit SHAP on training data only).
- **Packages:** **`shapviz`** for plots; **`kernelshap`** to compute values on the fitted **`ranger`** forest.

```{r}
#| label: day04-shap-sex-setup
#| include: false
suppressPackageStartupMessages({
  library(shapviz)
  library(kernelshap)
})
pg_sex <- prep_penguins_sex()
sex_shap <- fit_sex_shap_bundle(pg_sex, trees = 300, n_sample = 80, n_bg = 35, seed = 11)
shp_sex <- sex_shap$shp
```

## Compute SHAP in R (workflow + `bake`)

```{r}
#| label: day04-shap-sex-code
#| echo: true
#| eval: false
pg_sex <- prep_penguins_sex()

rec_sex <- recipe(sex ~ bill_length_mm + bill_depth_mm + flipper_length_mm +
                    body_mass_g + island + year, data = pg_sex) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

rf_sex_spec <- rand_forest(trees = 300, mtry = 3, min_n = 2) |>
  set_engine("ranger", probability = TRUE) |>
  set_mode("classification")

wf_sex <- workflow() |>
  add_recipe(rec_sex) |>
  add_model(rf_sex_spec)

fit_sex <- fit(wf_sex, pg_sex)
```

## Apply the recipe with `bake()` (model-ready columns)

```{r}
#| label: day04-shap-sex-bake
#| echo: true
#| eval: false
rec_prepped <- prep(extract_recipe(fit_sex, estimated = TRUE), training = pg_sex)
X_model <- bake(rec_prepped, new_data = pg_sex, all_predictors())

head(X_model, 3)
```

- **`prep()`** learned scaling and dummy columns when we **`fit()`** the workflow.
- **`bake()`** produces the **same encoded matrix** the forest was trained on  -  required before **SHAP** (same idea as Tuesday's `prep` / `bake` demo).

## KernelSHAP on the `ranger` object

```{r}
#| label: day04-shap-sex-shapviz
#| echo: true
#| eval: false
rf_engine <- extract_fit_parsnip(fit_sex)$fit

pred_fun <- function(object, X_new) {
  as.numeric(predict(object, data = X_new)$predictions[, "male"])
}

set.seed(11)
X_shap <- dplyr::slice_sample(X_model, n = 80)
bg_X <- dplyr::slice_sample(X_model, n = 35)

ks <- kernelshap(rf_engine, X = X_shap, pred_fun = pred_fun, bg_X = bg_X)
shp_sex <- shapviz(ks, X_pred = X_shap)
```

- **`kernelshap()`** needs numeric predictors only  -  that is why we **`bake()`** first.
- **`shapviz()`** wraps results for **beeswarm** and **waterfall** plots.

## SHAP beeswarm  -  global pattern

```{r}
#| label: day04-shap-beeswarm
#| echo: false
#| fig-width: 9
#| fig-height: 6
#| out-width: 100%
plot_shap_beeswarm(shp_sex)
```

- **y-axis:** features; **x-axis:** SHAP contribution toward **P(male)**.
- **Color:** feature value (high vs low). Wide spread = this predictor matters **for many birds** in this model.

## SHAP waterfall  -  two local explanations

**Male example** (one bird labeled male in the SHAP sample):

```{r}
#| label: day04-shap-waterfall-male
#| echo: false
#| fig-width: 9
#| fig-height: 6
#| out-width: 100%
plot_shap_waterfall(
  shp_sex,
  row_id = sex_shap$male_row,
  title = "Waterfall: one male penguin",
  subtitle = "Bars sum from baseline to model P(male) for this bird"
)
```

**Female example** (contrasting profile):

```{r}
#| label: day04-shap-waterfall-female
#| echo: false
#| fig-width: 9
#| fig-height: 6
#| out-width: 100%
plot_shap_waterfall(
  shp_sex,
  row_id = sex_shap$female_row,
  title = "Waterfall: one female penguin",
  subtitle = "Same model  -  different feature pushes"
)
```

## Reading SHAP without over-claiming

**Reasonable claims**

- For **this** penguin, **higher body mass** (in the model, holding the coalition story fixed) pushed **P(male)** up or down.
- **Globally**, bill depth and mass show the widest SHAP spread in **this fitted RF**.
- SHAP describes **the trained model**, not a randomized experiment in the wild.

**Avoid**

- "Bill length **causes** males"  -  observational data; morphometrics are **correlated** ([data card](../data/cards/palmer-penguins.qmd)).
- "We should **intervene** on flipper length"  -  policy from association.
- "SHAP proves **fairness** by island"  -  needs explicit subgroup metrics, not one plot.

**Humility prompt (assessment):** *What would you **not** claim from this SHAP beeswarm?*

Full write-up: [Module 06  -  SHAP](../modules/module-06-evaluation-and-interpretability.qmd#shap-shapley-values-on-penguins-sex)
