**Gene × disease — decision trees (`rpart`)** on the same simulation as Monday ([logistic notebook](../notebooks/logistic-regression-gene-disease.html)).

```{r}
#| label: day02-gene-rpart-setup
#| include: false
suppressPackageStartupMessages({
  library(rpart)
  library(rpart.plot)
  library(ggplot2)
  library(rlang)
})
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  out.width = "100%",
  dpi = 120
)

plot_rpart_boundary <- function(model, data, feature1, feature2, title = NULL) {
  r1 <- range(data[[feature1]])
  r2 <- range(data[[feature2]])
  grid <- expand.grid(
    seq(r1[1] - 1, r1[2] + 1, length.out = 180),
    seq(r2[1] - 1, r2[2] + 1, length.out = 180)
  )
  names(grid) <- c(feature1, feature2)
  for (feat in setdiff(names(data), c(feature1, feature2, "disease_presence"))) {
    grid[[feat]] <- mean(data[[feat]])
  }
  grid$response <- predict(model, newdata = grid, type = "class")
  grid$response <- factor(grid$response, levels = c("Absent", "Present"))
  ggplot(data, aes(!!sym(feature1), !!sym(feature2))) +
    geom_tile(
      data = grid,
      aes(!!sym(feature1), !!sym(feature2), fill = response),
      alpha = 0.32,
      inherit.aes = FALSE
    ) +
    geom_point(aes(fill = disease_presence), shape = 21, color = "black", size = 2.2) +
    scale_fill_manual(values = c(Absent = "turquoise", Present = "magenta")) +
    theme_minimal() +
    labs(
      title = title,
      x = feature1,
      y = feature2,
      fill = "Predicted"
    )
}

save_growth_composite <- function(model, depth, chunk_id, feature1 = "Gene1", feature2 = "Gene3") {
  path <- paste0(knitr::fig_chunk(chunk_id), ".png")
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  if (requireNamespace("ragg", quietly = TRUE)) {
    ragg::agg_png(path, width = 12, height = 6, units = "in", res = 120)
  } else {
    png(path, width = 12, height = 6, units = "in", res = 120)
  }
  on.exit(dev.off(), add = TRUE)
  layout(matrix(1:2, nrow = 1, ncol = 2))
  par(mar = c(1, 1, 3, 1))
  rpart.plot(
    model,
    type = 4,
    extra = 102,
    box.palette = "BuGn",
    fallen.leaves = TRUE,
    main = paste0("Tree (maxdepth = ", depth, ")"),
    cex = 0.85
  )
  print(
    plot_rpart_boundary(
      model, bio_data, feature1, feature2,
      title = paste0("Partition after depth ", depth)
    )
  )
  path
}

describe_top_split <- function(model) {
  if (is.null(model$splits) || nrow(model$splits) == 0) {
    return("At this depth the tree matches the previous level (no new split).")
  }
  var <- rownames(model$splits)[1]
  cut <- model$splits[1, "index"]
  paste0("Top split: **", var, "** at ", signif(cut, 3), " (left vs right branches in the tree plot).")
}
```

```{r}
#| label: day02-gene-sim
#| echo: false
#| results: hide
set.seed(123)
n <- 200
p <- 10
gene_names <- paste0("Gene", 1:p)
genes_data <- data.frame(matrix(rnorm(n * p), nrow = n))
colnames(genes_data) <- gene_names
disease_presence <- ifelse(
  0.5 * genes_data$Gene1 + 0.8 * genes_data$Gene3 - 0.3 * genes_data$Gene4 -
    genes_data$Gene1 * genes_data$Gene3 * 2 + rnorm(n, sd = 0.5) > 0,
  1, 0
)
disease_presence <- factor(disease_presence, levels = c(0, 1), labels = c("Absent", "Present"))
bio_data <- data.frame(genes_data, disease_presence)
```

```{r}
#| label: day02-rpart-final
#| echo: false
#| results: hide
rpart_model <- rpart(
  disease_presence ~ .,
  data = bio_data,
  method = "class",
  control = rpart.control(cp = 0.01, minsplit = 10)
)
```

## Final decision surface (Gene1 vs Gene3)

Other genes fixed at their sample means — same 2D slice as Monday’s logistic boundary.

```{r}
#| label: day02-rpart-final-boundary
#| echo: false
#| fig-height: 5
#| fig-width: 8
print(
  plot_rpart_boundary(
    rpart_model, bio_data, "Gene1", "Gene3",
    title = "rpart: predicted class in the Gene1–Gene3 plane"
  )
)
```

## Final tree (`rpart.plot`)

```{r}
#| label: day02-rpart-final-tree
#| echo: false
#| fig-height: 6
#| fig-width: 8
par(mar = c(1, 1, 3, 1))
rpart.plot(
  rpart_model,
  type = 4,
  extra = 102,
  box.palette = "BuGn",
  fallen.leaves = TRUE,
  main = "Final classification tree (all genes)"
)
```

```{r}
#| label: day02-rpart-step-prep
#| echo: false
#| results: hide
max_step_depth <- 5L
for (d in seq_len(max_step_depth)) {
  tree_d <- rpart(
    disease_presence ~ .,
    data = bio_data,
    method = "class",
    control = rpart.control(maxdepth = d, cp = 0.001, minsplit = 15)
  )
  assign(paste0("tree_depth_", d), tree_d)
}
```

## Growing the tree (step by step)

One split level at a time: **tree diagram** (left) and **decision surface** in Gene1 × Gene3 (right).

## Growing the tree: depth 1

```{r}
#| label: day02-rpart-step-1-fig
#| echo: false
#| fig-width: 10
#| fig-height: 5
#| out-width: 100%
knitr::include_graphics(save_growth_composite(get("tree_depth_1"), 1, "day02-rpart-step-1-fig"))
```

```{r}
#| label: day02-rpart-step-1-text
#| echo: false
#| results: asis
cat(describe_top_split(get("tree_depth_1")), "\n")
```

## Growing the tree: depth 2

```{r}
#| label: day02-rpart-step-2-fig
#| echo: false
#| fig-width: 10
#| fig-height: 5
#| out-width: 100%
knitr::include_graphics(save_growth_composite(get("tree_depth_2"), 2, "day02-rpart-step-2-fig"))
```

```{r}
#| label: day02-rpart-step-2-text
#| echo: false
#| results: asis
cat(describe_top_split(get("tree_depth_2")), "\n")
```

## Growing the tree: depth 3

```{r}
#| label: day02-rpart-step-3-fig
#| echo: false
#| fig-width: 10
#| fig-height: 5
#| out-width: 100%
knitr::include_graphics(save_growth_composite(get("tree_depth_3"), 3, "day02-rpart-step-3-fig"))
```

```{r}
#| label: day02-rpart-step-3-text
#| echo: false
#| results: asis
cat(describe_top_split(get("tree_depth_3")), "\n")
```

## Growing the tree: depth 4

```{r}
#| label: day02-rpart-step-4-fig
#| echo: false
#| fig-width: 10
#| fig-height: 5.5
#| out-width: 100%
knitr::include_graphics(save_growth_composite(get("tree_depth_4"), 4, "day02-rpart-step-4-fig"))
```

```{r}
#| label: day02-rpart-step-4-text
#| echo: false
#| results: asis
cat(describe_top_split(get("tree_depth_4")), "\n")
```

## Growing the tree: depth 5

```{r}
#| label: day02-rpart-step-5-fig
#| echo: false
#| fig-width: 10
#| fig-height: 5.5
#| out-width: 100%
knitr::include_graphics(save_growth_composite(get("tree_depth_5"), 5, "day02-rpart-step-5-fig"))
```

```{r}
#| label: day02-rpart-step-5-text
#| echo: false
#| results: asis
cat(describe_top_split(get("tree_depth_5")), "\n")
```

## Tree vs logistic (same data, in-sample accuracy)

```{r}
#| label: day02-rpart-accuracy
#| echo: false
#| results: asis
logistic_model <- glm(disease_presence ~ ., data = bio_data, family = binomial())
log_prob <- predict(logistic_model, type = "response")
log_pred <- factor(
  ifelse(log_prob > 0.5, "Present", "Absent"),
  levels = c("Absent", "Present")
)
rpart_pred <- predict(rpart_model, type = "class")
knitr::kable(
  data.frame(
    Model = c("Logistic (glm)", "Decision tree (rpart)"),
    Accuracy_in = c(
      mean(log_pred == bio_data$disease_presence),
      mean(rpart_pred == bio_data$disease_presence)
    )
  ),
  digits = 3,
  caption = "In-sample accuracy on all n rows (Monday vs Tuesday model class)"
)
```
