## Extension: PCA inside the recipe (`peng_pca`)

Add **`step_pca()`** to the Part A recipe chain — fit on training folds only, same as `step_normalize()`.

## PCA motivation: correlated measurements

For PCA we add back **flipper length** and **body mass** — bills and size track together on penguins.

```{r}
#| label: day04-pca-corr
#| echo: true
#| fig-width: 7
#| fig-height: 5
num_cols <- c("bill_length_mm", "bill_depth_mm", "flipper_length_mm", "body_mass_g", "year")
cor_pca <- cor(peng_pca[, num_cols])
cor_df <- as.data.frame(as.table(cor_pca)) |>
  dplyr::rename(var1 = Var1, var2 = Var2, r = Freq)
ggplot(cor_df, aes(var1, var2, fill = r)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick", midpoint = 0, limit = c(-1, 1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_fixed() +
  labs(title = "Correlated morphometrics (why PCA?)", fill = "r")
```

## Recipe with `step_pca()`

```{r}
#| label: day04-pca-recipe-code
#| echo: true
rec_pca <- recipe(y ~ ., data = peng_pca) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_pca(all_numeric_predictors(), num_comp = 2)

rec_pca
```

```{r}
#| label: day04-pca-variance
#| echo: false
#| fig-width: 7
#| fig-height: 4
num_scaled <- peng_pca |>
  dplyr::select(dplyr::all_of(num_cols)) |>
  scale()
pc_fit <- prcomp(num_scaled, center = FALSE, scale. = FALSE)
var_pct <- pc_fit$sdev^2 / sum(pc_fit$sdev^2) * 100
var_df <- tibble(PC = paste0("PC", seq_along(var_pct)), variance_pct = var_pct)
ggplot(var_df, aes(PC, variance_pct)) +
  geom_col(fill = "steelblue") +
  theme_minimal() +
  labs(
    title = "Variance explained (same numerics as step_pca)",
    x = NULL,
    y = "% variance"
  )
```

## Model with PCA vs without (same folds)

```{r}
#| label: day04-pca-prep-models
#| echo: false
#| results: hide
xgb_pca_spec <- boost_tree(trees = 100, tree_depth = 3, learn_rate = 0.05) |>
  set_engine("xgboost") |>
  set_mode("classification")

rec_no_pca <- recipe(y ~ ., data = peng_pca) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

wf_pca <- workflow() |> add_recipe(rec_pca) |> add_model(xgb_pca_spec)
wf_no_pca <- workflow() |> add_recipe(rec_no_pca) |> add_model(xgb_pca_spec)
```

```{r}
#| label: day04-pca-fit-compare
#| echo: true
#| message: false
set.seed(8)
rs_pca <- fit_resamples(wf_pca, folds_pca, metrics = metrics_acc)
set.seed(8)
rs_no_pca <- fit_resamples(wf_no_pca, folds_pca, metrics = metrics_acc)
bind_rows(
  collect_metrics(rs_pca) |> mutate(model = "XGB + PCA (2 comp.)"),
  collect_metrics(rs_no_pca) |> mutate(model = "XGB, no PCA")
) |>
  filter(.metric == "accuracy") |>
  select(model, mean, std_err)
```

```{r}
#| label: day04-pca-auc-bar
#| echo: false
#| fig-width: 7
#| fig-height: 3.5
bind_rows(
  collect_metrics(rs_pca) |> mutate(model = "With PCA"),
  collect_metrics(rs_no_pca) |> mutate(model = "No PCA")
) |>
  filter(.metric == "accuracy") |>
  ggplot(aes(model, mean, fill = model)) +
  geom_col(show.legend = FALSE) +
  geom_errorbar(aes(ymin = mean - std_err, ymax = mean + std_err), width = 0.12) +
  theme_minimal() +
  labs(title = "Accuracy: PCA vs all scaled numerics", x = NULL, y = "Mean accuracy")
```

**Tradeoff:** components are **not** “bill mm” anymore — VIP on PCA columns is harder to explain biologically.

**Leakage check:** always fit PCA **inside** `workflow()` + resampling, never on the full table before splitting.
