## Same recipe, swap the `spec`

Part A gave **`wf_tree`** (`rec_base` + `tree_spec`). Below: **random forest** (bagging) and **XGBoost** (boosting) on the same **`folds`**.

## Bagging: random forests

- **Many trees** on **bootstrap** samples of rows; each split sees only a **random subset** of predictors → lower variance than one deep tree.
- **`rand_forest(mtry = …, trees = …, min_n = …) |> set_engine("ranger")`**

## Boosting: gradient boosted trees

- Trees are added **sequentially**; each new tree targets **errors left by the ensemble** so far.
- **`boost_tree(trees = …, tree_depth = …, learn_rate = …) |> set_engine("xgboost")`**
- **`learn_rate`**: shrink each tree’s contribution (smaller → more trees needed, often smoother).

## Forest and boosting specs (code)

```{r}
#| label: day04-ensemble-specs
#| echo: true
#| message: false
rf_spec <- rand_forest(mtry = 3, trees = 400, min_n = 2) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("classification")
xgb_spec <- boost_tree(trees = 150, tree_depth = 3, learn_rate = 0.05) |>
  set_engine("xgboost") |>
  set_mode("classification")

list(tree = tree_spec, rf = rf_spec, xgb = xgb_spec)
```

```{r}
#| label: day04-ensemble-prep
#| echo: true
#| message: false
wf_rf <- workflow() |> add_recipe(rec_base) |> add_model(rf_spec)
wf_xgb <- workflow() |> add_recipe(rec_base) |> add_model(xgb_spec)
```

### Cross-validated ROC AUC (code + table)

```{r}
#| label: day04-ensemble-resamples
#| echo: true
#| message: false
set.seed(7)
rs_rf <- fit_resamples(wf_rf, folds, metrics = metrics_acc)
rs_xgb <- fit_resamples(wf_xgb, folds, metrics = metrics_acc)

cmp_ens <- bind_rows(
  collect_metrics(rs_tree) |> mutate(model = "Single tree"),
  collect_metrics(rs_rf) |> mutate(model = "Random forest"),
  collect_metrics(rs_xgb) |> mutate(model = "XGBoost")
) |>
  filter(.metric == "accuracy")

cmp_ens |>
  select(model, mean, std_err) |>
  arrange(desc(mean))
```

```{r}
#| label: day04-ensemble-auc-bar
#| echo: true
#| fig-width: 8
#| fig-height: 4
cmp_ens |>
  ggplot(aes(reorder(model, mean), mean, fill = model)) +
  geom_col(show.legend = FALSE) +
  geom_errorbar(aes(ymin = mean - std_err, ymax = mean + std_err), width = 0.12) +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Same folds, same recipe — different spec (accuracy)",
    x = NULL,
    y = "Mean accuracy"
  )
```

Fixed settings here for speed — in projects, wrap each spec in **`tune_grid()`** like [Module 04](../modules/module-04-pipeline.qmd#canonical-pipeline-tuesday).
