## Shared pipeline — libraries & example data

```{r}
#| label: tm-lib-data
#| echo: true
#| message: false
#| warning: false
suppressPackageStartupMessages({
  library(tidymodels)
  library(vip)
  library(rpart.plot)
  library(palmerpenguins)
  library(dplyr)
  library(tidyr)
})

# Adelie vs Gentoo (binary); complete cases — matches penguins-classification.Rmd
peng_cls <- palmerpenguins::penguins |>
  filter(species %in% c("Adelie", "Gentoo")) |>
  mutate(
    y = factor(species, levels = c("Adelie", "Gentoo")),
    year = as.numeric(year)
  ) |>
  dplyr::select(-species) |>
  tidyr::drop_na()
```

Binary outcome **`y`**: Adelie vs Gentoo on complete-case Palmer Penguins (bill, flipper, mass, island, sex, year). Same story as [penguins-classification.Rmd](../notebooks/penguins-classification.Rmd).

## Shared pipeline — recipe

```{r}
#| label: tm-recipe
#| echo: true
rec <- recipe(y ~ ., data = peng_cls) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())
```

Preprocessing lives in the **recipe** (impute, encode, normalize, …). Resampling fits this recipe **inside each fold** so nothing leaks from “future” rows.

## Shared pipeline — model spec

```{r}
#| label: tm-spec
#| echo: true
mod_spec <- decision_tree(
  tree_depth = tune(),
  min_n = tune()
) |>
  set_engine("rpart") |>
  set_mode("classification")
```

Swap **`mod_spec`** (and `set_engine`) for `rand_forest`, `boost_tree` + `xgboost`, `mlp`, etc.; keep the same workflow + resampling when you want **fair** comparisons.

## Shared pipeline — workflow

```{r}
#| label: tm-workflow
#| echo: true
wf <- workflow() |>
  add_recipe(rec) |>
  add_model(mod_spec)
```

One **`workflow()`** object = recipe + model, tuned and fitted as a **single** unit.

## Shared pipeline — resampling, metrics & tuning

```{r}
#| label: tm-tune
#| echo: true
folds <- vfold_cv(peng_cls, v = 5, strata = y)
metrics <- metric_set(roc_auc, accuracy, sens, spec)

tune_res <- tune_grid(
  wf,
  resamples = folds,
  grid = grid_regular(
    tree_depth(range = c(2, 6)),
    min_n(range = c(2, 20)),
    levels = 3
  ),
  metrics = metrics
)
```

**`metric_set`** fixes the scorecard before you peek at held-out data. **`tune_grid`** fits the workflow on each analysis set and evaluates on assessment folds.

## Shared pipeline — finalize & fit on all rows

```{r}
#| label: tm-final
#| echo: true
best <- select_best(tune_res, metric = "roc_auc")
final_wf <- finalize_workflow(wf, best)
final_fit <- fit(final_wf, data = peng_cls)
```

In a real project you would **`fit(final_wf, data = training(split))`** after choosing hyperparameters using **training data only**, then assess once on **`testing(split)`**. Here we fit on all of `peng_cls` for a compact teaching demo.

```{r}
#| label: tm-optional-split
#| echo: true
#| eval: false
# --- Optional: last_fit with a held-out split for honest performance ---------
# data_split <- initial_split(peng_cls, prop = 0.8, strata = y)
# train <- training(data_split)
# test  <- testing(data_split)
# last_fit(final_wf, split = data_split)  # after finalize_workflow(...)
```

## Figure — hyperparameter tuning

```{r}
#| label: tm-fig-autoplot
#| echo: true
#| fig-width: 10
#| fig-height: 5.2
tune_res |>
  autoplot() +
  theme_minimal() +
  labs(
    title = "Hyperparameter tuning",
    subtitle = "ROC AUC, accuracy, sensitivity, specificity across tree depth and min_n"
  )
```

Use **`autoplot()`** on `tune_res` to see how metrics move with the grid; follow up with **`collect_metrics()`** for tables.

## Figure — variable importance (model-based)

```{r}
#| label: tm-fig-vip
#| echo: true
#| fig-width: 9
#| fig-height: 5
final_fit |>
  extract_fit_parsnip() |>
  vip(geom = "point", aesthetics = list(color = "midnightblue", size = 3)) +
  theme_minimal() +
  labs(
    title = "Feature importance",
    subtitle = "Which predictors matter most for Adelie vs Gentoo in this tree?"
  )
```

**`vip`** summarizes how much this fitted learner leaned on each input; interpretation still depends on **correlation**, **measurement error**, and **causal** claims you do *not* make from importance alone.

## Figure — fitted tree structure

```{r}
#| label: tm-fig-rpart
#| echo: true
#| fig-width: 10
#| fig-height: 6
final_fit |>
  extract_fit_engine() |>
  rpart.plot(
    roundint = FALSE,
    type = 4,
    extra = 104,
    box.palette = "RdBu",
    main = "Final decision tree (rpart engine)"
  )
```

Connect this picture back to **partitioning** intuition from lecture: each box is a region of predictor space after sequential splits.

## Figure — ROC curve (in-sample look)

```{r}
#| label: tm-fig-roc
#| echo: true
#| fig-width: 8
#| fig-height: 5
augment(final_fit, new_data = peng_cls) |>
  roc_curve(truth = y, .pred_Gentoo, event_level = "second") |>
  autoplot() +
  labs(
    title = "ROC curve",
    subtitle = "Gentoo as event class; in-sample fit — use test data in practice"
  )
```

**`.pred_Gentoo`** with **`event_level = \"second\"`** matches `y` with levels Adelie (first), Gentoo (second). For honest performance, evaluate on **`testing(split)`** or use **`last_fit()`**.

*Edit only `teaching-hub/_includes/tidymodels-pipeline.qmd`; slide decks and Module 04 pull it in with Quarto `include`.*
