## Tidymodels walk-through — what you will see

We use **one table** all the way through: Palmer Penguins restricted to **Adelie vs Gentoo** (same as [penguins-classification.Rmd](../notebooks/penguins-classification.Rmd)).

Each **next block** follows the same rhythm: **first** intuition tied to penguin columns, **then** the matching R code (uses the **`tidymodels`** packages — names like `recipe` and `workflow` are just helpers around ideas you know from `lm`/`glm`).

You will move through: **data → recipe → model spec → workflow → resampling + tuning → finalize + fit → four diagnostic figures**.

### Mini-glossary (read once)

- **`peng_cls`** — the cleaned penguin table used in every chunk below (Adelie vs Gentoo, complete cases).
- **Recipe** — a reusable list of preprocessing steps (dummy coding for factors, scaling numbers, …) learned **inside** each cross-validation fold.
- **Model spec** — the **type** of model and which knobs are unknown **before** tuning (here: depth and minimum leaf size of a tree); not the fitted tree yet.
- **Workflow** — preprocessing + model stuck together so the computer never “forgets” to redo the preprocessing when you refit.
- **`vfold_cv`** — split data into *k* folds for **cross-validation** (here \(k=5\)), stratifying on `y` so each fold keeps both species.
- **`tune_grid`** — try a grid of tuning values; each fit uses only **training folds**, each score uses the **held-out fold**.
- **`finalize_workflow` + `fit`** — plug in the winning tuning values, then fit (in class we fit on **all** rows for speed; in research, reserve a true test set).

## Step 0 — The prediction task (Palmer Penguins)

- **Outcome `y`:** species with two levels, Adelie then Gentoo (so later ROC uses **Gentoo** as the second / “event” level).
- **Predictors:** bill length & depth, flipper length, body mass (all mm or g), plus **island**, **sex**, and **year** — mixed **numeric** and **nominal** columns, like most real labs.
- **Complete cases only:** `drop_na()` drops rows missing sex or measurements (talk in class about how much *N* you lose).


## Step 0 — Load packages and build `peng_cls`

```{r}
#| label: tm-lib-data
#| echo: true
#| message: false
#| warning: false
suppressPackageStartupMessages({
  library(tidymodels)
  library(vip)
  library(rpart.plot)
  library(palmerpenguins)
  library(dplyr)
  library(tidyr)
})

# Adelie vs Gentoo (binary); complete cases — matches penguins-classification.Rmd
peng_cls <- palmerpenguins::penguins |>
  filter(species %in% c("Adelie", "Gentoo")) |>
  mutate(
    y = factor(species, levels = c("Adelie", "Gentoo")),
    year = as.numeric(year)
  ) |>
  dplyr::select(-species) |>
  tidyr::drop_na()
```


## Step 0b — Inspect the rows we model (code)

```{r}
#| label: tm-glimpse
#| echo: true
#| message: false
dplyr::glimpse(peng_cls)
dplyr::count(peng_cls, y, name = "n")
```

Use **`glimpse`** so students see **types** (double vs factor) before we talk about recipe steps.


## Step 1 — Why a **recipe**? (Penguins vocabulary)

- **`recipe(y ~ ., data = peng_cls)`** means: “every other column helps predict `y`.”
- **`step_dummy(all_nominal_predictors())`** turns **island** and **sex** into numeric **0/1** columns so the tree engine can split on them.
- **`step_normalize(all_numeric_predictors())`** puts bill length, flipper length, body mass, year on a comparable scale before the tree sees them (important when you mix mm, g, and year).
- **`step_zv(all_predictors())`** drops **zero-variance** dummies if a level barely appears — avoids useless columns.

All of this is **declared now** and **fitted inside each cross-validation fold** later, which is what prevents **leakage** from future rows.


## Step 1 — Define the recipe (code)

```{r}
#| label: tm-recipe
#| echo: true
rec <- recipe(y ~ ., data = peng_cls) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())
```


## Step 2 — **Model spec** vs fitted model (penguins + trees)

- Here we pick a **classification tree** (`rpart` under the hood): good for teaching **splits** you can draw on the board.
- **`tree_depth = tune()`** and **`min_n = tune()`** say: “we will **try several** tree sizes and let cross-validation pick,” instead of guessing once by hand.
- **`set_mode("classification")`** matches our factor outcome **`y`** (Adelie / Gentoo).

Nothing is **fitted** yet — this is only the **blueprint** (parsnip’s job).


## Step 2 — Model specification (code)

```{r}
#| label: tm-spec
#| echo: true
mod_spec <- decision_tree(
  tree_depth = tune(),
  min_n = tune()
) |>
  set_engine("rpart") |>
  set_mode("classification")
```

Swap **`mod_spec`** (and `set_engine`) for `rand_forest`, `boost_tree` + `xgboost`, `mlp`, etc.; keep the same workflow + resampling when you want **fair** comparisons.


## Step 3 — **Workflow** = recipe + model (why one object?)

- **`workflow() |> add_recipe(rec) |> add_model(mod_spec)`** glues preprocessing + model into **one** object.
- **Why it matters for students:** when we cross-validate next, each fold **re-fits the recipe on the analysis rows** and evaluates on the held-out fold — the computer treats preprocessing as part of the model, not a forgotten earlier script.


## Step 3 — Build the workflow (code)

```{r}
#| label: tm-workflow
#| echo: true
wf <- workflow() |>
  add_recipe(rec) |>
  add_model(mod_spec)
```


## Step 4 — Resampling & metrics on penguins (concepts)

- **`vfold_cv(peng_cls, v = 5, strata = y)`**: five folds; **`strata = y`** tries to keep Adelie/Gentoo **balanced** in each fold so no fold is “all one species.”
- **`metric_set(roc_auc, accuracy, sens, spec)`**: we commit to **these** scores before looking at a true test set — good classroom habit.
- **`tune_grid(...)`**: for each combination of `tree_depth` and `min_n` in the grid, fit the **whole workflow** on each analysis set and score on each assessment set.


## Step 4 — Run cross-validation and tuning (code)

```{r}
#| label: tm-tune
#| echo: true
folds <- vfold_cv(peng_cls, v = 5, strata = y)
metrics <- metric_set(roc_auc, accuracy, sens, spec)

tune_res <- tune_grid(
  wf,
  resamples = folds,
  grid = grid_regular(
    tree_depth(range = c(2, 6)),
    min_n(range = c(2, 20)),
    levels = 3
  ),
  metrics = metrics
)
```


## Step 5 — Choose the winner and **finalize** (concepts)

- **`select_best(tune_res, metric = "roc_auc")`**: pick the depth / `min_n` pair with best **ROC AUC** on CV (you could choose another metric in real projects).
- **`finalize_workflow(wf, best)`**: plugs those numbers into the unfitted workflow — still no fit on the full dataset until the next line.
- **`fit(final_wf, data = peng_cls)`** below: for **teaching speed** we fit on **all** penguin rows here. In research you would normally **`fit` on training only** after tuning, then **`last_fit`** or a single test assessment.


## Step 5 — Finalize and fit (code)

```{r}
#| label: tm-final
#| echo: true
best <- select_best(tune_res, metric = "roc_auc")
final_wf <- finalize_workflow(wf, best)
final_fit <- fit(final_wf, data = peng_cls)
```

```{r}
#| label: tm-optional-split
#| echo: true
#| eval: false
# --- Optional: last_fit with a held-out split for honest performance ---------
# data_split <- initial_split(peng_cls, prop = 0.8, strata = y)
# train <- training(data_split)
# test  <- testing(data_split)
# last_fit(final_wf, split = data_split)  # after finalize_workflow(...)
```


## Figure 1 — How to read the **tuning** plot (Penguins)

- Each panel shows how **one metric** moves over the **grid** of `tree_depth` and `min_n`.
- Ask students: “Did ROC AUC and accuracy **agree** on the best region, or trade off?” — small *N* penguins often make metrics a bit noisy.


## Figure 1 — Tuning results (code)

```{r}
#| label: tm-fig-autoplot
#| echo: true
#| fig-width: 10
#| fig-height: 5.2
tune_res |>
  autoplot() +
  theme_minimal() +
  labs(
    title = "Hyperparameter tuning",
    subtitle = "ROC AUC, accuracy, sensitivity, specificity across tree depth and min_n"
  )
```

Use **`autoplot()`** on `tune_res` to see how metrics move with the grid; follow up with **`collect_metrics()`** for tables.


## Figure 2 — **Variable importance** on penguins (concepts)

- **`vip`** asks: “Which inputs did **this fitted tree** use most?” for separating Adelie vs Gentoo.
- Remind students: **bill length** might rank high because it **correlates** with mass and species ecology — not the same as a causal gene.


## Figure 2 — Variable importance (code)

```{r}
#| label: tm-fig-vip
#| echo: true
#| fig-width: 9
#| fig-height: 5
final_fit |>
  extract_fit_parsnip() |>
  vip(geom = "point", aesthetics = list(color = "midnightblue", size = 3)) +
  theme_minimal() +
  labs(
    title = "Feature importance",
    subtitle = "Which predictors matter most for Adelie vs Gentoo in this tree?"
  )
```


## Figure 3 — **Tree plot** — connect to real penguins (concepts)

- Each box is a **region** of predictor space after a **split** (e.g. flipper length below / above a threshold).
- Ask: “Where do Adelie vs Gentoo **separate first** in this fitted tree?”


## Figure 3 — Fitted tree structure (code)

```{r}
#| label: tm-fig-rpart
#| echo: true
#| fig-width: 10
#| fig-height: 6
final_fit |>
  extract_fit_engine() |>
  rpart.plot(
    roundint = FALSE,
    type = 4,
    extra = 104,
    box.palette = "RdBu",
    main = "Final decision tree (rpart engine)"
  )
```


## Figure 4 — **ROC curve** on penguins (concepts)

- We plot **Gentoo vs not** using **`.pred_Gentoo`** because **`y`** has levels **Adelie (first)**, **Gentoo (second)** — `event_level = "second"` tells `yardstick` which probability column defines the curve.
- This fit is **in-sample** (same rows as `fit`) — good for **shape** of the curve in class; **not** a substitute for test-set evaluation.


## Figure 4 — ROC curve (code)

```{r}
#| label: tm-fig-roc
#| echo: true
#| fig-width: 8
#| fig-height: 5
augment(final_fit, new_data = peng_cls) |>
  roc_curve(truth = y, .pred_Gentoo, event_level = "second") |>
  autoplot() +
  labs(
    title = "ROC curve",
    subtitle = "Gentoo as event class; in-sample fit — use test data in practice"
  )
```


*Edit only `teaching-hub/_includes/tidymodels-pipeline.qmd`; slide decks and Module 04 pull it in with Quarto `include`.*
