---
title: "Palmer Penguins — multiclass imbalance and upsampling"
author: "ML Summer School"
format:
  html:
    toc: true
    code-tools: true
engine: knitr
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(tidymodels)
  library(themis)
  library(dplyr)
  library(ggplot2)
})
source("../R/slide-viz-helpers.R")
```

# Overview

Minimal walkthrough for [Day 4 Part E](../slides/day-04-thursday.html#/part-e-metrics): three-species classification with a **rare Chinstrap** class, one model, and **`step_upsample()`** in the recipe.

See also [three-species notebook](penguins-species-multiclass.Rmd) and the [Palmer Penguins data card](../data/cards/palmer-penguins.qmd).

## Load imbalanced data

We keep Adelie and Gentoo abundant and retain **15 Chinstrap** rows (deterministic rule: closest to the Adelie morphometric cloud).

```{r}
peng_imb3 <- prep_penguins_multiclass_imbalance(
  minority_class = "Chinstrap",
  hard_to_class = "Adelie",
  n_minority = 15L
)

peng_imb3 |>
  count(y3) |>
  knitr::kable(col.names = c("Species", "n"))
```

```{r fig.width=6, fig.height=3.5}
peng_imb3 |>
  count(y3) |>
  ggplot(aes(y3, n, fill = y3)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n), vjust = -0.25, size = 4) +
  theme_minimal() +
  labs(title = "Class counts (rare Chinstrap)", x = NULL, y = "n")
```

## One model, two recipes

Same **`decision_tree`** spec; only the recipe differs (with vs without upsampling).

```{r}
mc_tree_spec <- decision_tree(tree_depth = 4, min_n = 20) |>
  set_engine("rpart") |>
  set_mode("classification")

rec_no <- recipe(y3 ~ ., data = peng_imb3) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

rec_up <- recipe(y3 ~ ., data = peng_imb3) |>
  step_upsample(y3) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

wf_no <- workflow() |> add_recipe(rec_no) |> add_model(mc_tree_spec)
wf_up <- workflow() |> add_recipe(rec_up) |> add_model(mc_tree_spec)
```

## Train / test split and fit

```{r}
set.seed(13)
split <- initial_split(peng_imb3, prop = 0.8, strata = y3)
train <- training(split)
test <- testing(split)

fit_no <- fit(wf_no, train)
fit_up <- fit(wf_up, train)

pred_no <- augment(fit_no, test)
pred_up <- augment(fit_up, test)
```

## Confusion matrices (holdout)

Rows = true species; columns = predicted species.

**No upsampling:**

```{r}
pred_no |>
  conf_mat(truth = y3, estimate = .pred_class)
```

**With `step_upsample(y3)`:**

```{r}
pred_up |>
  conf_mat(truth = y3, estimate = .pred_class)
```

## Per-class recall (what improves)

```{r}
bind_rows(
  multiclass_recall_table(pred_no, truth_col = "y3") |> mutate(model = "No upsample"),
  multiclass_recall_table(pred_up, truth_col = "y3") |> mutate(model = "Upsample")
) |>
  mutate(recall = round(recall, 3)) |>
  select(model, truth, recall) |>
  tidyr::pivot_wider(names_from = model, values_from = recall) |>
  knitr::kable(caption = "Holdout recall by species")
```

Focus on the **Chinstrap** row: upsampling is useful when it raises minority recall without hiding poor performance on other classes.
