---
title: "Palmer Penguins — canonical pipeline (three models, three metrics)"
author: "ML Summer School"
format:
  html:
    toc: true
    code-tools: true
engine: knitr
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(tidymodels)
  library(palmerpenguins)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(purrr)
})
```

# Overview

Minimal **canonical `tidymodels` pipeline** on Palmer Penguins (**Adelie vs Gentoo**):

1. **Recipe** → preprocess inside resampling  
2. **Three model specs** → same folds, fair comparison  
3. **`fit_resamples()`** → **accuracy**, **kappa**, **ROC AUC**  
4. **One summary figure** (mean ± SE across folds)

Companion: [Module 04 — canonical pipeline](../modules/module-04-pipeline.qmd#canonical-pipeline-tuesday), [Module 07](../modules/module-07-penguins-choose-metrics.qmd), [Module 08](../modules/module-08-penguins-compare-models.qmd).

## Data, recipe, and folds

```{r}
peng <- penguins |>
  filter(species %in% c("Adelie", "Gentoo")) |>
  mutate(
    y = factor(species, levels = c("Adelie", "Gentoo")),
    year = as.numeric(year)
  ) |>
  select(-species, -flipper_length_mm, -body_mass_g) |>
  drop_na()

rec <- recipe(y ~ ., data = peng) |>
  step_zv(all_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())

folds <- vfold_cv(peng, v = 5, strata = y)
metrics <- metric_set(accuracy, kap, roc_auc)
```

## Three models (fixed settings)

```{r}
tree_spec <- decision_tree(tree_depth = 4, min_n = 10) |>
  set_engine("rpart") |>
  set_mode("classification")

glm_spec <- logistic_reg() |>
  set_engine("glm") |>
  set_mode("classification")

rf_spec <- rand_forest(mtry = 4, trees = 300, min_n = 2) |>
  set_engine("ranger") |>
  set_mode("classification")

workflows <- list(
  decision_tree = workflow() |> add_recipe(rec) |> add_model(tree_spec),
  logistic_glm = workflow() |> add_recipe(rec) |> add_model(glm_spec),
  random_forest = workflow() |> add_recipe(rec) |> add_model(rf_spec)
)
```

## Cross-validation (`fit_resamples`)

```{r}
set.seed(7)
cv_results <- workflows |>
  imap(\(wf, name) fit_resamples(wf, folds, metrics = metrics)) |>
  set_names(names(workflows))
```

## Metrics table

```{r}
cmp <- imap_dfr(
  cv_results,
  \(rs, name) collect_metrics(rs) |> mutate(model = name)
)

cmp |>
  select(model, .metric, mean, std_err) |>
  mutate(
    mean = round(mean, 3),
    std_err = round(std_err, 3)
  ) |>
  arrange(.metric, desc(mean)) |>
  knitr::kable(col.names = c("Model", "Metric", "Mean", "Std err"))
```

## Performance summary (final figure)

Same recipe and folds for every model — differences reflect **model family** only.

```{r fig.width=9, fig.height=4.5}
plot_df <- cmp |>
  mutate(
    model = recode(
      model,
      decision_tree = "Decision tree",
      logistic_glm = "Logistic regression",
      random_forest = "Random forest"
    ),
    metric = recode(
      .metric,
      accuracy = "Accuracy",
      kap = "Kappa",
      roc_auc = "ROC AUC"
    )
  )

ggplot(plot_df, aes(model, mean, fill = model)) +
  geom_col(show.legend = FALSE, width = 0.72) +
  geom_errorbar(
    aes(ymin = mean - std_err, ymax = mean + std_err),
    width = 0.18,
    linewidth = 0.5
  ) +
  facet_wrap(~metric, scales = "free_y", ncol = 3) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Model comparison (5-fold CV, same recipe)",
    subtitle = "Bars = mean across folds; error bars = standard error",
    x = NULL,
    y = "Score (higher is better)"
  ) +
  theme_minimal(base_size = 13) +
  theme(strip.text = element_text(face = "bold"))
```

**Takeaway:** pick the **metric** that matches your goal (e.g. ROC AUC for ranking, kappa when classes are imbalanced), then compare models on that score — small gaps within the error bars are often noise on this \(n\).
